### 2023-12-13 EDA.Rmd
Added basic code - EDA, importing data, tree and logistic regression

---
title: "EDA"
author: "Mihir Kulkarni"
date: "2023-12-12"
output: html_document
---

```{r}
#Required libraries

library(survival)
library(ggplot2)
library(survival)
library(survminer)

```

### Survival Analysis Based on Treatment and Stage: How does the survival rate differ among patients at different stages of PBC who received D-penicillamine versus placebo?

This question investigates the effectiveness of the drug D-penicillamine compared to a placebo, considering the stage of PBC. By analyzing survival rates across different disease stages and treatment types, we can assess the drug's effectiveness at various disease stages. This information is vital for clinicians to make informed decisions about treatment plans and for researchers to understand the drug's impact.

```{r}
surv_object <- Surv(cirrhosis$N_Days, cirrhosis$Status == 'D')
surv_fit <- survfit(surv_object ~ cirrhosis$Drug + cirrhosis$Stage, data = cirrhosis)
ggsurvplot(surv_fit, 
           data = cirrhosis,
           conf.int = TRUE, 
           #palette = c("#00AFBB", "#E7B800", "#FC4E07"),
           xlab = "Days", 
           ylab = "Survival Probability", 
           title = "Survival Curves by Treatment and Stage")
```

### Impact of Liver Complications on Survival: How do the presence of ascites, hepatomegaly, and spiders relate to survival time?

Exploring how liver-related symptoms (ascites, hepatomegaly, and spiders) affect patient survival provides insights into the severity of these complications in PBC progression. This can help in identifying high-risk patients and understanding the disease's impact on liver function.

```{r}
surv_fit_complications <- survfit(surv_object ~ cirrhosis$Ascites + cirrhosis$Hepatomegaly + cirrhosis$Spiders, data = cirrhosis)
ggsurvplot(surv_fit_complications, 
           data = cirrhosis,
           conf.int = TRUE, 
           #palette = c("#2E9FDF", "#FC4E07", "#6ACC65"),
           xlab = "Days", 
           ylab = "Survival Probability", 
           title = "Survival Curves by Liver Complications")
```

### Correlation between Biochemical Markers and Disease Stage: What is the correlation between biochemical markers (Bilirubin, Cholesterol, Albumin, Copper, Alk_Phos, SGOT, Triglycerides) and the histologic stage of the disease?

This analysis is crucial to understand how different biochemical markers correlate with the disease's progression. Such correlations can aid in the early detection of disease severity, help in monitoring the disease progression, and potentially guide treatment adjustments.

```{r}
library(reshape2)
library(ggplot2)
cirrhosis_melted <- melt(cirrhosis, id.vars = "Stage", measure.vars = c("Bilirubin", "Cholesterol", "Albumin", "Copper", "Alk_Phos", "SGOT"))
# Worry about , "Triglycerides" later
ggplot(cirrhosis_melted, aes(x = Stage, y = value, color = variable)) +
    geom_point() +
    scale_color_brewer(palette = "Dark2") +
    theme_minimal() +
    labs(title = "Biochemical Markers vs. Disease Stage", x = "Stage", y = "Marker Level")
```

### Gender Differences in Disease Progression: Is there a difference in the progression of PBC between male and female patients?

First, let us plot the sex distribution

```{r}
library(ggplot2)
library(dplyr)

cirrhosis$Sex <- factor(ifelse(cirrhosis$Sex == 0, "Female", "Male"))

# Preparing the data for pie chart
sex_data <- cirrhosis %>% 
            count(Sex) %>% 
            mutate(Percentage = n / sum(n) * 100)

# Pie Chart
ggplot(sex_data, aes(x="", y=Percentage, fill=Sex)) +
    geom_bar(stat="identity", width=1) +
    coord_polar("y", start=0) +
    scale_fill_brewer(palette="Pastel1") +
    theme_minimal() +
    theme(axis.line = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(),
          plot.title = element_text(size=14, face="bold")) +
    labs(fill="Sex", title="Distribution of Sex in PBC Patients")

```

Investigating gender-based differences in disease progression can uncover any gender-specific patterns in PBC. This could lead to gender-tailored treatment approaches and a better understanding of the disease's biology, which might differ between males and females.

```{r}
surv_fit_gender <- survfit(surv_object ~ cirrhosis$Sex, data = cirrhosis)
ggsurvplot(surv_fit_gender, 
           data = cirrhosis,
           conf.int = TRUE, 
           palette = c("#FC4E07", "#2E9FDF"),
           xlab = "Days", 
           ylab = "Survival Probability", 
           title = "Survival Curves by Gender")

```

### Effect of Age at Diagnosis on Survival

```{r}
# Converting age from days to years for better readability
cirrhosis$Age_Years <- cirrhosis$Age / 365.25

# Histogram
ggplot(cirrhosis, aes(x=Age_Years)) +
    geom_histogram(binwidth=5, fill="#69b3a2", color="#e9ecef") +
    theme_minimal() +
    labs(title="Age Distribution of PBC Patients", x="Age (years)", y="Count") +
    theme(plot.title = element_text(size=14, face="bold"))

```

Age at diagnosis is a critical factor in many diseases. This analysis can reveal if younger or older patients have better survival rates, influencing screening practices and early intervention strategies.

```{r}
ggplot(cirrhosis, aes(x = Age, y = N_Days, color = Status)) +
    geom_point() +
    scale_color_brewer(palette = "Set1") +
    theme_minimal() +
    labs(title = "Age at Diagnosis vs. Survival Time", x = "Age (days)", y = "Survival Time (days)")
```
### Comparison of Edema Status on Patient Outcomes

Edema is a common complication in liver diseases and can significantly impact patient quality of life and survival. Understanding how different levels of edema (none, controlled by diuretics, or persistent despite treatment) affect survival can inform patient management strategies and highlight the need for aggressive interventions in certain cases.

```{r}
surv_fit_edema <- survfit(surv_object ~ cirrhosis$Edema, data = cirrhosis)
ggsurvplot(surv_fit_edema, 
           data = cirrhosis,
           conf.int = TRUE, 
           palette = c("#6ACC65", "#E7B800", "#00AFBB"),
           xlab = "Days", 
           ylab = "Survival Probability", 
           title = "Survival Curves by Edema Status")

```

### 2023-12-13 Final Project.Rmd
Added basic code - EDA, importing data, tree and logistic regression

---
title: "Project #4: Final Project Report: An exploration of Cirrhosis"
output: html_notebook
---

- Set the stage–explain the context you’re trying to learn something about. MUST ANSWER: Why is this important to you?

we're both interested in the health applications of AI and found an interesting article about cirrhosis and machine learning. We wanted to see if we could replicate the results and learn more about the disease.

- Explain the data provenance–Answer the Who, What, When, Where, Why, and How questions for the origin of the data. Include any additional information that would be useful in understanding the context in which the data are embedded.

- Data Preprocessing

```{r}
#load in the data
library(tidyverse)
#load data
cirrhosis <- read_csv("cirrhosis.csv")
head(cirrhosis)
```

# Pre-process the data
```{r}
#encode the sex, ascites, hepatomegaly, spiders, edema into binary variables
cirrhosis$Sex <- ifelse(cirrhosis$Sex == "F", 0, 1)
cirrhosis$Ascites <- ifelse(cirrhosis$Ascites == "Y", 1, 0)
cirrhosis$Hepatomegaly <- ifelse(cirrhosis$Hepatomegaly == "Y", 1, 0)
cirrhosis$Spiders <- ifelse(cirrhosis$Spiders == "Y", 1, 0)
#factorize the stage
cirrhosis$Stage <- factor(cirrhosis$Stage)
cirrhosis$Status <- factor(cirrhosis$Status)
#need to do more
#Drop the first column
cirrhosis <- cirrhosis[,-1]
```

- Engage in Exploratory Data Analysis

- Use Machine Learning Algorithms:
  - You must use at least one of each: a Regression-based algorithm, a Tree-based algorithm, and an Unsupervised Learning algorithm.
  - Each algorithm must be explained for a general audience.
  - All results should be explained.
  - Where appropriate, comparisons between methods should be discussed. This includes highlighting the strengths and weaknesses of the various approaches as well as when each is appropriate.

- A discussion of what you’ve learned from your work about the context and recommendations for future work.

- Author Contributions–list out who did what in the project.

- References–Citation style is your choice, but all sources should be documented (both in text and in the References section). This includes where you got your data.

- A commented Code Appendix

### 2023-12-13 Regression.Rmd
Added basic code - EDA, importing data, tree and logistic regression

---
title: "Regression-based"
author: "Mihir Kulkarni"
date: "2023-12-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2023-12-13 Tree.Rmd
Added basic code - EDA, importing data, tree and logistic regression

---
title: "Tree-based"
author: "Mihir Kulkarni"
date: "2023-12-12"
output: html_document
---

### Explain tree model

### Pre-processing the data

```{r}
#Combine C and CL status into one variable and binarize
cirrhosis$Status <- ifelse(cirrhosis$Status == "C" | cirrhosis$Status == "CL", 1, 0)
```

# Split the data sets into training and testing

```{r}
#Set up the data

# Wrangle the Graduate data to set up training and testing datasets
modelCirrhosis <- cirrhosis %>%
  #drop_na() %>%
  mutate(
    tempID = row_number(),
    .before = Status
  )

## Set seed for reproducibility and slice ----
set.seed(380)
trainingData <- modelCirrhosis %>%
  group_by(Status) %>%
  slice_sample(prop = 0.8)

testingData <- modelCirrhosis %>%
  filter(!(tempID %in% trainingData$tempID))
```

### Grow tree via rpart

```{r}
# Grow Graduate tree via rpart package
rPartStatus <- rpart(
  formula = Status ~ N_Days + Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage,
  #Lost Triglycerides + 
  data = trainingData,
  method = "class",
  parms = list(split = "information")
  # We did not need to use the control parameters
)

```

### Plot tree

```{r}
# Display rpart.plot ----
# library(rpart.plot)
rpart.plot(
  x = rPartStatus,
  type = 2,
  extra = 101
)
```

### Plot tree

```{r}
# Using the rattle package to visualize the tree ----
fancyRpartPlot(
  model = rPartStatus,
  main = NULL,
  sub = NULL
)
```

### Plot cpTable

```{r}
invisible(capture.output({cpTable <- printcp(rPartStatus)}))

kable(
  x = cpTable,
  col.names = c("CP", "Num. of splits", "Rel. Error",
                "Mean Error", "Std. Deviation of Error"),
  digits = 3,
  booktabs = TRUE,
  align = "c",
  table.attr = 'data-quarto-disable-processing="true"'
)
```

### Plot cp visualization

```{r}
plotcp(
  x = rPartStatus,
  minline = TRUE,
  upper = "size"
)
```

### Prune tree

```{r}
# Prune the rpart Graduate Tree ----
rPartStatus2 <- prune(
  tree = rPartStatus,
  cp = 0.016
)
```

### Plot pruned tree

```{r}
fancyRpartPlot(
  model = rPartStatus2,
  main = NULL,
  sub = NULL
)
```

### Predictions

```{r}
pred_StatusRpart2 <- predict(
  object = rPartStatus2,
  newdata = testingData,
  type = "prob"
)

# Data Wrangling the predictions ----
StatusPrediction <- data.frame(
  rpart2_non_death = pred_StatusRpart2[, 1],
  rpart2_death = pred_StatusRpart2[, 2]
) %>%
  mutate(
    rpart2_pred = ifelse(
      test = rpart2_death > rpart2_non_death,
      yes = 1,
      no = 0
    )
  )

## Set predictions as factors
StatusPrediction$rpart2_pred <- as.factor(StatusPrediction$rpart2_pred)

# Merge supervision column into predictions data frame ----
StatusPrediction <- cbind(
  tempID = testingData$tempID,
  Status = testingData$Status,
  StatusPrediction
)
```

```{r}
StatusPrediction$Status <- factor(StatusPrediction$Status)

# Build confusion matrix for second tree model
conf_matrix <- conf_mat(
  data = StatusPrediction,
  truth = Status,
  estimate = rpart2_pred
)$table

kable(
  conf_matrix,
  col.names = c("Prediction/Supervision", "0", "1"),
  digits = 3,
  booktabs = TRUE,
  caption = "Model 1: Confusion Matrix",
  align = "c"
) %>%
kable_styling(latex_options = "HOLD_position")

accuracy <- accuracy(StatusPrediction, Status, rpart2_pred)
specificity <- specificity(StatusPrediction, Status, rpart2_pred)
sensitivity <- sensitivity(StatusPrediction, Status, rpart2_pred)
```
```{r}
# Build a data frame with model metrics ----
StatusPreds <- StatusPrediction %>%
  dplyr::select(tempID, Status, contains("_pred")) %>%
  pivot_longer(
    cols = !c(tempID, Status),
    names_to = "model",
    values_to = "prediction"
  )

accuracy <- StatusPreds %>%
  group_by(model) %>%
  accuracy(
    truth = Status,
    estimate = prediction
  )

sensitivity <- StatusPreds %>%
  group_by(model) %>%
  sensitivity(
    truth = Status,
    estimate = prediction,
    event_level = "second"
  )

specificity <- StatusPreds %>%
  group_by(model) %>%
  specificity(
    truth = Status,
    estimate = prediction,
    event_level = "second"
  )

modelMetrics <- bind_rows(
  accuracy,
  sensitivity,
  specificity
)
```
```{r}
# Make a nice looking table of model metrics ----
modelMetrics %>%
  dplyr::select(model, .metric, .estimate) %>%
  pivot_wider(
    id_cols = model,
    names_from = .metric,
    values_from = .estimate
  ) %>%
  kable(
    digits = 3,
    booktabs = TRUE,
    align = "c",
    table.attr = 'data-quarto-disable-processing="true"'
  )
```

```{r}
#Set up the data

# Wrangle the Graduate data to set up training and testing datasets
modelCirrhosis2 <- cirrhosis %>%
  drop_na() %>%
  mutate(
    tempID = row_number(),
    .before = Status
  )

## Set seed for reproducibility and slice ----
set.seed(380)
trainingData <- modelCirrhosis2 %>%
  group_by(Status) %>%
  slice_sample(prop = 0.8)

testingData <- modelCirrhosis2 %>%
  filter(!(tempID %in% trainingData$tempID))
```

# Comparison with logistic regression

```{r}
# Lower bound (Intercept only)
lower <- glm(
  formula = Status ~ 1,
  data = trainingData,
  family = binomial
)

# Upper bound 
upper <- glm(
  formula = Status ~ N_Days + Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage,
  data = trainingData,
  family = binomial
)

# Stepwise search
model2 <- step(
  object = lower,
  scope = list(
    lower = lower,
    upper = upper
  ),
  data = trainingData,
  direction = "both",
  k = 2
)
trainingResults <- trainingData
```

```{r}
# Store the predicted and actual values for Model 3:
trainingResults$model2Pred <- predict(model2, type = "response")
trainingResults$model2Class <- ifelse(trainingResults$model2Pred > 0.5, 1, 0)
trainingResults$Actual <- ifelse(trainingData$Status == 1, 1, 0)

# Create confusion matrix using table()
confusionMatrix <- table(Predicted = trainingResults$model2Class, Actual = trainingResults$Actual)

# Print and format confusion matrix
kable(confusionMatrix, booktabs = TRUE, align = "c")%>%kable_styling(latex_options = "HOLD_position")
```

```{r}
accuracy <- (confusionMatrix[1, 1] + confusionMatrix[0, 0]) / sum(confusionMatrix)
sensitivity <- confusionMatrix[1, 1] / (confusionMatrix[1, 1] + confusionMatrix[0, 1])
specificity <- confusionMatrix[0, 0] / (confusionMatrix[0, 0] + confusionMatrix[1, 0])

metrics_df <- data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity"),
  Value = c(accuracy, sensitivity, specificity)
)

# Use kable to create the table
kable(metrics_df, format = "html", caption = "Model Performance Metrics")
```

### 2023-12-13 Final Project.Rmd
said hi
changed to hello
heyyy

@@
 ---
+hello
 
 - Set the stage–explain the context you’re trying to learn something about. MUST ANSWER: Why is this important to you?
+
+Cirrohosis is a condition where the liver is scarred and permanently damaged.
+As this condition, chances of liver failure and liver cancer increase.
+We are intrested in creating model that can predict stages of Cirrohosis and also death based on certain conditions.
+We hope our work can be useful in catching Cirrosis earlier on.
 

### 2023-12-13 Tree.Rmd
heyyy

@@
 title: "Tree-based"
-author: "Mihir Kulkarni"
+author: "Mihir Kulkarni, Nithika Menon"
 date: "2023-12-12"
@@
 # Grow Graduate tree via rpart package
+library(rpart)
 rPartStatus <- rpart(
-  formula = Status ~ N_Days + Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage,
-  #Lost Triglycerides + 
+  formula = Status ~ N_Days + Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage + Tryglicerides,
   data = trainingData,
@@
 # Display rpart.plot ----
-# library(rpart.plot)
+ library(rpart.plot)
 rpart.plot(
@@
 # Using the rattle package to visualize the tree ----
+library(rattle)
+
 fancyRpartPlot(
@@
 invisible(capture.output({cpTable <- printcp(rPartStatus)}))
+
+library(kableExtra)
 
@@
 StatusPrediction$Status <- factor(StatusPrediction$Status)
+
+library(yardstick)
 

### 2023-12-13 Tree.Rmd
Removed N_days from tree formula

@@
 rPartStatus <- rpart(
-  formula = Status ~ N_Days + Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage + Tryglicerides,
+  formula = Status ~ Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage + Tryglicerides, 
   data = trainingData,

### 2023-12-13 Final Project.Rmd
Added Linear Regression

@@
 cirrhosis$Status <- factor(cirrhosis$Status)
+cirrhosis$Drug <- factor(cirrhosis$Drug)
+cirrhosis$Sex <- factor(cirrhosis$Sex)
+cirrhosis$Ascites <- factor(cirrhosis$Ascites)
+cirrhosis$Hepatomegaly <- factor(cirrhosis$Hepatomegaly)
+cirrhosis$Spiders <- factor(cirrhosis$Spiders)
+cirrhosis$Edema <- factor(cirrhosis$Edema)
 #need to do more

### 2023-12-13 Regression.Rmd
heyyy
Added Linear Regression

@@
 title: "Regression-based"
-author: "Mihir Kulkarni"
+author: "Mihir Kulkarni, Nithika Menon"
 date: "2023-12-12"
@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+```{r}
+count <- colSums(is.na(cirrhosis))
+
+kable(count)
+```
+##### Splitting:
+
+we'll split the dataset into training and testing sets.
+The training set will be used to train our model, while the testing set will help evaluate its performance.
+We'll use 80% of the data for training and the remaining 20% for testing.
+
+```{r}
+cirrhosisLRData <- cirrhosis %>%
+  filter(Status == "D") %>%
+  drop_na()
+
+cirrhosisLRData <- cirrhosisLRData[, -2] #Remove status column
+
+#cirrhosisLRData <- na.omit(cirrhosisLRData)
+
+set.seed(123)  #reproducibility
+
+# Sampling 80% 
+train_data <- cirrhosisLRData %>% sample_frac(0.8)
+
+# Getting the remaining 20%
+test_data <- setdiff(cirrhosisLRData, train_data)
+
 ```
 
-## R Markdown
+##### Test out basic model
+```{r}
+LRModel1 <- lm( #train the insurance model
+  formula = N_Days ~ Platelets,
+  data = test_data,
+  na.action = "na.omit"
+)
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
-
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
-
-```{r cars}
-summary(cars)
+summary(LRModel1)
 ```
 
-## Including Plots
+##### Define the models
 
-You can also embed plots, for example:
+- fullModel predicts N_Days using all variables from the train_data 
+- interceptModel uses only the intercept from the non-missing data
+- initialModel uses only the intercept from the entire dataset.
 
-```{r pressure, echo=FALSE}
-plot(pressure)
+```{r}
+fullModel <- lm(
+  formula = N_Days ~ .,
+  data = na.omit(train_data)
+)
+
+interceptModel <- lm(
+  formula = N_Days ~ 1,
+  data = na.omit(train_data)
+)
+
+initialModel <- lm(
+  formula = N_Days ~ 1,
+  data = train_data
+)
 ```
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+##### Calculating RMSE for first model
+```{r}
+sqrt(mean(residuals(initialModel)^2, na.rm = TRUE))
+```
+
+#####Using stepwise regression
+
+```{r}
+stepwiseTrainModel <- step(
+  object = initialModel, 
+  scope = list(
+    lower = initialModel,
+    upper = fullModel
+  ),
+  data = train_data, 
+  direction = "both",
+  k = log(nrow(na.omit(train_data))),
+  trace = 0
+)
+```
+
+#### Trying various algorithms to choose the best features/attributes
+
+##### Backward Selection
+- Backward selection is a method of fitting regression models in which the choice of predictive variables is carried out by an automatic procedure.
+- step() function is employed to perform backward selection.
+The fullModel is used as the starting model, and variables are iteratively removed until the model reaches the specification of interceptModel, which only contains the intercept.
+The data used for this selection process omits any rows with missing values.
+The direction = backward argument specifies the backward elimination method, and k = 2 determines the penalty for adding a variable.
+Finally, trace = 0 indicates that the intermediate steps of the selection process won't be printed.
+The final model after backward selection is stored in ModelBS
+
+summary of backward:
+```{r}
+ModelBS <- step(
+  object = fullModel,
+  scope = list(
+    lower = interceptModel,
+    upper = fullModel
+  ),
+  data = na.omit(test_data), 
+  direction = "backward",
+  k = 2,
+  trace = 0
+)
+
+# Summary of the final backward selection model ----
+summary(ModelBS)
+```
+
+##### Forward Selection
+
+- Forward selection is a type of stepwise regression which begins with an empty model and adds in variables one by one.
+In each forward step, you add the one variable that gives the single best improvement to your model.
+- The step() function to perform forward model selection.
+Starting from a basic model (interceptModel), the function iteratively adds predictors from the fullModel if they improve the model fit based on some criterion (in this case, likely AIC with a penalty factor of 2, given by k = 2).
+The trace = 0 argument suppresses intermediate output, providing a cleaner console output.
+This technique helps in identifying the most relevant predictors for a regression model in a data-driven manner.
+
+Summary of foward:
+
+```{r}
+ModelFS <- step(
+  object = interceptModel,
+  scope = list(
+    lower = interceptModel,
+    upper = fullModel
+  ),
+  data = na.omit(test_data), 
+  direction = "forward",
+  k = 2,
+  trace = 0
+)
+
+# Summary of the final forward selection model ----
+summary(ModelFS)
+```
+
+##### Stepwise
+
+- Stepwise regression is the step-by-step iterative construction of a regression model that involves the selection of independent variables to be used in a final model.
+It involves adding or removing potential explanatory variables in succession and testing for statistical significance after each iteration.
+
+- the step function conducts stepwise selection for a regression model.
+Initially, it starts with an interceptModel and evaluates the inclusion or exclusion of predictor variables based on their significance and contribution to the model fit.
+The scope argument specifies the simplest and most complex models.
+In this case, the simplest is the interceptModel and the most complex is the fullModel.
+The selection process can both add and remove predictors based on their importance.
+The penalty for model complexity is determined by the logarithm of the number of observations.
+
+Summary of stepwise:
+
+```{r}
+# Use Stepwise Selection ---
+ModelSS <- step(
+  object = interceptModel, 
+  scope = list(
+    lower = interceptModel,
+    upper = fullModel
+  ),
+  data = na.omit(test_data), 
+  direction = "both",
+  k = log(nrow(na.omit(test_data))),
+  trace = 0
+)
+
+# Summary of the final stepwise selection model ----
+summary(ModelSS)
+```
+
+```{r}
+## Backward Selection testing
+sqrt(mean(residuals(stepwiseTrainModel)^2, na.rm = TRUE))
+```
+
+##### Build QQ Plot for the stepwise Train Model
+
+```{r}
+
+library(car)
+car::qqPlot(
+  x = residuals(stepwiseTrainModel),
+  distribution = "norm",
+  id = FALSE,
+  envelope = 0.9,
+  pch = 20,
+  xlab = "Theoretical quantiles",
+  ylab = "Residuals (grams)",
+  main = "Assess Gaussian Errors"
+)
+```
+
+##### Calculate Skewness of Residuals from the Stepwise Training Model
+
+```{r}
+library(psych)
+# Sample Skewness ----
+skew(residuals(stepwiseTrainModel))
+```
+
+##### Sample Excess Kurtosis
+```{r}
+kurtosi(residuals(stepwiseTrainModel))
+```
+
+##### Built a plot for the model
+
+```{r}
+ggplot(
+  data = data.frame(
+    residuals = residuals(stepwiseTrainModel),
+    fitted = fitted(stepwiseTrainModel)
+  ),
+  mapping = aes(x = fitted, y = residuals)
+) +
+  geom_point() +
+  geom_smooth(
+    formula = y ~ x,
+    method = stats::loess,
+    method.args = list(degree = 1),
+    se = FALSE,
+    linewidth = 0.5
+  ) +
+  geom_hline(
+    yintercept = mean(residuals(stepwiseTrainModel)),
+    color = "darkgrey",
+    linetype = "solid",
+    linewidth = 1
+  ) +
+  geom_hline(
+    yintercept = 0,
+    color = "red",
+    linetype = "dashed"
+  ) +
+  theme_bw() +
+  labs(
+    x = "Fitted values (grams)",
+    y = "Residuals (grams)"
+  )
+```
+
+```{r}
+library(car)
+library(dplyr)
+library(knitr)
+library(kableExtra)
+
+# Calculate VIFs
+vif_values <- vif(ModelBS)
+
+# Convert to data frame
+vif_df <- data.frame(
+  Variable = names(vif_values),
+  VIF = as.numeric(vif_values)
+)
+
+# Add squared VIF column
+vif_df <- vif_df %>%
+  mutate(
+    VIF_Squared = VIF^2
+  )
+
+# Display the results
+kable(
+  x = vif_df,
+  digits = 3,
+  align = "cccc",
+  booktab = TRUE,
+  table.attr = 'data-quarto-disable-processing="true"'
+) %>%
+  kableExtra::kable_classic(
+    latex_options = c("HOLD_position"),
+    full_width = FALSE
+  )
+
+```
+
+### PREDICT
+```{r}
+predictVals <- predict(
+  object = stepwiseTrainModel,
+  newdata = test_data
+)
+
+# Add predictions to data frame and find residuals ----
+test_data$prediction <- predictVals
+test_data <- test_data %>%
+  mutate(
+    residuals = N_Days - prediction
+  )
+
+# Calculate RMSE for Testing Data ----
+rmse_testing <- sqrt(mean(test_data$residuals^2, na.rm = TRUE))
+rmse_testing
+```
+
+##### Plot comparing actual charges to predicted charges from the test_data
+
+the blue reference line indicates wher predictions match actual values perfectly.
+```{r}
+ggplot(
+  data = test_data,
+  mapping = aes(x = N_Days, y = prediction)
+) +
+  geom_point() +
+  geom_abline( 
+    slope = 1,
+    intercept = 0,
+    color = "blue",
+    linetype = "solid",
+    linewidth = 1
+  ) +
+  theme_bw() +
+  labs(
+    x = "Actual",
+    y = "Predicted"
+  )
+```

### 2023-12-13 Tree.Rmd
Wrote the descriptions for CART

@@
 
-### Explain tree model
+# Part I: Classification Algorithm for the Status of the Patient
 
-### Pre-processing the data
+To address the first sub-goal of this project, we will be exploring the prediction of the status of a patient at the end of the study.
+Our main objective through this is to gain a stronger understanding of what factors played in the role of the death of the patient.
+To do this, we will be using Decision Trees.
+Let us start by exploring what these are and why we chose to use them.
+
+## Introduction
+
+Decision trees are a type of model used in statistics for making predictions based on data.
+They work by breaking down a dataset into smaller subsets through "splits," resembling a tree with branches.
+Each branch represents a possible decision or outcome, leading to a final prediction or classification.
+
+The main advantages of decision trees include their simplicity and interpretability, as they are easy to visualize and understand.
+Another important advantage is that they are able to learn with data with N/A values, something alternatives such as Logistic Regression.
+Considering the limitation we have with our data size and number of N/A values, this an important advantage.
+However, there are some downsides such as a tendency to overfit the data, which is something we need to be careful about.
+
+![[Decision Tree Example (]{.underline}[Source](https://data-flair.training/blogs/r-decision-trees/))](DecTreeEg.png)
+
+## Pre-processing the data
+
+For this data, we will be dealing with the Status variable. The Status variable was divided into 3 classes:
+
+-   Class 0 (D): The patient didn't survive.
+
+```{=html}
+<!-- -->
+```
+-   Class 1 (C): The patient is censored, meaning that the observation period ended without the death being recorded.
+
+-   Class 2 (CL): Similar to Class 1, the patient is censored due to liver transplantation.
+
+Thus, we can group Class 1 and 2.
 
@@
 #Combine C and CL status into one variable and binarize
-cirrhosis$Status <- ifelse(cirrhosis$Status == "C" | cirrhosis$Status == "CL", 1, 0)
+cirrhosisTreeData <- cirrhosis
+cirrhosisTreeData$Status <- ifelse(cirrhosisTreeData$Status == "C" | cirrhosisTreeData$Status == "CL", 1, 0)
 ```
 
-# Split the data sets into training and testing
+## Methodology
+
+Our objective is to create a classifier capable of predicting a patient's outcome.
+To achieve this, we will be testing our data with the cart algorithm.
+To ensure model validation, we'll be using a 80% training and 20% testing data division.
+I will also be stratifying the data based on the status.
+
+To guarantee consistency and reproducibility in our results, we have fixed the seed for our 80/20 data split at 380.
+With these steps, we are now well-positioned to finalize our training and testing data sets.
+
+The predictors in these models will be guided by the results from the EDA. We will also use a variety of tools to understand the model's performance.
 
 ```{r}
-#Set up the data
-
 # Wrangle the Graduate data to set up training and testing datasets
-modelCirrhosis <- cirrhosis %>%
+modelCirrhosis <- cirrhosisTreeData %>%
   #drop_na() %>%
@@
 
-### Grow tree via rpart
+### Step 1: Growing the tree
+
+There are useful packages to build decision trees in R: the {tree} and the {rpart} (recursive partitioning) packages.
+
+In this report, I have decided to use {rpart} because it provides more flexibility for surrogate splits and the trees are a bit easier to make attractive looking.
 
@@
 
-### Plot tree
+### Part 2: Visualizing the tree
+
+With the tree grown, we can now visualize it for an easy understanding of its functioning. This is an important advantage for CART over logistic regression.
+
+The following is a basic diagram for the tree that was just grown.
 
@@
 
-### Plot tree
+(Explanation of above) To gain a further understanding of the data, we can plot a tree yielding Collection Node style trees. This can help us understand how the data is split.
 
@@
 
-### Plot cpTable
+### Part 3: Pruning the tree
+
+Pruning reduces the size of decision trees by removing parts of the tree that do not provide power to classify instances.
+The first step of pruning a tree is understanding the complexity parameter used.
+The complexity parameter (cp) in rpart is the minimum improvement in the model needed at each node.
+This is used when building the tree.
+We can see the results based on the cross validation from the table below.
 
@@
 
-### Plot cp visualization
+This can also be visualized in a graph to gain a better understanding of the data. The graph below shows the connection between the cp, size of tree and the x-val relative error.
 
@@
 
-### Prune tree
+From the graph we can see that a cp of 0.011 is ideal as it is under the horizontal (dotted) reference line. We can prune the tree with this CP value.
 
@@
 
-### Plot pruned tree
+We can plot the pruned tree
 
@@
 
-### Predictions
+### Part 4: Results
+
+Now, we can evaluate the results of the tree on the testing data from the initial 80-20 split.
+As is true whenever we use validation approaches, we need to test out our model on the testing data set.
+This will give us a more accurate understanding of how well the model fits the context we're seeking to build our understanding of.
+
+An important part of our results is understanding the role of prediction and inference.
+In a broad sense, prediction refers to the process of making forecasts about future events or unknown values based on a model while inference generally refers to the process of drawing conclusions from data.
+For the basic tree, I will be mainly focusing on prediction aspects of the results.
+However, later in the report, we will also be exploring inference findings.
 
@@
 
+We can evaluate the results of this through a confusion matrix.
+
 ```{r}
@@
 ```
+
 ```{r}
@@
 ```
+
+From this, we can get the model's
+
+-   True Positive (TP): Predicted Graduated and actually Graduated. In this case, 2217
+
+-   False Positive (FP): Predicted Graduated but actually Not Graduated. In this case, 1000
+
+-   False Negative (FN): Predicted Not Graduated but actually Graduated. In this case, 244
+
+-   True Negative (TN): Predicted Not Graduated and actually Not Graduated. In this case, 659
+
+With this, we can also calculate the model's metrics.
+
 ```{r}

### 2023-12-13 Regression.Rmd
Did some descriptiosn for the linear regression model

@@
 
-```{r}
-count <- colSums(is.na(cirrhosis))
+# Part II: Prediction for the Number of Days until an event (death)
 
-kable(count)
-```
-##### Splitting:
+Next, let us address our next sub-goal: predicting the number of days a patient survives.
+The objective is to use the medical test data help doctors make better decisions.
+To do this, we will be using linear regression.
 
-we'll split the dataset into training and testing sets.
+## Introduction
+
+Linear regression is a statistical technique to model the relationship between a dependent variable and one or more independent variables.
+The model tries to draw a straight line that best fits the data points on a graph, representing how changes in the independent variables are associated with changes in the dependent variable.
+The primary advantage of linear regression is its simplicity and ease of interpretation.
+However, its main drawbacks are that it assumes a linear relationship between variables, which may not always be true.
+This type of model can be very sensitive to outliers, which affects the model's accuracy, especially dealing with medical data.
+Furthermore, unlike the decision tree, linear regression models cannot ignore N/A values.
+
+![[Linear Regression Example (]{.underline}[Source](https://www.spiceworks.com/tech/artificial-intelligence/articles/what-is-linear-regression/))](LinRegEg.png)
+
+## Methodology
+
+Similar to the tree, we will start by splitting the data set into training and testing sets.
+Note that the data now onluy contains instances where the patient was deceased.
 The training set will be used to train our model, while the testing set will help evaluate its performance.
 We'll use 80% of the data for training and the remaining 20% for testing.
+To allow reproducible code, we have fixed the seed at 380.
 
@@
 
-#cirrhosisLRData <- na.omit(cirrhosisLRData)
-
-set.seed(123)  #reproducibility
+set.seed(380)  #reproducibility
 
@@
 
-##### Test out basic model
+### Step 1: Train basic tree
+
 ```{r}
-LRModel1 <- lm( #train the insurance model
+LRModel1 <- lm( #train the basic tree
   formula = N_Days ~ Platelets,
@@
 
-##### Define the models
+### Step 2: Train basic models
 
@@
 ##### Calculating RMSE for first model
+
 ```{r}
@@
 
-#####Using stepwise regression
+### Step 3: Testing selection models
 
@@
 
-#### Trying various algorithms to choose the best features/attributes
+Here, we will be trying various algorithms to choose the best features/attributes
 
-##### Backward Selection
+#### (a) Backward Selection
+
 -   Backward selection is a method of fitting regression models in which the choice of predictive variables is carried out by an automatic procedure.
+
 -   step() function is employed to perform backward selection.
@@
 summary of backward:
+
 ```{r}
@@
 
-##### Forward Selection
+#### (b) Forward Selection
 
@@
 
-##### Stepwise
+#### (c) Stepwise
 
@@
 
-##### Build QQ Plot for the stepwise Train Model
+## Step 4: Results
+
+### Build QQ Plot for the stepwise Train Model
 
@@
 
-##### Calculate Skewness of Residuals from the Stepwise Training Model
+### Calculate Skewness of Residuals from the Stepwise Training Model
 
@@
 
-##### Sample Excess Kurtosis
+#### Sample Excess Kurtosis
+
 ```{r}
@@
 
-##### Built a plot for the model
+#### Built a plot for the model
 
@@
 
-### PREDICT
+### Predictions
+
 ```{r}
@@
 
-##### Plot comparing actual charges to predicted charges from the test_data
+### Plot predictions
 
-the blue reference line indicates wher predictions match actual values perfectly.
+the blue reference line indicates where predictions match actual values perfectly.
+
 ```{r}

### 2023-12-13 Regression.Rmd
yuh

--- 2023-12-13 Regression.Rmd
@@
 ```
-
-## R Markdown
-
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
-
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
-
-```{r cars}
-summary(cars)
-```
-
-## Including Plots
-
-You can also embed plots, for example:
-
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
-
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2023-12-14 CART.Rmd
b

--- 2023-12-13 Tree.Rmd
@@
 
-```{=html}
-<!-- -->
-```
 -   Class 1 (C): The patient is censored, meaning that the observation period ended without the death being recorded.
@@
 
+### Dataset for Tree
 ```{r}
-#Combine C and CL status into one variable and binarize
-cirrhosisTreeData <- cirrhosis
-cirrhosisTreeData$Status <- ifelse(cirrhosisTreeData$Status == "C" | cirrhosisTreeData$Status == "CL", 1, 0)
+
+cirrhosisTree <- cirrhosis
+
+```
+
+### 
+```{r}
+
+cirrhosisTree <- cirrhosisTree %>%
+                                  filter(Status != "D") %>%
+                                                      mutate(Transplant = ifelse(Status == 'C', 0, 1 ))
+
 ```
@@
 ```
+
+## Discussion

### 2023-12-14 EDA.Rmd
heyyy
b

@@
 title: "EDA"
-author: "Mihir Kulkarni"
+author: "Mihir Kulkarni, Nithika Menon"
 date: "2023-12-12"
@@
 
-```{r}
-library(reshape2)
-library(ggplot2)
-cirrhosis_melted <- melt(cirrhosis, id.vars = "Stage", measure.vars = c("Bilirubin", "Cholesterol", "Albumin", "Copper", "Alk_Phos", "SGOT"))
-# Worry about , "Triglycerides" later
-ggplot(cirrhosis_melted, aes(x = Stage, y = value, color = variable)) +
-    geom_point() +
-    scale_color_brewer(palette = "Dark2") +
-    theme_minimal() +
-    labs(title = "Biochemical Markers vs. Disease Stage", x = "Stage", y = "Marker Level")
-```
-
-### Gender Differences in Disease Progression: Is there a difference in the progression of PBC between male and female patients?
-
-First, let us plot the sex distribution
-
+### Sex Distribution of Patients
 ```{r}
@@
 ```
-
+Primary Biliary Cholangitis affects women much more than men at a 10-1 ratio.
+The data of this clinical study mimics the greater population.
 Investigating gender-based differences in disease progression can uncover any gender-specific patterns in PBC.
@@
 
-```{r}
-surv_fit_gender <- survfit(surv_object ~ cirrhosis$Sex, data = cirrhosis)
-ggsurvplot(surv_fit_gender, 
-           data = cirrhosis,
-           conf.int = TRUE, 
-           palette = c("#FC4E07", "#2E9FDF"),
-           xlab = "Days", 
-           ylab = "Survival Probability", 
-           title = "Survival Curves by Gender")
-
-```
-
-### Effect of Age at Diagnosis on Survival
+### Age Distribution of Patients
 
@@
 ```
+The distribution of ages is quite normal.
 
-Age at diagnosis is a critical factor in many diseases.
-This analysis can reveal if younger or older patients have better survival rates, influencing screening practices and early intervention strategies.
+### might have to mutate data below to filter out alive ppl but get the gist
+```{r}
+
+ggplot(data = cirrhosis, aes (Drug, Stage, fill = N_Days))  + geom_tile() + scale_fill_distiller(palette = "RdPu")
+
+```
+
+### Individuals with Edema and if they were on the drug
+```{r}
+
+library(ggmosaic)
+cirrhosis %>%
+  ggplot() +
+  geom_mosaic(aes( x = product(Edema), fill = Drug))
+#ggplot(data = cirrhosis) + geom_mosaic(aes(x = Drug , fill = Status))
+
+```
+Looking at the dist. of stages
+```{r}
+
+cirrhosis %>%
+      ggplot(aes(x = Stage)) + geom_bar() # need color
+
+```
 
 ```{r}
-ggplot(cirrhosis, aes(x = Age, y = N_Days, color = Status)) +
-    geom_point() +
-    scale_color_brewer(palette = "Set1") +
-    theme_minimal() +
-    labs(title = "Age at Diagnosis vs. Survival Time", x = "Age (days)", y = "Survival Time (days)")
+cirrhosis %>%
+      ggplot(aes(x = Cholesterol, y = Bilirubin)) + geom_line()
+
 ```
+
 ### Comparison of Edema Status on Patient Outcomes
@@
 Understanding how different levels of edema (none, controlled by diuretics, or persistent despite treatment) affect survival can inform patient management strategies and highlight the need for aggressive interventions in certain cases.
-
-```{r}
-surv_fit_edema <- survfit(surv_object ~ cirrhosis$Edema, data = cirrhosis)
-ggsurvplot(surv_fit_edema, 
-           data = cirrhosis,
-           conf.int = TRUE, 
-           palette = c("#6ACC65", "#E7B800", "#00AFBB"),
-           xlab = "Days", 
-           ylab = "Survival Probability", 
-           title = "Survival Curves by Edema Status")
-
-```

### 2023-12-14 Final Project.Rmd
Did some descriptiosn for the linear regression model
yuh
b

@@
 ---
-title: "Project #4: Final Project Report: An exploration of Cirrhosis"
+title: "Final Project Report: An exploration of Biliary Cirrhosis and Treatments"
 output: html_notebook
+authors: "Mihir Kulkarni, Nithika Menon"
 ---
-hello
 
-- Set the stage–explain the context you’re trying to learn something about. MUST ANSWER: Why is this important to you?
+# Background
 
-Cirrohosis is a condition where the liver is scarred and permanently damaged.
-As this condition, chances of liver failure and liver cancer increase.
-We are intrested in creating model that can predict stages of Cirrohosis and also death based on certain conditions.
-We hope our work can be useful in catching Cirrosis earlier on.
+## What is Biliary Cholangitis?
 
-we're both interested in the health applications of AI and found an interesting article about cirrhosis and machine learning.
-We wanted to see if we could replicate the results and learn more about the disease.
+Primary Biliary Cholangitis (previously called Primary Biliary Cirrhosis) is an autoimmune disease where bile ducts become swollen and inflamed and block the flow of bile.
+Bile is a substance that aids with digestion.
+The bile ducts carry bile from the liver to the small intestine.
+The swelling and inflammation can lead to scarring of the liver which is cirrhosis.
+Advanced cirrhosis can lead to liver failure or liver cancer.
+Medication can slow progression.
+There is no definite cure at this time.
 
-- Explain the data provenance–Answer the Who, What, When, Where, Why, and How questions for the origin of the data.
-Include any additional information that would be useful in understanding the context in which the data are embedded.
+## Analysis Objectives
+
+### Q1: What predictors, if any, increases an individual's chance of receiving a liver transplant?
+
+### Q2: Are there factors that can help distinguish different stages of Biliary Cholangitis not currently included in medical definition?
+
+## Clinical Study + Dataset Background :
+
+The dataset is produced from a clinical study by Mayo Clinic run from 1974-1984.
+The final 'Status' of each patient was observed in 1986.
+'Status' was either Dead, Censored, or Censored due to liver transplant.
+
+A link to the original study is here: https://faculty.washington.edu/abansal/ShortCourse_DynamicDecisionMaking/Dickson1989_MayoPBCOriginalArticle.pdf
 
@@
 
+```{r}
+count <- colSums(is.na(cirrhosis))
+
+kable(count)
+```
 - Engage in Exploratory Data Analysis
@@
 - A commented Code Appendix
+
+```{r}
+
+cirrhosis %>% 
+  ggplot(aes(x = N_Days)) + geom_histogram() +
+  facet_wrap(facets = vars(Status))
+
+```

### 2023-12-14 EDA.Rmd
hi

@@
 date: "2023-12-12"
-output: html_document
+output: html_notebook
 ---
@@
 
+# EDA
+
+## Exploring the demographics of the patients
+
+### Sex Ratio of Patients
+```{r}
+
+#cirrhosis$Sex <- factor(ifelse(cirrhosis$Sex == 0, "Female", "Male"))
+
+# Preparing the data for pie chart
+sex_data <- cirrhosis %>% 
+            count(Sex) %>% 
+            mutate(Percentage = n / sum(n) * 100)
+
+# Pie Chart
+ggplot(sex_data, aes(x="", y=Percentage, fill=Sex)) +
+    geom_bar(stat="identity", width=1) +
+    coord_polar("y", start=0) +
+    scale_fill_brewer(palette="Pastel1") +
+    theme_minimal() +
+    theme(axis.line = element_blank(),
+          axis.text = element_blank(),
+          axis.ticks = element_blank(),
+          panel.grid = element_blank(),
+          plot.title = element_text(size=14, face="bold")) +
+    labs(fill="Sex", title=" Sex Amongst Patients")
+
+```
+Primary Biliary Cholangitis affects women much more than men at a 10-1 ratio.
+The data of this clinical study mimics the greater population.
+Investigating gender-based differences in disease progression can uncover any gender-specific patterns in PBC.
+This could lead to gender-tailored treatment approaches and a better understanding of the disease's biology, which might differ between males and females.
+
+### Age Distribution of Patients
+```{r}
+# Converting age from days to years for better readability
+cirrhosis$Age_Years <- cirrhosis$Age / 365.25
+
+# Histogram
+ggplot(cirrhosis, aes(x=Age_Years)) +
+    geom_histogram(binwidth=5, fill="#69b3a2", color="#e9ecef") +
+    theme_minimal() +
+    labs(title="Age Distribution of PBC Patients", x="Age (years)", y="Count") +
+    theme(plot.title = element_text(size=14, face="bold"))
+
+```
+The distribution of ages is quite normal.
+
+## Looking at the data itself
+
+### NA counts amongst our variables
+```{r}
+count <-  data.frame(NA_CountsPerVar = colSums(is.na(cirrhosis)))
+kable(count)
+```
+An observation is that many of the variables have 106 NAs. This indicates that a good fraction [106 patients] were equally tracked and measured in a less extensive way.
+
+## Distribution of Stages of Biliary Cholangitis
+```{r}
+
+cirrhosis %>%
+      ggplot(aes(x = Stage)) + geom_bar(fill = "#097969") # need color
+
+```
+This distribution is left skewed and not symmetric. Most patients have stage 3 and then 4 of Biliary Cholangitis.
+
+## Exploring some interactions of variables
+
+### How does prescense of a D-penicillamine with stage of Biliary Cholangitis impact number of days till death?
+```{r}
+
+ggplot(data = cirrhosis, aes (Drug, Stage, fill = N_Days))+ geom_tile() + scale_fill_distiller(palette = "RdPu", trans = "reverse")
+
+```
+This heat map shows that being a placebo in stage 1 gives you a greater amount of days till death in this sample of patients.
+
+### Individuals with Edema, and if they were on the drug
+```{r}
+
+hi = c("#40B5AD", "#009E60", "#9FE2BF")
+library(ggmosaic)
+cirrhosis %>%
+  ggplot() +
+  geom_mosaic(aes( x = product(Edema), fill = Drug)) + scale_fill_manual(values = hi)
+```
+
+Edema is swelling due to too much liquid trapped in the body's tissues.
+It's common complication in liver diseases and can significantly impact patient quality of life and survival.
+Understanding how different levels of edema (none, controlled by diuretics, or persistent despite treatment) affect survival can inform patient management strategies and highlight the need for aggressive interventions in certain cases.
+From the graph we can see a even split between the placebo and drug for Edema status.
+There is no NA values for Edema persistent despite diuretics.
+We cannot  directly compare it the splits for other statuses given their NA values if known could change the look of the graph.
+
+### mihir choose your preffered graph and explain
+
 ### Survival Analysis Based on Treatment and Stage: How does the survival rate differ among patients at different stages of PBC who received D-penicillamine versus placebo?
@@
 This information is vital for clinicians to make informed decisions about treatment plans and for researchers to understand the drug's impact.
+(LOLL this is what the study acc sought to discover)
 
@@
 Such correlations can aid in the early detection of disease severity, help in monitoring the disease progression, and potentially guide treatment adjustments.
-
-### Sex Distribution of Patients
-```{r}
-library(ggplot2)
-library(dplyr)
-
-cirrhosis$Sex <- factor(ifelse(cirrhosis$Sex == 0, "Female", "Male"))
-
-# Preparing the data for pie chart
-sex_data <- cirrhosis %>% 
-            count(Sex) %>% 
-            mutate(Percentage = n / sum(n) * 100)
-
-# Pie Chart
-ggplot(sex_data, aes(x="", y=Percentage, fill=Sex)) +
-    geom_bar(stat="identity", width=1) +
-    coord_polar("y", start=0) +
-    scale_fill_brewer(palette="Pastel1") +
-    theme_minimal() +
-    theme(axis.line = element_blank(),
-          axis.text = element_blank(),
-          axis.ticks = element_blank(),
-          panel.grid = element_blank(),
-          plot.title = element_text(size=14, face="bold")) +
-    labs(fill="Sex", title="Distribution of Sex in PBC Patients")
-
-```
-Primary Biliary Cholangitis affects women much more than men at a 10-1 ratio.
-The data of this clinical study mimics the greater population.
-Investigating gender-based differences in disease progression can uncover any gender-specific patterns in PBC.
-This could lead to gender-tailored treatment approaches and a better understanding of the disease's biology, which might differ between males and females.
-
-### Age Distribution of Patients
-
-```{r}
-# Converting age from days to years for better readability
-cirrhosis$Age_Years <- cirrhosis$Age / 365.25
-
-# Histogram
-ggplot(cirrhosis, aes(x=Age_Years)) +
-    geom_histogram(binwidth=5, fill="#69b3a2", color="#e9ecef") +
-    theme_minimal() +
-    labs(title="Age Distribution of PBC Patients", x="Age (years)", y="Count") +
-    theme(plot.title = element_text(size=14, face="bold"))
-
-```
-The distribution of ages is quite normal.
-
-### might have to mutate data below to filter out alive ppl but get the gist
-```{r}
-
-ggplot(data = cirrhosis, aes (Drug, Stage, fill = N_Days))  + geom_tile() + scale_fill_distiller(palette = "RdPu")
-
-```
-
-### Individuals with Edema and if they were on the drug
-```{r}
-
-library(ggmosaic)
-cirrhosis %>%
-  ggplot() +
-  geom_mosaic(aes( x = product(Edema), fill = Drug))
-#ggplot(data = cirrhosis) + geom_mosaic(aes(x = Drug , fill = Status))
-
-```
-Looking at the dist. of stages
-```{r}
-
-cirrhosis %>%
-      ggplot(aes(x = Stage)) + geom_bar() # need color
-
-```
-
-```{r}
-cirrhosis %>%
-      ggplot(aes(x = Cholesterol, y = Bilirubin)) + geom_line()
-
-```
-
-### Comparison of Edema Status on Patient Outcomes
-
-Edema is a common complication in liver diseases and can significantly impact patient quality of life and survival.
-Understanding how different levels of edema (none, controlled by diuretics, or persistent despite treatment) affect survival can inform patient management strategies and highlight the need for aggressive interventions in certain cases.

### 2023-12-14 Final Project.Rmd
hi

@@
 
+# ASK MIHIR TO PUT IN IMAGES HERE !!!
+
 ## Analysis Objectives
@@
 
-### Q2: Are there factors that can help distinguish different stages of Biliary Cholangitis not currently included in medical definition?
+### Q2: Are there factors not currently included in medical definition that can help distinguish different stages of Biliary Cholangitis?
 
@@
 
-- Data Preprocessing
+### Load in Packages
+```{r}
 
+library(survival)
+library(ggplot2)
+library(survival)
+library(survminer)
+library(kableExtra)
+library(factoextra)
+
+```
+
+### Data Processing
 ```{r}
@@
 
-# Pre-process the data
+### Pre-process the data
 ```{r}
@@
 cirrhosis$Sex <- factor(cirrhosis$Sex)
-cirrhosis$Ascites <- factor(cirrhosis$Ascites)
-cirrhosis$Hepatomegaly <- factor(cirrhosis$Hepatomegaly)
-cirrhosis$Spiders <- factor(cirrhosis$Spiders)
+#cirrhosis$Ascites <- factor(cirrhosis$Ascites)
+#cirrhosis$Hepatomegaly <- factor(cirrhosis$Hepatomegaly)
+#cirrhosis$Spiders <- factor(cirrhosis$Spiders)
 cirrhosis$Edema <- factor(cirrhosis$Edema)
-#need to do more
+cirrhosis$Sex <- factor(ifelse(cirrhosis$Sex == 0, "Female", "Male"))
+
 #Drop the first column
@@
 
-```{r}
-count <- colSums(is.na(cirrhosis))
-
-kable(count)
-```
 - Engage in Exploratory Data Analysis

### 2023-12-14 Unsupervised.Rmd
Added basic code - EDA, importing data, tree and logistic regression
hi

--- 2023-12-13 Regression.Rmd
@@
 ---
-title: "Regression-based"
+title: "Unsupervised"
 author: "Mihir Kulkarni"
 date: "2023-12-12"
-output: html_document
+output: html_notebook
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+# Unsupevised Learning
+
+ In this section, we sought to if there are sub-collections of patients.
+We expect sub-collections to mimic medical definitions of the stages of Biliary Cirrhosis but are curious to see if clustering can reveal some newer observations.
+
+### Stages of Biliary Cirrhosis
+
+Stage 1: There's inflammation and damange to the walls of medium-sized bile ducts.
+
+Stage 2: There's blockage of the small bile ducts.
+
+Stage 3: This stage marks the beginning of scarring.
+
+Stage 4: Cirrhosis has developed. This permanent, severe, scarring and damange to the liver.
+
+Source of information: https://www.healthline.com/health/primary-biliary-cirrhosis#stages
+
+#do the ml, take about observational stages vs now quantify, look into the different variables and what they mean.
+
+### Data Wrangling
+```{r}
+
+cirrhosisCluster <- cirrhosis
+
+cirrhosisCluster <- cirrhosisCluster %>%                 
+                                  select(-Stage)
+
+cirrhosisCluster <- na.omit(cirrhosisCluster) #cannot have NA values in clustering
+
+cirrhosisCluster$Age_Years <- round(cirrhosisCluster$Age_Years) #round age to whole numbers
+
+#reode sex... recode others later if need be #female is 0
+
+#cirrhosisCluster  <- cirrhosisCluster %>%
+#  select( -c(Status, Drug, Edema)) %>%
+
+# mutate(Sex = ifelse(Sex == 'Female', 0, 1))
+
+  cirrhosisCluster  <- cirrhosisCluster %>%
+    mutate(Sex = ifelse(Sex == 'Female', 0, 1)) %>%
+          mutate(Transplant = ifelse(Status == "CL", 1, 0)) %>%
+              mutate(Status = ifelse(Status %in% c('C', 'CL'), 0 , 1)) %>%
+                    mutate(Drug = ifelse(Drug == "D-penicillamine", 0, 1)) %>%
+                        mutate(EdemaDiurectics = ifelse(Edema %in% c('S', 'Y'), 1, 0)) %>%
+                              mutate(NoEdemaORD = ifelse(Edema == 'N' , 1, 0)) %>%
+                                        mutate(EdemaANDD = ifelse(Edema == "Y", 1, 0)) %>%
+                                                mutate(EdemaORD = ifelse(Edema == "S", 1, 0)) %>%
+                                                                              select(-Edema)
+
+#Data is already factored
+
+#mutate(Sex = ifelse(Sex == 'Female', 0, 1)) %>%
+
+   #mutate(Status = ifelse(Status %in% c('C', 'CL'), 0 , 1)) %>%
+          #    mutate(Drug = ifelse(Drug == "D-penicillamine", 0, 1))
 ```
 
-## R Markdown
+#make column yes no edema
+#then yes no under treatment
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+#same for tranplant stuff
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+```{r}
+distCirrhosis <- dist(
+  x = cirrhosisCluster,
+  method = "euclidean"
+)
 
-```{r cars}
-summary(cars)
 ```
 
-## Including Plots
+### A mixed hierarchial and non-hierarchial
 
-You can also embed plots, for example:
+```{r}
 
-```{r pressure, echo=FALSE}
-plot(pressure)
+hybridCirrhosis <- hkmeans(
+  x = cirrhosisCluster,
+  k = 4,
+  hc.metric = "euclidean",
+  hc.method = "ward.D",
+  iter.max = 10
+)
+
 ```
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+### Visualize Results
+
+### Making a color palette
+```{r}
+
+StagesPalette <- c("#AA336A", "#770737", "#40B5AD", "#009E60")
+
+```
+
+```{r}
+
+## MAKE A NEW PALLETE TO VISUALIZE
+
+# Plot the initial dendrogram for hybrid approach ----
+set.seed(380)
+hkmeans_tree(
+  hkmeans = hybridCirrhosis,
+  rect.col = StagesPalette,
+  cex = 0.4,
+  main = "Initial Hierarchical Clusters"
+)
+
+```
+
+```{r}
+# Plot the final dendrogram for hybrid approach ----
+# library(factoextra)
+fviz_dend(
+  x = hybridCirrhosis,
+  cex = 0.4,
+  palette = StagesPalette,
+  rect = FALSE,
+  horiz = TRUE,
+  repel = TRUE,
+  main = "Final Dendrogram"
+)
+
+```
+
+```{r}
+
+# Plot the final clustering for hybrid approach ----
+# library(factoextra)
+fviz_cluster(
+  object = hybridCirrhosis,
+  stand = FALSE,
+  geom = "point",
+  main = "Hybrid Cluster Plot"
+) +
+  scale_color_manual(values = StagesPalette) +
+  scale_fill_manual(values = StagesPalette) +
+  theme_bw()
+
+```

### 2023-12-15 Unsupervised.Rmd
yuh

@@
 
+## DO I NEED THIS DIST??
 ```{r}
@@
 
-StagesPalette <- c("#AA336A", "#770737", "#40B5AD", "#009E60")
+StagesPalette <- c("#AA336A", "#770737", "#40B5AD", "#009E60", "#9FE2BF")
 
@@
 ```
+### Testing for optimal k value
+
+```{r}
+# Create scree plot for choosing k ----
+library(factoextra)
+set.seed(380)
+fviz_nbclust(
+  x = cirrhosisCluster,
+  diss = NULL,
+  FUNcluster = kmeans,
+  method = "wss",
+  k.max = 10
+)
+
+```
+
+### Trying k = 5
+
+```{r}
+
+hybridCirrhosis2 <- hkmeans(
+  x = cirrhosisCluster,
+  k = 5,
+  hc.metric = "euclidean",
+  hc.method = "ward.D",
+  iter.max = 10
+)
+
+```
 
@@
 fviz_dend(
-  x = hybridCirrhosis,
+  x = hybridCirrhosis2,
   cex = 0.4,
@@
 ```
+
+```{r}
+
+# Plot the final clustering for hybrid approach ----
+# library(factoextra)
+fviz_cluster(
+  object = hybridCirrhosis2,
+  stand = FALSE,
+  geom = "point",
+  main = "Hybrid Cluster Plot"
+) +
+  scale_color_manual(values = StagesPalette) +
+  scale_fill_manual(values = StagesPalette) +
+  theme_bw()
+
+```

### 2023-12-15 EDA.Rmd
Started working on Logistical model

@@
 #Required libraries
-
 library(survival)
@@
 ### Sex Ratio of Patients
+
 ```{r}
@@
 ```
+
 Primary Biliary Cholangitis affects women much more than men at a 10-1 ratio.
@@
 ### Age Distribution of Patients
+
 ```{r}
@@
 ```
+
 The distribution of ages is quite normal.
@@
 ### NA counts amongst our variables
+
 ```{r}
@@
 ```
+
 An observation is that many of the variables have 106 NAs. This indicates that a good fraction [106 patients] were equally tracked and measured in a less extensive way.
@@
 ## Distribution of Stages of Biliary Cholangitis
+
 ```{r}
@@
 ```
+
 This distribution is left skewed and not symmetric. Most patients have stage 3 and then 4 of Biliary Cholangitis.
@@
 ### How does prescense of a D-penicillamine with stage of Biliary Cholangitis impact number of days till death?
+
 ```{r}
@@
 ```
+
 This heat map shows that being a placebo in stage 1 gives you a greater amount of days till death in this sample of patients.
@@
 ### Individuals with Edema, and if they were on the drug
+
 ```{r}

### 2023-12-15 Regression.Rmd
h
b
Started working on Logistical model

--- 2023-12-13 Regression.Rmd
@@
 
-# Part II: Prediction for the Number of Days until an event (death)
+# Part II: Exploring the predictors for a Liver Transplant using logistic regression
 
-Next, let us address our next sub-goal: predicting the number of days a patient survives.
-The objective is to use the medical test data help doctors make better decisions.
-To do this, we will be using linear regression.
+Next, let us explore our question with a different type of model. To do this, we will be using (binary) logistic regression model
 
@@
 
-Linear regression is a statistical technique to model the relationship between a dependent variable and one or more independent variables.
-The model tries to draw a straight line that best fits the data points on a graph, representing how changes in the independent variables are associated with changes in the dependent variable.
-The primary advantage of linear regression is its simplicity and ease of interpretation.
-However, its main drawbacks are that it assumes a linear relationship between variables, which may not always be true.
-This type of model can be very sensitive to outliers, which affects the model's accuracy, especially dealing with medical data.
+Logistic regression is a statistical method used for binary classification, which predicts the probability of an outcome that can be either true or false.
+This is done by understanding the relationship between a dependent binary variable and one or more independent variables.
+Logistic regression is easy to implement and interpret.
+However there are also some drawbacks, the model assumes a linear relationship between the independent variables and the log odds of the dependent variable, which may not always hold true in complex real-world scenarios.
 Furthermore, unlike the decision tree, linear regression models cannot ignore N/A values.
 
-![[Linear Regression Example (]{.underline}[Source](https://www.spiceworks.com/tech/artificial-intelligence/articles/what-is-linear-regression/))](LinRegEg.png)
+![Logistic Regression Model ([Source](https://www.spiceworks.com/tech/artificial-intelligence/articles/what-is-logistic-regression/))](46-4.png)
 
@@
 Similar to the tree, we will start by splitting the data set into training and testing sets.
-Note that the data now onluy contains instances where the patient was deceased.
+Note that the data now only contains instances where the patient was deceased.
 The training set will be used to train our model, while the testing set will help evaluate its performance.
@@
 
-### Step 1: Train basic tree
+### Dataset for Regression
+
+```{r}
+cirrhosisRegression <- cirrhosis
+```
+
+```{r}
+
+cirrhosisRegression <- cirrhosisRegression %>%
+                                        filter(Status != "D") %>%
+                                                          mutate(Transplant = ifelse(Status == 'C', 0, 1 ))
+
+```
+
+### Logistic regression model
+
+```{r}
+
+TransplantModel <- glm()
+
+```
+
+======= \### Step 1: Train basic tree
 

### 2023-12-15 Regression.Rmd
added old tree code

--- 2023-12-13 Regression.Rmd
@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+# Part II: Exploring the predictors for a Liver Transplant using logistic regression
+
+Next, let us explore our question with a different type of model. To do this, we will be using (binary) logistic regression model
+
+## Introduction
+
+Logistic regression is a statistical method used for binary classification, which predicts the probability of an outcome that can be either true or false.
+This is done by understanding the relationship between a dependent binary variable and one or more independent variables.
+Logistic regression is easy to implement and interpret.
+However there are also some drawbacks, the model assumes a linear relationship between the independent variables and the log odds of the dependent variable, which may not always hold true in complex real-world scenarios.
+Furthermore, unlike the decision tree, linear regression models cannot ignore N/A values.
+
+![Logistic Regression Model ([Source](https://www.spiceworks.com/tech/artificial-intelligence/articles/what-is-logistic-regression/))](46-4.png)
+
+## Methodology
+
+Similar to the tree, we will start by splitting the data set into training and testing sets.
+Note that the data now only contains instances where the patient was deceased.
+The training set will be used to train our model, while the testing set will help evaluate its performance.
+We'll use 80% of the data for training and the remaining 20% for testing.
+To allow reproducible code, we have fixed the seed at 380.
+
+```{r}
+#Set up dataset for regression
+cirrhosisRegression <- cirrhosis
+
+#Remove all rows where the patient status is D
+cirrhosisRegression <- cirrhosisRegression %>%
+                                        filter(Status != "D") %>%
+                                                          mutate(Transplant = ifelse(Status == 'C', 0, 1 ))
+
 ```
+
+### Logistic regression model
+
+```{r}
+#model data
+LRmodelData <- cirrhosisRegression %>%
+  drop_na() %>%
+  mutate(
+    tempID = row_number(),
+    .before = Transplant
+  )
+
+# Set seed for reproducibility and slice
+set.seed(380)
+trainingData <- LRmodelData %>%
+  group_by(Transplant) %>%  # group_by() function ensures that the data
+  slice_sample(prop = 0.75)
+
+testingData <- LRmodelData %>%
+  filter(!(tempID %in% trainingData$tempID))
+
+trainingResults <- trainingData
+```
+
+```{r}
+# Form Candidate Model 1
+model1 <- glm(
+  formula = Transplant ~ Platelets,
+  data = trainingData,
+  family = binomial
+)
+```
+
+```{r models}
+# Lower bound (Intercept only)
+lower <- glm(
+  formula = Transplant ~ 1,
+  data = trainingData,
+  family = binomial
+)
+
+# Upper bound 
+upper <- glm(
+  formula = Transplant ~ Status + Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage,
+  data = trainingData,
+  family = binomial
+)
+
+# Stepwise search
+model2 <- step(
+  object = lower,
+  scope = list(
+    lower = lower,
+    upper = upper
+  ),
+  data = trainingData,
+  direction = "both",
+  k = 2
+)
+```
+
+# Results
+
+Initially, we'll delve into the two preliminary models independently to
+understand where they stand. Following that, we'll be deploying the best
+candidate model on our test data. Regarding confusion matrices, we'll
+employ a basic rule: if the predicted probability of a student
+graduating exceeds 0.5, we'll categorize them as having graduated (naïve
+rule).
+
+## Model 1
+
+```{r}
+# Model 1 Coefficient Table
+as.data.frame(summary(model1)$coefficients) %>%
+  rownames_to_column(var = "term") %>%
+  rename(coefficient = Estimate) %>% 
+  mutate(
+    prob_odds = case_when(
+      coefficient == "(Intercept)" ~ exp(coefficient)/(1 + exp(coefficient)),
+      .default = exp(coefficient)
+    ),
+    .after = coefficient
+  ) %>%
+  mutate(
+    `Pr(>|z|)` = ifelse(
+      test = `Pr(>|z|)` < 0.001,
+      yes = paste("< 0.001"),
+      no = `Pr(>|z|)`
+    ),
+    term = case_when(
+      term == "(Intercept)" ~ "Intercept",
+      grepl(x = term, pattern = "Platelets") ~ "Platelets"
+    )
+  ) %>%
+  kable()
+
+```
+
+```{r}
+library(janitor)
+# Building confidence intervals for Model 1 coefficients
+model1CI <- confint(
+  object = model1,
+  parm = "Platelets",
+  level = 0.9
+)
+
+trainingResults <- trainingData %>%
+  ungroup() %>%
+  mutate(model1Pred = predict(object = model1, newdata = ., type = "response"))
+
+# Apply naïve rule ----
+trainingResults <- trainingResults %>%
+  mutate(
+    model1Class = case_when(
+      model1Pred > 0.5 ~ 1,
+      .default = 0
+    )
+  )
+
+#Confusion Matrix for Model 1
+trainingResults %>%
+  mutate(Transplant_status = ifelse(Transplant == 1, 1, 0)) %>%
+  tabyl(var1 = model1Class, var2 = Transplant_status) %>%
+  adorn_title(
+    placement = "combined",
+    row_name = "Predicted",
+    col_name = "Actual"
+  ) %>%
+  kable(
+    booktabs = TRUE,
+    align = "c",
+    caption = "Model 1 Confusion Matrix"
+  )
+
+```
+
+Next let us look at our Model 2, which has multiple factors as discussed
+earlier.
+
+```{r}
+#Coeff for model 2
+as.data.frame(summary(model2)$coefficients) %>%
+  tail(11) %>%
+  rename(coefficient = Estimate) %>% 
+  mutate(
+    prob_odds = case_when(
+      coefficient == "(Intercept)" ~ exp(coefficient)/(1 + exp(coefficient)),
+      TRUE ~ exp(coefficient)
+    ),
+    .after = coefficient
+  ) %>%
+  kable()
+
+```
+
+```{r}
+#do the Tukey-Anscombe plo
+ggplot(
+  data = data.frame(
+    residuals = residuals(model2, type = "pearson"),
+    fitted = fitted(model2)
+  ),
+  mapping = aes(x = fitted, y = residuals)
+) +
+  geom_point() +
+  geom_smooth(
+    formula = y ~ x,
+    method = stats::loess,
+    method.args = list(degree = 1),
+    se = FALSE,
+    linewidth = 0.5
+  ) +
+  theme_bw() +
+  labs(
+    x = "Fitted",
+    y = "Pearson Residuals"
+  )
+```
+
+```{r}
+#plot the gvif
+as.data.frame(car::vif(model2)) %>%
+  kable(
+    digits = 3,
+    align = "lcccc",
+    booktab = TRUE,
+    format.args = list(big.mark = ","),
+    table.attr = 'data-quarto-disable-processing="true"',
+    label = "GVIF analsyis"
+  )
+
+```
+
+Here we can see the generalized variance inflation factors for candidate
+Model 2. Multicollinearity is a statistical concept where several
+independent variables in a model are correlated and this is an important
+consideration especially when we deal with so many categories.
+
+```{r}
+#Store the predicted and actual values for Model 2
+trainingResults$model2Pred <- predict(model2, type = "response")
+trainingResults$model2Class <- ifelse(trainingResults$model2Pred > 0.5, 1, 0)
+trainingResults$Actual <- ifelse(trainingData$graduated == 1, 1, 0)
+
+# Create confusion matrix using table
+confusionMatrix <- table(Predicted = trainingResults$model2Class, Actual = trainingResults$Actual)
+
+kable(confusionMatrix, caption = "Confusion matrix for Model 2") %>%
+  kable_classic(latex_options = "HOLD_position")
+
+```
+
+Lastly, let us look at the separation plots for each of the models.
+
+```{r}
+# Fit ROC Curves for later
+## Model 1
+model1ROC <- roc(
+  formula = graduated ~ model1Pred,
+  data = trainingResults
+)
+model1ROC_df <- data.frame(
+  threshold = model1ROC$thresholds,
+  sensitivity = model1ROC$sensitivities,
+  specificity = model1ROC$specificities,
+  model = "Model 1"
+)
+## Model 2
+model2ROC <- roc(
+  formula = graduated ~ model2Pred,
+  data = trainingResults
+)
+model2ROC_df <- data.frame(
+  threshold = model2ROC$thresholds,
+  sensitivity = model2ROC$sensitivities,
+  specificity = model2ROC$specificities,
+  model = "Model 2"
+)
+```
+
+```{r}
+# Convert 'Actual' column to numeric 0/1
+trainingResults <- trainingResults %>%
+  mutate(
+    actualNum = if_else(Actual == 0, 0, 1)
+  )
+
+#Sepeation Plot
+par(mar = c(4,0,0,0))
+separationplot(
+  pred = trainingResults$model1Pred, 
+  actual = trainingResults$actualNum, 
+  type = "rect",
+  line = TRUE, 
+  lwd2 = 2,
+  show.expected = TRUE, 
+  newplot = FALSE,
+  heading = "Model 1"
+)
+
+```
+
+```{r}
+#Sepeation Plot
+par(mar = c(4,0,0,0))
+separationplot(
+  pred = trainingResults$model2Pred, 
+  actual = trainingResults$actualNum, 
+  type = "rect",
+  line = TRUE, 
+  lwd2 = 2,
+  show.expected = TRUE, 
+  newplot = FALSE,
+  heading = "Model 2"
+)
+```
+
+The separation plot assesses the the fit of the model by providing the
+model's ability to predict occurrences with a high probability and
+non-occurrences with low probability. The separation plot above
+separation plot suggests that Model 2 has a reasonably good performance
+in predicting graduates, especially for the observations on the
+left-most side of the plot compared to Model 1 with the training data.
+We will later compare this graph with the testing data.
+
+Lastly, let us look at the ROC curves for bothe the models.
+
+```{r}
+## Merge into existing data frame
+rocData <- rbind(model1ROC_df, model2ROC_df)
+
+## AUC Data
+aucData <- data.frame(
+  model = c("Model 1", "Model 2"),
+  auc = c(model1ROC$auc, model2ROC$auc)
+)
+```
+
+```{r}
+#ROC plot
+ggplot(
+  data = rocData,
+  mapping = aes(x = 1 - specificity, y = sensitivity, color = model)
+) +
+  geom_path() +
+  geom_abline(
+    slope = 1,
+    intercept = 0,
+    linetype = "dotted"
+  ) +
+  geom_text(
+  inherit.aes = FALSE,
+  data = aucData,
+  mapping = aes(label = paste(model, "AUC: \n", round(auc, 3))),
+  x = c(0.25, 0.15),
+  y = c(0.4, 1.05)
+)
+```
+
+From the graphs we can interpret that:
+
+Model 1: - Its ROC curve is above the line of no discrimination,
+indicating that the model has some predictive capabilities. - The AUC is
+0.621, which is better than random guessing but suggests there's room
+for improvement.
+
+Model 2: - The ROC curve for Model 2 is significantly above that of
+Model 1, and much closer to the top-left corner, indicating better
+predictive performance. - The AUC is 0.768, which suggests a "good"
+classification performance, and it's notably better than Model 1. - Its
+ability to discriminate between positive and negative classes is
+superior to that of Model 1.
+
+Lastly, let us plot our influence plot.
+
+```{r}
+# Influence Plot for Model 2
+idCases <- car::influencePlot(model2)
+```
+
+The influence plot shows several data points with high leverage and
+large residuals, indicating potential outliers. Some observations,
+notably those labeled like "23881," have significant influence on the
+regression model due to their Cook's D values.
+
+# Testing our model
+
+Using our final model, we now turn to see how well this classifier does
+on our testing data. Recall that we initally set the test data during
+the train test split.
+
+The confusion matrix below shows the performance of our model using the
+naïve decision rule.
+
+```{r}
+# Set up testing data results
+testingData <- testingData %>%
+  mutate(
+    gradNum = case_when(
+      graduated == 0 ~ 0,
+      graduated == 1 ~ 1
+    ),
+    .after = graduated
+  )
+testingData$predict <- predict(
+  object = model2,
+  newdata = testingData,
+  type = "response"
+)
+testingData <- testingData %>%
+  mutate(
+    model2Class = case_when(
+      predict > 0.5 ~ 1,
+      .default = 0
+    )
+  )
+```
+
+```{r}
+testingData$graduated <- ifelse(testingData$graduated == 1, 1, 0)
+
+# Build Confusion Matrix for Testing Data
+testingData %>%
+  tabyl(var1 = model2Class, var2 = graduated) %>%
+  adorn_title(
+    placement = "combined",
+    row_name = "Predicted",
+    col_name = "Actual"
+  ) %>%
+  kable(
+    caption = "Confusion Matrix for Test data"
+  )
+```
+
+From the matrix, we can see that the model has the following metrics:
+
+-   True Positive (TP): Predicted Graduated and actually Graduated. In
+    this case, 7251.
+-   False Positive (FP): Predicted Graduated but actually Not Graduated.
+    In this case, 3505.
+-   False Negative (FN): Predicted Not Graduated but actually Graduated.
+    In this case, 1936.
+-   True Negative (TN): Predicted Not Graduated and actually Not
+    Graduated. In this case, 5395.
+
+From this, we can interpret that.
+
+-   Accuracy: Approximately 69.92%
+-   Precision: Approximately 67.41%
+-   Recall (Sensitivity): Approximately 78.93%
+-   F1 Score: Approximately 72.72%
+
+This shows that our model is still capable with new data, with a slight
+reduction to the overall performance.
+
+```{r}
+#Sepeation Plot
+par(mar = c(4,0,0,0))
+separationplot(
+  pred = testingData$predict, 
+  actual = testingData$gradNum, 
+  type = "rect",
+  line = TRUE, 
+  lwd2 = 2,
+  show.expected = TRUE, 
+  newplot = FALSE
+)
+
+```

### 2023-12-15 Tree.Rmd
added old tree code

@@
 
--   Class 0 (D): The patient didn't survive.
-
-```{=html}
-<!-- -->
-```
--   Class 1 (C): The patient is censored, meaning that the observation period ended without the death being recorded.
-
--   Class 2 (CL): Similar to Class 1, the patient is censored due to liver transplantation.
+-   Class 0 (D): The patient didn't survive by the end of the observation
+-   Class 1 (C): The patient is censored, meaning that the observation period ended without the death being recorded
+-   Class 2 (CL): Similar to Class 1, the patient is censored due to liver transplantation
 
@@
   tree = rPartStatus,
-  cp = 0.016
+  cp = 0.029
 )
@@
 # Wrangle the Graduate data to set up training and testing datasets
-modelCirrhosis2 <- cirrhosis %>%
+modelCirrhosis2 <- cirrhosisTreeData %>%
   drop_na() %>%
@@
 upper <- glm(
-  formula = Status ~ N_Days + Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage,
+  formula = Status ~ Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage,
   data = trainingData,

### 2023-12-15 Regression.Rmd
Re-did logistic regression with new goals

@@
 ```{r}
-#Set up dataset for regression
 cirrhosisRegression <- cirrhosis
-
-#Remove all rows where the patient status is D
-cirrhosisRegression <- cirrhosisRegression %>%
-                                        filter(Status != "D") %>%
-                                                          mutate(Transplant = ifelse(Status == 'C', 0, 1 ))
-
+cirrhosisRegression$Status <- ifelse(cirrhosisRegression$Status == "C" | cirrhosisRegression$Status == "CL", 1, 0)
 ```
@@
     tempID = row_number(),
-    .before = Transplant
+    .before = Status
   )
@@
 trainingData <- LRmodelData %>%
-  group_by(Transplant) %>%  # group_by() function ensures that the data
+  group_by(Status) %>%  # group_by() function ensures that the data
   slice_sample(prop = 0.75)
@@
 model1 <- glm(
-  formula = Transplant ~ Platelets,
+  formula = Status ~ Platelets,
   data = trainingData,
@@
 lower <- glm(
-  formula = Transplant ~ 1,
+  formula = Status ~ 1,
   data = trainingData,
@@
 upper <- glm(
-  formula = Transplant ~ Status + Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage,
+  formula = Status ~ Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage,
   data = trainingData,
@@
 trainingResults %>%
-  mutate(Transplant_status = ifelse(Transplant == 1, 1, 0)) %>%
-  tabyl(var1 = model1Class, var2 = Transplant_status) %>%
+  mutate(Patient_status = ifelse(Status == 1, 1, 0)) %>%
+  tabyl(var1 = model1Class, var2 = Patient_status) %>%
   adorn_title(
@@
 as.data.frame(summary(model2)$coefficients) %>%
-  tail(11) %>%
   rename(coefficient = Estimate) %>% 
@@
 ```{r}
-#do the Tukey-Anscombe plo
+#do the Tukey-Anscombe plot
 ggplot(
@@
 trainingResults$model2Class <- ifelse(trainingResults$model2Pred > 0.5, 1, 0)
-trainingResults$Actual <- ifelse(trainingData$graduated == 1, 1, 0)
+trainingResults$Actual <- ifelse(trainingData$Status == 1, 1, 0)
 
@@
 ```{r}
+library(pROC)
+library(separationplot)
 # Fit ROC Curves for later
@@
 model1ROC <- roc(
-  formula = graduated ~ model1Pred,
+  formula = Status ~ model1Pred,
   data = trainingResults
@@
 model2ROC <- roc(
-  formula = graduated ~ model2Pred,
+  formula = Status ~ model2Pred,
   data = trainingResults
@@
     gradNum = case_when(
-      graduated == 0 ~ 0,
-      graduated == 1 ~ 1
+      Status == 0 ~ 0,
+      Status == 1 ~ 1
     ),
-    .after = graduated
+    .after = Status
   )
@@
 ```{r}
-testingData$graduated <- ifelse(testingData$graduated == 1, 1, 0)
+testingData$Status <- ifelse(testingData$Status == 1, 1, 0)
 
@@
 testingData %>%
-  tabyl(var1 = model2Class, var2 = graduated) %>%
+  tabyl(var1 = model2Class, var2 = Status) %>%
   adorn_title(

### 2023-12-15 Unsupervised.Rmd
Re-did logistic regression with new goals

@@
 
-# Unsupevised Learning
+# Part III: Exploring the sub-clusters of the cirrhosis data using unsupervised learning
 
@@
 
-### Stages of Biliary Cirrhosis
+## Introduction
 
-Stage 1: There's inflammation and damange to the walls of medium-sized bile ducts.
+Biliary Cirrhosis patients are typically clustered in four main clusters ([Source](https://www.healthline.com/health/primary-biliary-cirrhosis#stages){.uri}):
 
-Stage 2: There's blockage of the small bile ducts.
+-   Stage 1: There's inflammation and damage to the walls of medium-sized bile ducts.
 
-Stage 3: This stage marks the beginning of scarring.
+-   Stage 2: There's blockage of the small bile ducts.
 
-Stage 4: Cirrhosis has developed. This permanent, severe, scarring and damange to the liver.
+-   Stage 3: This stage marks the beginning of scarring.
 
-Source of information: https://www.healthline.com/health/primary-biliary-cirrhosis#stages
+-   Stage 4: Cirrhosis has developed. This permanent, severe, scarring and damage to the liver.
 
+In this section we will be using clustering on our data.
+K-means Clustering is an unsupervised learning technique where data points are grouped based on their similarities.
+It's commonly used to identify patterns and structures within datasets without prior knowledge of the groups.
+The main advantage of clustering is its ability to discover hidden patterns in data.
+However, a significant drawback is the subjectivity in defining the 'similarity' criteria, which can lead to varying results and interpretations.
+
+[![K-means clustering ([Source](https://medium.datadriveninvestor.com/k-means-clustering-4a700d4a4720))](1*fz-rjYPPRlGEMdTI-RLbDg.png)](https://medium.datadriveninvestor.com/k-means-clustering-4a700d4a4720)
+
+An important motivation for this part of the study is explore how does "Stage" which is typically defined based on medical professionals' observations compares to the data that we have in this data set.
+
+## Pre-processing the data
+
+We had to pre-process the data for the best performance.
+First, we removed the 'Stage' column to avoid using outcome-related features in the unsupervised learning.
+The rows with N/A values were also removed.
+Lastly, we transformed various categorical variables into numeric formats, which is necessary for clustering algorithms.
+
+```{r}
 #do the ml, take about observational stages vs now quantify, look into the different variables and what they mean.
-
-### Data Wrangling
-```{r}
-
 cirrhosisCluster <- cirrhosis
@@
           #    mutate(Drug = ifelse(Drug == "D-penicillamine", 0, 1))
+
+#???:
+
+#make column yes no edema #then yes no under treatment
+
+#same for transplant stuff
 ```
 
-#make column yes no edema
-#then yes no under treatment
-
-#same for tranplant stuff
-
-## DO I NEED THIS DIST??
 ```{r}
@@
 
-### A mixed hierarchial and non-hierarchial
+## Methodology
+
+To apply the k-means algorithm.
+A mixed hierarchical and non-hierarchical was applied.
+This was done using the hkmeansm module, with k = 4 as our initial value.
+Based on the data properties, the euclidean (L2) distance works best.
 
@@
 
-### Visualize Results
+## Results
 
-### Making a color palette
+We were able to visualize the results of the k-means algorithm. The first step was making a color palette and plotting the dendrogram tree.
+
 ```{r}
@@
 ```{r}
-
 ## MAKE A NEW PALLETE TO VISUALIZE
-
 # Plot the initial dendrogram for hybrid approach ----
@@
 ```
-### Testing for optimal k value
+
+As, you can see, the model was able to create clear clusters for the data.
+However, it is important to find the best value of k to get the optimal clusters.
+To do this we can use a scree plot.
+Here, we\'re looking for the number of clusters that corresponds to the \"elbow\".
 
@@
 
-### Trying k = 5
+From this, we identified that 5 was the ideal value of k, where the Total Within [Cluster] Sums of Squares begins leveling off.
+Let us create a new k-means model with k = 5.
+We can plot this refined model, with a format more easy to visualize.
 
@@
 
+As you can see, the 5 clusters were identified, which are color coded in our updated diagram.
+
+Lastly, let us plot these clusters. Here is a plot of the initial model where k = 4
+
 ```{r}
-
 # Plot the final clustering for hybrid approach ----
@@
   geom = "point",
-  main = "Hybrid Cluster Plot"
+  main = "Hybrid Cluster Plot - Initial model"
 ) +
@@
 ```
+
+We can also plot our refined model, as you can see the plot below identifies 5 clear clusters.
 
@@
   geom = "point",
-  main = "Hybrid Cluster Plot"
+  main = "Hybrid Cluster Plot - Refined model"
 ) +

### 2023-12-15 Final Project.Rmd
refined the models

@@
 
-# ASK MIHIR TO PUT IN IMAGES HERE !!!
+![[Primary Biliary Cholangitis ([Source](https://www.google.com/search?client=safari&sca_esv=591276386&rls=en&sxsrf=AM9HkKmtJVOC2Jrf2BSYVB2O-QcnTSgx6g:1702664705100&q=Primary+Biliary+Cholangitis&tbm=isch&source=lnms&sa=X&ved=2ahUKEwid0qyniJKDAxX9FFkFHVH3CcoQ0pQJegQIDxAB&biw=1440&bih=848&dpr=2#imgrc=UeV_fhP-8vPOlM))]{.underline}](17715-primary-biliary-cholangitis.webp)
 
-## Analysis Objectives
+## Objectives
 
@@
 
-A link to the original study is here: https://faculty.washington.edu/abansal/ShortCourse_DynamicDecisionMaking/Dickson1989_MayoPBCOriginalArticle.pdf
+A link to the original study is here: <https://faculty.washington.edu/abansal/ShortCourse_DynamicDecisionMaking/Dickson1989_MayoPBCOriginalArticle.pdf>
 
 ### Load in Packages
+
 ```{r}
@@
 ### Data Processing
+
 ```{r}
@@
 ### Pre-process the data
+
 ```{r}
@@
 -   Use Machine Learning Algorithms:
+
     -   You must use at least one of each: a Regression-based algorithm, a Tree-based algorithm, and an Unsupervised Learning algorithm.
@@
 
-- A discussion of what you’ve learned from your work about the context and recommendations for future work.
+-   A discussion of what you've learned from your work about the context and recommendations for future work.
 
-- Author Contributions–list out who did what in the project.
+-   Author Contributions--list out who did what in the project.
 
-- References–Citation style is your choice, but all sources should be documented (both in text and in the References section). This includes where you got your data.
+-   References--Citation style is your choice, but all sources should be documented (both in text and in the References section). This includes where you got your data.
 

### 2023-12-15 Regression.Rmd
refined the models

@@
 
-# Part II: Exploring the predictors for a Liver Transplant using logistic regression
+# Part II: Exploring the predictors for the patient's status using logistic regression
 
-Next, let us explore our question with a different type of model. To do this, we will be using (binary) logistic regression model
+Next, let us explore our question with a different type of model and compare the results with our tree. To do this, we will be using (binary) logistic regression model.
 
@@
 
-![Logistic Regression Model ([Source](https://www.spiceworks.com/tech/artificial-intelligence/articles/what-is-logistic-regression/))](46-4.png)
+![[Logistic Regression Model ([Source](https://www.spiceworks.com/tech/artificial-intelligence/articles/what-is-logistic-regression/))]{.underline}](46-4.png)
 
@@
 
+With this, we will build two candidate models:
+
+-   The first model will test the classification based on just the SGOT
+
+-   The second model will use a step wise function using various predictors to see the best performance.
+
+Another important consideration is the application of prediction (estimating an outcome based on input variables) and inference (understanding the relationships between variables).
+We will evaluate the inference through the coefficient analysis and prediction through roc curves and confusion matrix.
+It is important to note that these metrics will complement each other in our understanding of the data.
+However, the main focus of this analysis will be prediction and we will work with several metrics to evaluate it.
+
+We will use a variety of tools to understand the model's performance.
+
 ```{r}
@@
 ```
-
-### Logistic regression model
 
@@
   group_by(Status) %>%  # group_by() function ensures that the data
-  slice_sample(prop = 0.75)
+  slice_sample(prop = 0.80)
 
@@
 model1 <- glm(
-  formula = Status ~ Platelets,
+  formula = Status ~ SGOT,
   data = trainingData,
@@
 ```
+
+Stepwise results:
 
@@
 
-Initially, we'll delve into the two preliminary models independently to
-understand where they stand. Following that, we'll be deploying the best
-candidate model on our test data. Regarding confusion matrices, we'll
-employ a basic rule: if the predicted probability of a student
-graduating exceeds 0.5, we'll categorize them as having graduated (naïve
-rule).
+Initially, we'll delve into the two preliminary models independently to understand where they stand.
+Following that, we'll be deploying the best candidate model on our test data.
+Regarding confusion matrices, we'll employ a basic rule: if the predicted probability of a the patient's status being "deceased" exceeds 0.5, we'll categorize them as deceased (naïve rule).
 
@@
 as.data.frame(summary(model1)$coefficients) %>%
-  rownames_to_column(var = "term") %>%
+  rownames_to_column(var = "X") %>%
   rename(coefficient = Estimate) %>% 
@@
     ),
-    term = case_when(
-      term == "(Intercept)" ~ "Intercept",
-      grepl(x = term, pattern = "Platelets") ~ "Platelets"
+    X = case_when(
+      X == "(Intercept)" ~ "Intercept",
+      grepl(x = X, pattern = "SGOT") ~ "SGOT"
     )
@@
 
+This table shows us the results of our first model.
+We can see that, holding other variables constant, a one-unit increase in SGOT is associated with a decrease in the log-odds of the response variable by 0.0076446.
+Furthermore, the odds-ratio indicates that for each one-unit increase in SGOT, the odds of the event occurring decrease by about 0.76%.
+
+We can also plot the confusion matrix for this model:
+
 ```{r}
@@
   object = model1,
-  parm = "Platelets",
+  parm = "SGOT",
   level = 0.9
@@
     model1Class = case_when(
-      model1Pred > 0.5 ~ 1,
-      .default = 0
+      model1Pred > 0.5 ~ "Censored",
+      .default = "Deceased"
     )
@@
 trainingResults %>%
-  mutate(Patient_status = ifelse(Status == 1, 1, 0)) %>%
+  mutate(Patient_status = ifelse(Status == 1, "Censored", "Deceased")) %>%
   tabyl(var1 = model1Class, var2 = Patient_status) %>%
@@
     caption = "Model 1 Confusion Matrix"
-  )
+  )%>%kable_styling(latex_options = "HOLD_position")
 
@@
 
-Next let us look at our Model 2, which has multiple factors as discussed
-earlier.
+We can see that this model tends to over predict censored values. This shows the need of bringing in more factors. Next let us look at our Model 2, which has multiple factors as discussed earlier.
 
@@
 ```
+
+This analysis suggests significant associations between several predictors (Bilirubin, Age, Alk_Phos, Prothrombin, Spiders) and the status of the patient.
+Each predictor, including the Intercept, shows a statistically significant effect, as indicated by their small p-values (Pr(\>\|z\|)).
+The negative coefficients for Bilirubin, Age, Alk_Phos, Prothrombin, and Spiders imply that increases in these variables are associated with a decrease in the log-odds of the outcome.
+Bilirubin and Spiders are showing notably strong effects in the model.
 
@@
 
+This figure shows us the Tukey-Anscombe plot using Pearson residuals for Model 2.
+In an ideal fit, the residuals should be evenly distributed about zero with constant mean and variance.
+The shape of the line suggests that the model is not capturing someunder lying structure in the data, especially in extreme cases.
+
 ```{r}
@@
 
-Here we can see the generalized variance inflation factors for candidate
-Model 2. Multicollinearity is a statistical concept where several
-independent variables in a model are correlated and this is an important
-consideration especially when we deal with so many categories.
+The Variance Inflation Factor (VIF) values for the variables in the model (Bilirubin, Age, Alk_Phos, Prothrombin, Spiders) are all close to 1, indicating minimal multicollinearity.
+This means that these predictors are relatively independent of each other, enhancing the reliability of the model.
 
@@
 trainingResults$model2Pred <- predict(model2, type = "response")
-trainingResults$model2Class <- ifelse(trainingResults$model2Pred > 0.5, 1, 0)
-trainingResults$Actual <- ifelse(trainingData$Status == 1, 1, 0)
+trainingResults$model2Class <- ifelse(trainingResults$model2Pred > 0.5, "Censored", "Deceased")
+trainingResults$Actual <- ifelse(trainingData$Status == 1, "Censored", "Deceased")
 
 # Create confusion matrix using table
-confusionMatrix <- table(Predicted = trainingResults$model2Class, Actual = trainingResults$Actual)
+confusionMatrixRegression <- table(Predicted = trainingResults$model2Class, Actual = trainingResults$Actual)
 
-kable(confusionMatrix, caption = "Confusion matrix for Model 2") %>%
+kable(confusionMatrixRegression, caption = "Confusion matrix for Model 2") %>%
   kable_classic(latex_options = "HOLD_position")
@@
 ```
+
+From this confusion matrix we can see the relationships between the True Positive, True Negative, False Positive and False Negative values. From this we can calculate:
+
+-   **Accuracy**: Approximately 81.36%
+
+-   **Recall**: Approximately 79.01%
+
+-   **Precision**: Approximately 72.73%
+
+-   **F1 Score**: Approximately 75.74%
 
@@
   mutate(
-    actualNum = if_else(Actual == 0, 0, 1)
+    actualNum = if_else(Actual == "Deceased", 0, 1)
   )
@@
 
-The separation plot assesses the the fit of the model by providing the
-model's ability to predict occurrences with a high probability and
-non-occurrences with low probability. The separation plot above
-separation plot suggests that Model 2 has a reasonably good performance
-in predicting graduates, especially for the observations on the
-left-most side of the plot compared to Model 1 with the training data.
+The separation plot assesses the the fit of the model by providing the model's ability to predict occurrences with a high probability and non-occurrences with low probability.
+The separation plot above separation plot suggests that Model 2 has a reasonably good performance in predicting the patient's status, especially for the observations on the left-most side of the plot compared to Model 1 with the training data.
 We will later compare this graph with the testing data.
@@
 
-Model 1: - Its ROC curve is above the line of no discrimination,
-indicating that the model has some predictive capabilities. - The AUC is
-0.621, which is better than random guessing but suggests there's room
-for improvement.
+Model 1: - Its ROC curve is above the line of no discrimination, indicating that the model has some predictive capabilities.
+- The AUC is 0.64, which is better than random guessing but suggests there's room for improvement.
 
-Model 2: - The ROC curve for Model 2 is significantly above that of
-Model 1, and much closer to the top-left corner, indicating better
-predictive performance. - The AUC is 0.768, which suggests a "good"
-classification performance, and it's notably better than Model 1. - Its
-ability to discriminate between positive and negative classes is
-superior to that of Model 1.
+Model 2: - The ROC curve for Model 2 is significantly above that of Model 1, and much closer to the top-left corner, indicating better predictive performance.
+- The AUC is 0.899, which suggests a good classification performance, and it's notably better than Model 1.
+- Its ability to discriminate between positive and negative classes is superior to that of Model 1.
 
@@
 
-The influence plot shows several data points with high leverage and
-large residuals, indicating potential outliers. Some observations,
-notably those labeled like "23881," have significant influence on the
-regression model due to their Cook's D values.
+The influence plot shows several data points with high leverage and large residuals, indicating potential outliers.
+Some observations, notably those labeled like "149," have significant influence on the regression model due to their Cook's D values.
 
@@
 
-Using our final model, we now turn to see how well this classifier does
-on our testing data. Recall that we initally set the test data during
-the train test split.
+Using our final model, we now turn to see how well this classifier does on our testing data. Recall that we initally set the test data during the train test split.
 
-The confusion matrix below shows the performance of our model using the
-naïve decision rule.
+The confusion matrix below shows the performance of our model using the naïve decision rule.
 
@@
     model2Class = case_when(
-      predict > 0.5 ~ 1,
-      .default = 0
+      predict > 0.5 ~ "Censored",
+      .default = "Deceased"
     )
@@
 ```{r}
-testingData$Status <- ifelse(testingData$Status == 1, 1, 0)
+testingData$Status <- ifelse(testingData$Status == 1, "Censored", "Deceased")
 
@@
 
-From the matrix, we can see that the model has the following metrics:
+-   **Accuracy**: Approximately 71.43%
 
--   True Positive (TP): Predicted Graduated and actually Graduated. In
-    this case, 7251.
--   False Positive (FP): Predicted Graduated but actually Not Graduated.
-    In this case, 3505.
--   False Negative (FN): Predicted Not Graduated but actually Graduated.
-    In this case, 1936.
--   True Negative (TN): Predicted Not Graduated and actually Not
-    Graduated. In this case, 5395.
+-   **Recall**: Approximately 76.92%
 
-From this, we can interpret that.
+-   **Precision**: Approximately 43.48%
 
--   Accuracy: Approximately 69.92%
--   Precision: Approximately 67.41%
--   Recall (Sensitivity): Approximately 78.93%
--   F1 Score: Approximately 72.72%
+-   **F1 Score**: Approximately 55.56%
 
-This shows that our model is still capable with new data, with a slight
-reduction to the overall performance.
+This shows that our model is has struggled with overfitting, with an especially low precision score.
+We will discuss this in the comparison with the tree model, but this is a limitation of our data size.
+Lastly, we can plot the separation plot.
 

### 2023-12-15 Tree.Rmd
refined the models

@@
 ```
-
-```{r}
-#Set up the data
-
-# Wrangle the Graduate data to set up training and testing datasets
-modelCirrhosis2 <- cirrhosisTreeData %>%
-  drop_na() %>%
-  mutate(
-    tempID = row_number(),
-    .before = Status
-  )
-
-## Set seed for reproducibility and slice ----
-set.seed(380)
-trainingData <- modelCirrhosis2 %>%
-  group_by(Status) %>%
-  slice_sample(prop = 0.8)
-
-testingData <- modelCirrhosis2 %>%
-  filter(!(tempID %in% trainingData$tempID))
-```
-
-# Comparison with logistic regression
-
-```{r}
-# Lower bound (Intercept only)
-lower <- glm(
-  formula = Status ~ 1,
-  data = trainingData,
-  family = binomial
-)
-
-# Upper bound 
-upper <- glm(
-  formula = Status ~ Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage,
-  data = trainingData,
-  family = binomial
-)
-
-# Stepwise search
-model2 <- step(
-  object = lower,
-  scope = list(
-    lower = lower,
-    upper = upper
-  ),
-  data = trainingData,
-  direction = "both",
-  k = 2
-)
-trainingResults <- trainingData
-```
-
-```{r}
-# Store the predicted and actual values for Model 3:
-trainingResults$model2Pred <- predict(model2, type = "response")
-trainingResults$model2Class <- ifelse(trainingResults$model2Pred > 0.5, 1, 0)
-trainingResults$Actual <- ifelse(trainingData$Status == 1, 1, 0)
-
-# Create confusion matrix using table()
-confusionMatrix <- table(Predicted = trainingResults$model2Class, Actual = trainingResults$Actual)
-
-# Print and format confusion matrix
-kable(confusionMatrix, booktabs = TRUE, align = "c")%>%kable_styling(latex_options = "HOLD_position")
-```
-
-```{r}
-accuracy <- (confusionMatrix[1, 1] + confusionMatrix[0, 0]) / sum(confusionMatrix)
-sensitivity <- confusionMatrix[1, 1] / (confusionMatrix[1, 1] + confusionMatrix[0, 1])
-specificity <- confusionMatrix[0, 0] / (confusionMatrix[0, 0] + confusionMatrix[1, 0])
-
-metrics_df <- data.frame(
-  Metric = c("Accuracy", "Sensitivity", "Specificity"),
-  Value = c(accuracy, sensitivity, specificity)
-)
-
-# Use kable to create the table
-kable(metrics_df, format = "html", caption = "Model Performance Metrics")
-```

### 2023-12-15 Submission.Rmd
Added comparison to variables in unspervised

---  
title: "Submission"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    df_print: paged
---

```{r child = 'Final Project.Rmd'}
```

```{r child = 'EDA.Rmd'}
```

```{r child = 'Tree.Rmd'}
```

```{r child = 'Regression.Rmd'}
```

```{r child = 'Unsupervised.Rmd'}
```

### 2023-12-15 Submission.tex
Added comparison to variables in unspervised

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\usepackage{soul}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Submission},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Submission}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{background}{%
\section{Background}\label{background}}

\hypertarget{what-is-biliary-cholangitis}{%
\subsection{What is Biliary
Cholangitis?}\label{what-is-biliary-cholangitis}}

Primary Biliary Cholangitis (previously called Primary Biliary
Cirrhosis) is an autoimmune disease where bile ducts become swollen and
inflamed and block the flow of bile. Bile is a substance that aids with
digestion. The bile ducts carry bile from the liver to the small
intestine. The swelling and inflammation can lead to scarring of the
liver which is cirrhosis. Advanced cirrhosis can lead to liver failure
or liver cancer. Medication can slow progression. There is no definite
cure at this time.

\begin{figure}
\centering
\includegraphics{17715-primary-biliary-cholangitis.webp}
\caption{\ul{Primary Biliary Cholangitis
(\href{https://www.google.com/search?client=safari\&sca_esv=591276386\&rls=en\&sxsrf=AM9HkKmtJVOC2Jrf2BSYVB2O-QcnTSgx6g:1702664705100\&q=Primary+Biliary+Cholangitis\&tbm=isch\&source=lnms\&sa=X\&ved=2ahUKEwid0qyniJKDAxX9FFkFHVH3CcoQ0pQJegQIDxAB\&biw=1440\&bih=848\&dpr=2\#imgrc=UeV_fhP-8vPOlM}{Source})}}
\end{figure}

\hypertarget{objectives}{%
\subsection{Objectives}\label{objectives}}

\hypertarget{q1-what-predictors-if-any-increases-an-individuals-chance-of-receiving-a-liver-transplant}{%
\subsubsection{Q1: What predictors, if any, increases an individual's
chance of receiving a liver
transplant?}\label{q1-what-predictors-if-any-increases-an-individuals-chance-of-receiving-a-liver-transplant}}

\hypertarget{q2-are-there-factors-not-currently-included-in-medical-definition-that-can-help-distinguish-different-stages-of-biliary-cholangitis}{%
\subsubsection{Q2: Are there factors not currently included in medical
definition that can help distinguish different stages of Biliary
Cholangitis?}\label{q2-are-there-factors-not-currently-included-in-medical-definition-that-can-help-distinguish-different-stages-of-biliary-cholangitis}}

\hypertarget{clinical-study-dataset-background}{%
\subsection{Clinical Study + Dataset Background
:}\label{clinical-study-dataset-background}}

The dataset is produced from a clinical study by Mayo Clinic run from
1974-1984. The final `Status' of each patient was observed in 1986.
`Status' was either Dead, Censored, or Censored due to liver transplant.

A link to the original study is here:
\url{https://faculty.washington.edu/abansal/ShortCourse_DynamicDecisionMaking/Dickson1989_MayoPBCOriginalArticle.pdf}

\hypertarget{load-in-packages}{%
\subsubsection{Load in Packages}\label{load-in-packages}}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(survival)}
\FunctionTok{library}\NormalTok{(ggplot2)}
\FunctionTok{library}\NormalTok{(survival)}
\FunctionTok{library}\NormalTok{(survminer)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Loading required package: ggpubr
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'survminer'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:survival':
## 
##     myeloma
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(kableExtra)}
\FunctionTok{library}\NormalTok{(factoextra)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Welcome! Want to learn more? See two factoextra-related books at https://goo.gl/ve3WBa
\end{verbatim}

\hypertarget{data-processing}{%
\subsubsection{Data Processing}\label{data-processing}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#load in the data}
\FunctionTok{library}\NormalTok{(tidyverse)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## -- Attaching core tidyverse packages ------------------------ tidyverse 2.0.0 --
## v dplyr     1.1.3     v readr     2.1.4
## v forcats   1.0.0     v stringr   1.5.0
## v lubridate 1.9.2     v tibble    3.2.1
## v purrr     1.0.2     v tidyr     1.3.0
## -- Conflicts ------------------------------------------ tidyverse_conflicts() --
## x dplyr::filter()     masks stats::filter()
## x dplyr::group_rows() masks kableExtra::group_rows()
## x dplyr::lag()        masks stats::lag()
## i Use the conflicted package (<http://conflicted.r-lib.org/>) to force all conflicts to become errors
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#load data}
\NormalTok{cirrhosis }\OtherTok{\textless{}{-}} \FunctionTok{read\_csv}\NormalTok{(}\StringTok{"cirrhosis.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Rows: 418 Columns: 20
## -- Column specification --------------------------------------------------------
## Delimiter: ","
## chr  (7): Status, Drug, Sex, Ascites, Hepatomegaly, Spiders, Edema
## dbl (13): ID, N_Days, Age, Bilirubin, Cholesterol, Albumin, Copper, Alk_Phos...
## 
## i Use `spec()` to retrieve the full column specification for this data.
## i Specify the column types or set `show_col_types = FALSE` to quiet this message.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{head}\NormalTok{(cirrhosis)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 6 x 20
##      ID N_Days Status Drug          Age Sex   Ascites Hepatomegaly Spiders Edema
##   <dbl>  <dbl> <chr>  <chr>       <dbl> <chr> <chr>   <chr>        <chr>   <chr>
## 1     1    400 D      D-penicill~ 21464 F     Y       Y            Y       Y    
## 2     2   4500 C      D-penicill~ 20617 F     N       Y            Y       N    
## 3     3   1012 D      D-penicill~ 25594 M     N       N            N       S    
## 4     4   1925 D      D-penicill~ 19994 F     N       Y            Y       S    
## 5     5   1504 CL     Placebo     13918 F     N       Y            Y       N    
## 6     6   2503 D      Placebo     24201 F     N       Y            N       N    
## # i 10 more variables: Bilirubin <dbl>, Cholesterol <dbl>, Albumin <dbl>,
## #   Copper <dbl>, Alk_Phos <dbl>, SGOT <dbl>, Tryglicerides <dbl>,
## #   Platelets <dbl>, Prothrombin <dbl>, Stage <dbl>
\end{verbatim}

\hypertarget{pre-process-the-data}{%
\subsubsection{Pre-process the data}\label{pre-process-the-data}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#encode the sex, ascites, hepatomegaly, spiders, edema into binary variables}
\NormalTok{cirrhosis}\SpecialCharTok{$}\NormalTok{Sex }\OtherTok{\textless{}{-}} \FunctionTok{ifelse}\NormalTok{(cirrhosis}\SpecialCharTok{$}\NormalTok{Sex }\SpecialCharTok{==} \StringTok{"F"}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{)}
\NormalTok{cirrhosis}\SpecialCharTok{$}\NormalTok{Ascites }\OtherTok{\textless{}{-}} \FunctionTok{ifelse}\NormalTok{(cirrhosis}\SpecialCharTok{$}\NormalTok{Ascites }\SpecialCharTok{==} \StringTok{"Y"}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)}
\NormalTok{cirrhosis}\SpecialCharTok{$}\NormalTok{Hepatomegaly }\OtherTok{\textless{}{-}} \FunctionTok{ifelse}\NormalTok{(cirrhosis}\SpecialCharTok{$}\NormalTok{Hepatomegaly }\SpecialCharTok{==} \StringTok{"Y"}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)}
\NormalTok{cirrhosis}\SpecialCharTok{$}\NormalTok{Spiders }\OtherTok{\textless{}{-}} \FunctionTok{ifelse}\NormalTok{(cirrhosis}\SpecialCharTok{$}\NormalTok{Spiders }\SpecialCharTok{==} \StringTok{"Y"}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)}
\CommentTok{\#factorize the stage}
\NormalTok{cirrhosis}\SpecialCharTok{$}\NormalTok{Stage }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(cirrhosis}\SpecialCharTok{$}\NormalTok{Stage)}
\NormalTok{cirrhosis}\SpecialCharTok{$}\NormalTok{Status }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(cirrhosis}\SpecialCharTok{$}\NormalTok{Status)}
\NormalTok{cirrhosis}\SpecialCharTok{$}\NormalTok{Drug }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(cirrhosis}\SpecialCharTok{$}\NormalTok{Drug)}
\NormalTok{cirrhosis}\SpecialCharTok{$}\NormalTok{Sex }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(cirrhosis}\SpecialCharTok{$}\NormalTok{Sex)}
\CommentTok{\#cirrhosis$Ascites \textless{}{-} factor(cirrhosis$Ascites)}
\CommentTok{\#cirrhosis$Hepatomegaly \textless{}{-} factor(cirrhosis$Hepatomegaly)}
\CommentTok{\#cirrhosis$Spiders \textless{}{-} factor(cirrhosis$Spiders)}
\NormalTok{cirrhosis}\SpecialCharTok{$}\NormalTok{Edema }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(cirrhosis}\SpecialCharTok{$}\NormalTok{Edema)}
\NormalTok{cirrhosis}\SpecialCharTok{$}\NormalTok{Sex }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{ifelse}\NormalTok{(cirrhosis}\SpecialCharTok{$}\NormalTok{Sex }\SpecialCharTok{==} \DecValTok{0}\NormalTok{, }\StringTok{"Female"}\NormalTok{, }\StringTok{"Male"}\NormalTok{))}

\CommentTok{\#Drop the first column}
\NormalTok{cirrhosis }\OtherTok{\textless{}{-}}\NormalTok{ cirrhosis[,}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{itemize}
\item
  Engage in Exploratory Data Analysis
\item
  Use Machine Learning Algorithms:

  \begin{itemize}
  \tightlist
  \item
    You must use at least one of each: a Regression-based algorithm, a
    Tree-based algorithm, and an Unsupervised Learning algorithm.
  \item
    Each algorithm must be explained for a general audience.
  \item
    All results should be explained.
  \item
    Where appropriate, comparisons between methods should be discussed.
    This includes highlighting the strengths and weaknesses of the
    various approaches as well as when each is appropriate.
  \end{itemize}
\item
  A discussion of what you've learned from your work about the context
  and recommendations for future work.
\item
  Author Contributions--list out who did what in the project.
\item
  References--Citation style is your choice, but all sources should be
  documented (both in text and in the References section). This includes
  where you got your data.
\item
  A commented Code Appendix
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cirrhosis }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ N\_Days)) }\SpecialCharTok{+} \FunctionTok{geom\_histogram}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{facet\_wrap}\NormalTok{(}\AttributeTok{facets =} \FunctionTok{vars}\NormalTok{(Status))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.
\end{verbatim}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-9-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#Required libraries}
\FunctionTok{library}\NormalTok{(survival)}
\FunctionTok{library}\NormalTok{(ggplot2)}
\FunctionTok{library}\NormalTok{(survival)}
\FunctionTok{library}\NormalTok{(survminer)}
\end{Highlighting}
\end{Shaded}

\hypertarget{eda}{%
\section{EDA}\label{eda}}

\hypertarget{exploring-the-demographics-of-the-patients}{%
\subsection{Exploring the demographics of the
patients}\label{exploring-the-demographics-of-the-patients}}

\hypertarget{sex-ratio-of-patients}{%
\subsubsection{Sex Ratio of Patients}\label{sex-ratio-of-patients}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#cirrhosis$Sex \textless{}{-} factor(ifelse(cirrhosis$Sex == 0, "Female", "Male"))}

\CommentTok{\# Preparing the data for pie chart}
\NormalTok{sex\_data }\OtherTok{\textless{}{-}}\NormalTok{ cirrhosis }\SpecialCharTok{\%\textgreater{}\%} 
            \FunctionTok{count}\NormalTok{(Sex) }\SpecialCharTok{\%\textgreater{}\%} 
            \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Percentage =}\NormalTok{ n }\SpecialCharTok{/} \FunctionTok{sum}\NormalTok{(n) }\SpecialCharTok{*} \DecValTok{100}\NormalTok{)}

\CommentTok{\# Pie Chart}
\FunctionTok{ggplot}\NormalTok{(sex\_data, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\StringTok{""}\NormalTok{, }\AttributeTok{y=}\NormalTok{Percentage, }\AttributeTok{fill=}\NormalTok{Sex)) }\SpecialCharTok{+}
    \FunctionTok{geom\_bar}\NormalTok{(}\AttributeTok{stat=}\StringTok{"identity"}\NormalTok{, }\AttributeTok{width=}\DecValTok{1}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{coord\_polar}\NormalTok{(}\StringTok{"y"}\NormalTok{, }\AttributeTok{start=}\DecValTok{0}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{scale\_fill\_brewer}\NormalTok{(}\AttributeTok{palette=}\StringTok{"Pastel1"}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{theme\_minimal}\NormalTok{() }\SpecialCharTok{+}
    \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.line =} \FunctionTok{element\_blank}\NormalTok{(),}
          \AttributeTok{axis.text =} \FunctionTok{element\_blank}\NormalTok{(),}
          \AttributeTok{axis.ticks =} \FunctionTok{element\_blank}\NormalTok{(),}
          \AttributeTok{panel.grid =} \FunctionTok{element\_blank}\NormalTok{(),}
          \AttributeTok{plot.title =} \FunctionTok{element\_text}\NormalTok{(}\AttributeTok{size=}\DecValTok{14}\NormalTok{, }\AttributeTok{face=}\StringTok{"bold"}\NormalTok{)) }\SpecialCharTok{+}
    \FunctionTok{labs}\NormalTok{(}\AttributeTok{fill=}\StringTok{"Sex"}\NormalTok{, }\AttributeTok{title=}\StringTok{" Sex Amongst Patients"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-11-1.pdf}

Primary Biliary Cholangitis affects women much more than men at a 10-1
ratio. The data of this clinical study mimics the greater population.
Investigating gender-based differences in disease progression can
uncover any gender-specific patterns in PBC. This could lead to
gender-tailored treatment approaches and a better understanding of the
disease's biology, which might differ between males and females.

\hypertarget{age-distribution-of-patients}{%
\subsubsection{Age Distribution of
Patients}\label{age-distribution-of-patients}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Converting age from days to years for better readability}
\NormalTok{cirrhosis}\SpecialCharTok{$}\NormalTok{Age\_Years }\OtherTok{\textless{}{-}}\NormalTok{ cirrhosis}\SpecialCharTok{$}\NormalTok{Age }\SpecialCharTok{/} \FloatTok{365.25}

\CommentTok{\# Histogram}
\FunctionTok{ggplot}\NormalTok{(cirrhosis, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{Age\_Years)) }\SpecialCharTok{+}
    \FunctionTok{geom\_histogram}\NormalTok{(}\AttributeTok{binwidth=}\DecValTok{5}\NormalTok{, }\AttributeTok{fill=}\StringTok{"\#69b3a2"}\NormalTok{, }\AttributeTok{color=}\StringTok{"\#e9ecef"}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{theme\_minimal}\NormalTok{() }\SpecialCharTok{+}
    \FunctionTok{labs}\NormalTok{(}\AttributeTok{title=}\StringTok{"Age Distribution of PBC Patients"}\NormalTok{, }\AttributeTok{x=}\StringTok{"Age (years)"}\NormalTok{, }\AttributeTok{y=}\StringTok{"Count"}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{theme}\NormalTok{(}\AttributeTok{plot.title =} \FunctionTok{element\_text}\NormalTok{(}\AttributeTok{size=}\DecValTok{14}\NormalTok{, }\AttributeTok{face=}\StringTok{"bold"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-12-1.pdf}

The distribution of ages is quite normal.

\hypertarget{looking-at-the-data-itself}{%
\subsection{Looking at the data
itself}\label{looking-at-the-data-itself}}

\hypertarget{na-counts-amongst-our-variables}{%
\subsubsection{NA counts amongst our
variables}\label{na-counts-amongst-our-variables}}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{count }\OtherTok{\textless{}{-}}  \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{NA\_CountsPerVar =} \FunctionTok{colSums}\NormalTok{(}\FunctionTok{is.na}\NormalTok{(cirrhosis)))}
\FunctionTok{kable}\NormalTok{(count)}
\end{Highlighting}
\end{Shaded}

\begin{tabular}{l|r}
\hline
  & NA\_CountsPerVar\\
\hline
N\_Days & 0\\
\hline
Status & 0\\
\hline
Drug & 106\\
\hline
Age & 0\\
\hline
Sex & 0\\
\hline
Ascites & 106\\
\hline
Hepatomegaly & 106\\
\hline
Spiders & 106\\
\hline
Edema & 0\\
\hline
Bilirubin & 0\\
\hline
Cholesterol & 134\\
\hline
Albumin & 0\\
\hline
Copper & 108\\
\hline
Alk\_Phos & 106\\
\hline
SGOT & 106\\
\hline
Tryglicerides & 136\\
\hline
Platelets & 11\\
\hline
Prothrombin & 2\\
\hline
Stage & 6\\
\hline
Age\_Years & 0\\
\hline
\end{tabular}

An observation is that many of the variables have 106 NAs. This
indicates that a good fraction {[}106 patients{]} were equally tracked
and measured in a less extensive way.

\hypertarget{distribution-of-stages-of-biliary-cholangitis}{%
\subsection{Distribution of Stages of Biliary
Cholangitis}\label{distribution-of-stages-of-biliary-cholangitis}}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cirrhosis }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ Stage)) }\SpecialCharTok{+} \FunctionTok{geom\_bar}\NormalTok{(}\AttributeTok{fill =} \StringTok{"\#097969"}\NormalTok{) }\CommentTok{\# need color}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-14-1.pdf}

This distribution is left skewed and not symmetric. Most patients have
stage 3 and then 4 of Biliary Cholangitis.

\hypertarget{exploring-some-interactions-of-variables}{%
\subsection{Exploring some interactions of
variables}\label{exploring-some-interactions-of-variables}}

\hypertarget{how-does-prescense-of-a-d-penicillamine-with-stage-of-biliary-cholangitis-impact-number-of-days-till-death}{%
\subsubsection{How does prescense of a D-penicillamine with stage of
Biliary Cholangitis impact number of days till
death?}\label{how-does-prescense-of-a-d-penicillamine-with-stage-of-biliary-cholangitis-impact-number-of-days-till-death}}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{ggplot}\NormalTok{(}\AttributeTok{data =}\NormalTok{ cirrhosis, }\FunctionTok{aes}\NormalTok{ (Drug, Stage, }\AttributeTok{fill =}\NormalTok{ N\_Days))}\SpecialCharTok{+} \FunctionTok{geom\_tile}\NormalTok{() }\SpecialCharTok{+} \FunctionTok{scale\_fill\_distiller}\NormalTok{(}\AttributeTok{palette =} \StringTok{"RdPu"}\NormalTok{, }\AttributeTok{trans =} \StringTok{"reverse"}\NormalTok{) }
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-15-1.pdf}

This heat map shows that being a placebo in stage 1 gives you a greater
amount of days till death in this sample of patients.

\hypertarget{individuals-with-edema-and-if-they-were-on-the-drug}{%
\subsubsection{Individuals with Edema, and if they were on the
drug}\label{individuals-with-edema-and-if-they-were-on-the-drug}}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hi }\OtherTok{=} \FunctionTok{c}\NormalTok{(}\StringTok{"\#40B5AD"}\NormalTok{, }\StringTok{"\#009E60"}\NormalTok{, }\StringTok{"\#9FE2BF"}\NormalTok{)}
\FunctionTok{library}\NormalTok{(ggmosaic)}
\NormalTok{cirrhosis }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{ggplot}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{geom\_mosaic}\NormalTok{(}\FunctionTok{aes}\NormalTok{( }\AttributeTok{x =} \FunctionTok{product}\NormalTok{(Edema), }\AttributeTok{fill =}\NormalTok{ Drug)) }\SpecialCharTok{+} \FunctionTok{scale\_fill\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ hi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: `unite_()` was deprecated in tidyr 1.2.0.
## i Please use `unite()` instead.
## i The deprecated feature was likely used in the ggmosaic package.
##   Please report the issue at <https://github.com/haleyjeppson/ggmosaic>.
## This warning is displayed once every 8 hours.
## Call `lifecycle::last_lifecycle_warnings()` to see where this warning was
## generated.
\end{verbatim}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-16-1.pdf}

Edema is swelling due to too much liquid trapped in the body's tissues.
It's common complication in liver diseases and can significantly impact
patient quality of life and survival. Understanding how different levels
of edema (none, controlled by diuretics, or persistent despite
treatment) affect survival can inform patient management strategies and
highlight the need for aggressive interventions in certain cases. From
the graph we can see a even split between the placebo and drug for Edema
status. There is no NA values for Edema persistent despite diuretics. We
cannot directly compare it the splits for other statuses given their NA
values if known could change the look of the graph.

\hypertarget{mihir-choose-your-preffered-graph-and-explain}{%
\subsubsection{mihir choose your preffered graph and
explain}\label{mihir-choose-your-preffered-graph-and-explain}}

\hypertarget{survival-analysis-based-on-treatment-and-stage-how-does-the-survival-rate-differ-among-patients-at-different-stages-of-pbc-who-received-d-penicillamine-versus-placebo}{%
\subsubsection{Survival Analysis Based on Treatment and Stage: How does
the survival rate differ among patients at different stages of PBC who
received D-penicillamine versus
placebo?}\label{survival-analysis-based-on-treatment-and-stage-how-does-the-survival-rate-differ-among-patients-at-different-stages-of-pbc-who-received-d-penicillamine-versus-placebo}}

This question investigates the effectiveness of the drug D-penicillamine
compared to a placebo, considering the stage of PBC. By analyzing
survival rates across different disease stages and treatment types, we
can assess the drug's effectiveness at various disease stages. This
information is vital for clinicians to make informed decisions about
treatment plans and for researchers to understand the drug's impact.
(LOLL this is what the study acc sought to discover)

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{surv\_object }\OtherTok{\textless{}{-}} \FunctionTok{Surv}\NormalTok{(cirrhosis}\SpecialCharTok{$}\NormalTok{N\_Days, cirrhosis}\SpecialCharTok{$}\NormalTok{Status }\SpecialCharTok{==} \StringTok{\textquotesingle{}D\textquotesingle{}}\NormalTok{)}
\NormalTok{surv\_fit }\OtherTok{\textless{}{-}} \FunctionTok{survfit}\NormalTok{(surv\_object }\SpecialCharTok{\textasciitilde{}}\NormalTok{ cirrhosis}\SpecialCharTok{$}\NormalTok{Drug }\SpecialCharTok{+}\NormalTok{ cirrhosis}\SpecialCharTok{$}\NormalTok{Stage, }\AttributeTok{data =}\NormalTok{ cirrhosis)}
\FunctionTok{ggsurvplot}\NormalTok{(surv\_fit, }
           \AttributeTok{data =}\NormalTok{ cirrhosis,}
           \AttributeTok{conf.int =} \ConstantTok{TRUE}\NormalTok{, }
           \CommentTok{\#palette = c("\#00AFBB", "\#E7B800", "\#FC4E07"),}
           \AttributeTok{xlab =} \StringTok{"Days"}\NormalTok{, }
           \AttributeTok{ylab =} \StringTok{"Survival Probability"}\NormalTok{, }
           \AttributeTok{title =} \StringTok{"Survival Curves by Treatment and Stage"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-17-1.pdf}

\hypertarget{impact-of-liver-complications-on-survival-how-do-the-presence-of-ascites-hepatomegaly-and-spiders-relate-to-survival-time}{%
\subsubsection{Impact of Liver Complications on Survival: How do the
presence of ascites, hepatomegaly, and spiders relate to survival
time?}\label{impact-of-liver-complications-on-survival-how-do-the-presence-of-ascites-hepatomegaly-and-spiders-relate-to-survival-time}}

Exploring how liver-related symptoms (ascites, hepatomegaly, and
spiders) affect patient survival provides insights into the severity of
these complications in PBC progression. This can help in identifying
high-risk patients and understanding the disease's impact on liver
function.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{surv\_fit\_complications }\OtherTok{\textless{}{-}} \FunctionTok{survfit}\NormalTok{(surv\_object }\SpecialCharTok{\textasciitilde{}}\NormalTok{ cirrhosis}\SpecialCharTok{$}\NormalTok{Ascites }\SpecialCharTok{+}\NormalTok{ cirrhosis}\SpecialCharTok{$}\NormalTok{Hepatomegaly }\SpecialCharTok{+}\NormalTok{ cirrhosis}\SpecialCharTok{$}\NormalTok{Spiders, }\AttributeTok{data =}\NormalTok{ cirrhosis)}
\FunctionTok{ggsurvplot}\NormalTok{(surv\_fit\_complications, }
           \AttributeTok{data =}\NormalTok{ cirrhosis,}
           \AttributeTok{conf.int =} \ConstantTok{TRUE}\NormalTok{, }
           \CommentTok{\#palette = c("\#2E9FDF", "\#FC4E07", "\#6ACC65"),}
           \AttributeTok{xlab =} \StringTok{"Days"}\NormalTok{, }
           \AttributeTok{ylab =} \StringTok{"Survival Probability"}\NormalTok{, }
           \AttributeTok{title =} \StringTok{"Survival Curves by Liver Complications"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-18-1.pdf}

\hypertarget{correlation-between-biochemical-markers-and-disease-stage-what-is-the-correlation-between-biochemical-markers-bilirubin-cholesterol-albumin-copper-alk_phos-sgot-triglycerides-and-the-histologic-stage-of-the-disease}{%
\subsubsection{Correlation between Biochemical Markers and Disease
Stage: What is the correlation between biochemical markers (Bilirubin,
Cholesterol, Albumin, Copper, Alk\_Phos, SGOT, Triglycerides) and the
histologic stage of the
disease?}\label{correlation-between-biochemical-markers-and-disease-stage-what-is-the-correlation-between-biochemical-markers-bilirubin-cholesterol-albumin-copper-alk_phos-sgot-triglycerides-and-the-histologic-stage-of-the-disease}}

This analysis is crucial to understand how different biochemical markers
correlate with the disease's progression. Such correlations can aid in
the early detection of disease severity, help in monitoring the disease
progression, and potentially guide treatment adjustments.

\hypertarget{part-i-classification-algorithm-for-the-status-of-the-patient}{%
\section{Part I: Classification Algorithm for the Status of the
Patient}\label{part-i-classification-algorithm-for-the-status-of-the-patient}}

To address the first sub-goal of this project, we will be exploring the
prediction of the status of a patient at the end of the study. Our main
objective through this is to gain a stronger understanding of what
factors played in the role of the death of the patient. To do this, we
will be using Decision Trees. Let us start by exploring what these are
and why we chose to use them.

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Decision trees are a type of model used in statistics for making
predictions based on data. They work by breaking down a dataset into
smaller subsets through ``splits,'' resembling a tree with branches.
Each branch represents a possible decision or outcome, leading to a
final prediction or classification.

The main advantages of decision trees include their simplicity and
interpretability, as they are easy to visualize and understand. Another
important advantage is that they are able to learn with data with N/A
values, something alternatives such as Logistic Regression. Considering
the limitation we have with our data size and number of N/A values, this
an important advantage. However, there are some downsides such as a
tendency to overfit the data, which is something we need to be careful
about.

\begin{figure}
\centering
\includegraphics{DecTreeEg.png}
\caption{\ul{Decision Tree Example
(}\href{https://data-flair.training/blogs/r-decision-trees/}{Source})}
\end{figure}

\hypertarget{pre-processing-the-data}{%
\subsection{Pre-processing the data}\label{pre-processing-the-data}}

For this data, we will be dealing with the Status variable. The Status
variable was divided into 3 classes:

\begin{itemize}
\tightlist
\item
  Class 0 (D): The patient didn't survive by the end of the observation
\item
  Class 1 (C): The patient is censored, meaning that the observation
  period ended without the death being recorded
\item
  Class 2 (CL): Similar to Class 1, the patient is censored due to liver
  transplantation
\end{itemize}

Thus, we can group Class 1 and 2.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#Combine C and CL status into one variable and binarize}
\NormalTok{cirrhosisTreeData }\OtherTok{\textless{}{-}}\NormalTok{ cirrhosis}
\NormalTok{cirrhosisTreeData}\SpecialCharTok{$}\NormalTok{Status }\OtherTok{\textless{}{-}} \FunctionTok{ifelse}\NormalTok{(cirrhosisTreeData}\SpecialCharTok{$}\NormalTok{Status }\SpecialCharTok{==} \StringTok{"C"} \SpecialCharTok{|}\NormalTok{ cirrhosisTreeData}\SpecialCharTok{$}\NormalTok{Status }\SpecialCharTok{==} \StringTok{"CL"}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{methodology}{%
\subsection{Methodology}\label{methodology}}

Our objective is to create a classifier capable of predicting a
patient's outcome. To achieve this, we will be testing our data with the
cart algorithm. To ensure model validation, we'll be using a 80\%
training and 20\% testing data division. I will also be stratifying the
data based on the status.

To guarantee consistency and reproducibility in our results, we have
fixed the seed for our 80/20 data split at 380. With these steps, we are
now well-positioned to finalize our training and testing data sets.

The predictors in these models will be guided by the results from the
EDA. We will also use a variety of tools to understand the model's
performance.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Wrangle the Graduate data to set up training and testing datasets}
\NormalTok{modelCirrhosis }\OtherTok{\textless{}{-}}\NormalTok{ cirrhosisTreeData }\SpecialCharTok{\%\textgreater{}\%}
  \CommentTok{\#drop\_na() \%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}
    \AttributeTok{tempID =} \FunctionTok{row\_number}\NormalTok{(),}
    \AttributeTok{.before =}\NormalTok{ Status}
\NormalTok{  )}

\DocumentationTok{\#\# Set seed for reproducibility and slice {-}{-}{-}{-}}
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{380}\NormalTok{)}
\NormalTok{trainingData }\OtherTok{\textless{}{-}}\NormalTok{ modelCirrhosis }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(Status) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{slice\_sample}\NormalTok{(}\AttributeTok{prop =} \FloatTok{0.8}\NormalTok{)}

\NormalTok{testingData }\OtherTok{\textless{}{-}}\NormalTok{ modelCirrhosis }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{filter}\NormalTok{(}\SpecialCharTok{!}\NormalTok{(tempID }\SpecialCharTok{\%in\%}\NormalTok{ trainingData}\SpecialCharTok{$}\NormalTok{tempID))}
\end{Highlighting}
\end{Shaded}

\hypertarget{step-1-growing-the-tree}{%
\subsubsection{Step 1: Growing the tree}\label{step-1-growing-the-tree}}

There are useful packages to build decision trees in R: the \{tree\} and
the \{rpart\} (recursive partitioning) packages.

In this report, I have decided to use \{rpart\} because it provides more
flexibility for surrogate splits and the trees are a bit easier to make
attractive looking.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Grow Graduate tree via rpart package}
\FunctionTok{library}\NormalTok{(rpart)}
\NormalTok{rPartStatus }\OtherTok{\textless{}{-}} \FunctionTok{rpart}\NormalTok{(}
  \AttributeTok{formula =}\NormalTok{ Status }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Drug }\SpecialCharTok{+}\NormalTok{ Age }\SpecialCharTok{+}\NormalTok{ Sex }\SpecialCharTok{+}\NormalTok{ Ascites }\SpecialCharTok{+}\NormalTok{ Hepatomegaly }\SpecialCharTok{+}\NormalTok{ Spiders }\SpecialCharTok{+}\NormalTok{ Edema }\SpecialCharTok{+}\NormalTok{ Bilirubin }\SpecialCharTok{+}\NormalTok{ Cholesterol }\SpecialCharTok{+}\NormalTok{ Albumin }\SpecialCharTok{+}\NormalTok{ Copper }\SpecialCharTok{+}\NormalTok{ Alk\_Phos }\SpecialCharTok{+}\NormalTok{ SGOT }\SpecialCharTok{+}\NormalTok{ Platelets }\SpecialCharTok{+}\NormalTok{ Prothrombin }\SpecialCharTok{+}\NormalTok{ Stage }\SpecialCharTok{+}\NormalTok{ Tryglicerides, }
  \AttributeTok{data =}\NormalTok{ trainingData,}
  \AttributeTok{method =} \StringTok{"class"}\NormalTok{,}
  \AttributeTok{parms =} \FunctionTok{list}\NormalTok{(}\AttributeTok{split =} \StringTok{"information"}\NormalTok{)}
  \CommentTok{\# We did not need to use the control parameters}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{part-2-visualizing-the-tree}{%
\subsubsection{Part 2: Visualizing the
tree}\label{part-2-visualizing-the-tree}}

With the tree grown, we can now visualize it for an easy understanding
of its functioning. This is an important advantage for CART over
logistic regression.

The following is a basic diagram for the tree that was just grown.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Display rpart.plot {-}{-}{-}{-}}
 \FunctionTok{library}\NormalTok{(rpart.plot)}
\FunctionTok{rpart.plot}\NormalTok{(}
  \AttributeTok{x =}\NormalTok{ rPartStatus,}
  \AttributeTok{type =} \DecValTok{2}\NormalTok{,}
  \AttributeTok{extra =} \DecValTok{101}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-22-1.pdf}

(Explanation of above) To gain a further understanding of the data, we
can plot a tree yielding Collection Node style trees. This can help us
understand how the data is split.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Using the rattle package to visualize the tree {-}{-}{-}{-}}
\FunctionTok{library}\NormalTok{(rattle)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Loading required package: bitops
\end{verbatim}

\begin{verbatim}
## Rattle: A free graphical interface for data science with R.
## Version 5.5.1 Copyright (c) 2006-2021 Togaware Pty Ltd.
## Type 'rattle()' to shake, rattle, and roll your data.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{fancyRpartPlot}\NormalTok{(}
  \AttributeTok{model =}\NormalTok{ rPartStatus,}
  \AttributeTok{main =} \ConstantTok{NULL}\NormalTok{,}
  \AttributeTok{sub =} \ConstantTok{NULL}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-23-1.pdf}

\hypertarget{part-3-pruning-the-tree}{%
\subsubsection{Part 3: Pruning the tree}\label{part-3-pruning-the-tree}}

Pruning reduces the size of decision trees by removing parts of the tree
that do not provide power to classify instances. The first step of
pruning a tree is understanding the complexity parameter used. The
complexity parameter (cp) in rpart is the minimum improvement in the
model needed at each node. This is used when building the tree. We can
see the results based on the cross validation from the table below.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{invisible}\NormalTok{(}\FunctionTok{capture.output}\NormalTok{(\{cpTable }\OtherTok{\textless{}{-}} \FunctionTok{printcp}\NormalTok{(rPartStatus)\}))}

\FunctionTok{library}\NormalTok{(kableExtra)}

\FunctionTok{kable}\NormalTok{(}
  \AttributeTok{x =}\NormalTok{ cpTable,}
  \AttributeTok{col.names =} \FunctionTok{c}\NormalTok{(}\StringTok{"CP"}\NormalTok{, }\StringTok{"Num. of splits"}\NormalTok{, }\StringTok{"Rel. Error"}\NormalTok{,}
                \StringTok{"Mean Error"}\NormalTok{, }\StringTok{"Std. Deviation of Error"}\NormalTok{),}
  \AttributeTok{digits =} \DecValTok{3}\NormalTok{,}
  \AttributeTok{booktabs =} \ConstantTok{TRUE}\NormalTok{,}
  \AttributeTok{align =} \StringTok{"c"}\NormalTok{,}
  \AttributeTok{table.attr =} \StringTok{\textquotesingle{}data{-}quarto{-}disable{-}processing="true"\textquotesingle{}}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{tabular}{ccccc}
\toprule
CP & Num. of splits & Rel. Error & Mean Error & Std. Deviation of Error\\
\midrule
0.375 & 0 & 1.000 & 1.000 & 0.069\\
0.109 & 1 & 0.625 & 0.711 & 0.064\\
0.035 & 2 & 0.516 & 0.695 & 0.063\\
0.023 & 4 & 0.445 & 0.633 & 0.061\\
0.020 & 5 & 0.422 & 0.641 & 0.061\\
\addlinespace
0.010 & 7 & 0.383 & 0.648 & 0.062\\
\bottomrule
\end{tabular}

This can also be visualized in a graph to gain a better understanding of
the data. The graph below shows the connection between the cp, size of
tree and the x-val relative error.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plotcp}\NormalTok{(}
  \AttributeTok{x =}\NormalTok{ rPartStatus,}
  \AttributeTok{minline =} \ConstantTok{TRUE}\NormalTok{,}
  \AttributeTok{upper =} \StringTok{"size"}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-25-1.pdf}

From the graph we can see that a cp of 0.011 is ideal as it is under the
horizontal (dotted) reference line. We can prune the tree with this CP
value.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Prune the rpart Graduate Tree {-}{-}{-}{-}}
\NormalTok{rPartStatus2 }\OtherTok{\textless{}{-}} \FunctionTok{prune}\NormalTok{(}
  \AttributeTok{tree =}\NormalTok{ rPartStatus,}
  \AttributeTok{cp =} \FloatTok{0.029}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

We can plot the pruned tree

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{fancyRpartPlot}\NormalTok{(}
  \AttributeTok{model =}\NormalTok{ rPartStatus2,}
  \AttributeTok{main =} \ConstantTok{NULL}\NormalTok{,}
  \AttributeTok{sub =} \ConstantTok{NULL}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-27-1.pdf}

\hypertarget{part-4-results}{%
\subsubsection{Part 4: Results}\label{part-4-results}}

Now, we can evaluate the results of the tree on the testing data from
the initial 80-20 split. As is true whenever we use validation
approaches, we need to test out our model on the testing data set. This
will give us a more accurate understanding of how well the model fits
the context we're seeking to build our understanding of.

An important part of our results is understanding the role of prediction
and inference. In a broad sense, prediction refers to the process of
making forecasts about future events or unknown values based on a model
while inference generally refers to the process of drawing conclusions
from data. For the basic tree, I will be mainly focusing on prediction
aspects of the results. However, later in the report, we will also be
exploring inference findings.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{pred\_StatusRpart2 }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(}
  \AttributeTok{object =}\NormalTok{ rPartStatus2,}
  \AttributeTok{newdata =}\NormalTok{ testingData,}
  \AttributeTok{type =} \StringTok{"prob"}
\NormalTok{)}

\CommentTok{\# Data Wrangling the predictions {-}{-}{-}{-}}
\NormalTok{StatusPrediction }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
  \AttributeTok{rpart2\_non\_death =}\NormalTok{ pred\_StatusRpart2[, }\DecValTok{1}\NormalTok{],}
  \AttributeTok{rpart2\_death =}\NormalTok{ pred\_StatusRpart2[, }\DecValTok{2}\NormalTok{]}
\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}
    \AttributeTok{rpart2\_pred =} \FunctionTok{ifelse}\NormalTok{(}
      \AttributeTok{test =}\NormalTok{ rpart2\_death }\SpecialCharTok{\textgreater{}}\NormalTok{ rpart2\_non\_death,}
      \AttributeTok{yes =} \DecValTok{1}\NormalTok{,}
      \AttributeTok{no =} \DecValTok{0}
\NormalTok{    )}
\NormalTok{  )}

\DocumentationTok{\#\# Set predictions as factors}
\NormalTok{StatusPrediction}\SpecialCharTok{$}\NormalTok{rpart2\_pred }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(StatusPrediction}\SpecialCharTok{$}\NormalTok{rpart2\_pred)}

\CommentTok{\# Merge supervision column into predictions data frame {-}{-}{-}{-}}
\NormalTok{StatusPrediction }\OtherTok{\textless{}{-}} \FunctionTok{cbind}\NormalTok{(}
  \AttributeTok{tempID =}\NormalTok{ testingData}\SpecialCharTok{$}\NormalTok{tempID,}
  \AttributeTok{Status =}\NormalTok{ testingData}\SpecialCharTok{$}\NormalTok{Status,}
\NormalTok{  StatusPrediction}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

We can evaluate the results of this through a confusion matrix.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{StatusPrediction}\SpecialCharTok{$}\NormalTok{Status }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(StatusPrediction}\SpecialCharTok{$}\NormalTok{Status)}

\FunctionTok{library}\NormalTok{(yardstick)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
## Attaching package: 'yardstick'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:readr':
## 
##     spec
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Build confusion matrix for second tree model}
\NormalTok{conf\_matrix }\OtherTok{\textless{}{-}} \FunctionTok{conf\_mat}\NormalTok{(}
  \AttributeTok{data =}\NormalTok{ StatusPrediction,}
  \AttributeTok{truth =}\NormalTok{ Status,}
  \AttributeTok{estimate =}\NormalTok{ rpart2\_pred}
\NormalTok{)}\SpecialCharTok{$}\NormalTok{table}

\FunctionTok{kable}\NormalTok{(}
\NormalTok{  conf\_matrix,}
  \AttributeTok{col.names =} \FunctionTok{c}\NormalTok{(}\StringTok{"Prediction/Supervision"}\NormalTok{, }\StringTok{"0"}\NormalTok{, }\StringTok{"1"}\NormalTok{),}
  \AttributeTok{digits =} \DecValTok{3}\NormalTok{,}
  \AttributeTok{booktabs =} \ConstantTok{TRUE}\NormalTok{,}
  \AttributeTok{caption =} \StringTok{"Model 1: Confusion Matrix"}\NormalTok{,}
  \AttributeTok{align =} \StringTok{"c"}
\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
\FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options =} \StringTok{"HOLD\_position"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{table}[H]

\caption{\label{tab:unnamed-chunk-29}Model 1: Confusion Matrix}
\centering
\begin{tabular}[t]{lcc}
\toprule
Prediction/Supervision & 0 & 1\\
\midrule
0 & 19 & 9\\
1 & 14 & 43\\
\bottomrule
\end{tabular}
\end{table}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{accuracy }\OtherTok{\textless{}{-}} \FunctionTok{accuracy}\NormalTok{(StatusPrediction, Status, rpart2\_pred)}
\NormalTok{specificity }\OtherTok{\textless{}{-}} \FunctionTok{specificity}\NormalTok{(StatusPrediction, Status, rpart2\_pred)}
\NormalTok{sensitivity }\OtherTok{\textless{}{-}} \FunctionTok{sensitivity}\NormalTok{(StatusPrediction, Status, rpart2\_pred)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Build a data frame with model metrics {-}{-}{-}{-}}
\NormalTok{StatusPreds }\OtherTok{\textless{}{-}}\NormalTok{ StatusPrediction }\SpecialCharTok{\%\textgreater{}\%}
\NormalTok{  dplyr}\SpecialCharTok{::}\FunctionTok{select}\NormalTok{(tempID, Status, }\FunctionTok{contains}\NormalTok{(}\StringTok{"\_pred"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{pivot\_longer}\NormalTok{(}
    \AttributeTok{cols =} \SpecialCharTok{!}\FunctionTok{c}\NormalTok{(tempID, Status),}
    \AttributeTok{names\_to =} \StringTok{"model"}\NormalTok{,}
    \AttributeTok{values\_to =} \StringTok{"prediction"}
\NormalTok{  )}

\NormalTok{accuracy }\OtherTok{\textless{}{-}}\NormalTok{ StatusPreds }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(model) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{accuracy}\NormalTok{(}
    \AttributeTok{truth =}\NormalTok{ Status,}
    \AttributeTok{estimate =}\NormalTok{ prediction}
\NormalTok{  )}

\NormalTok{sensitivity }\OtherTok{\textless{}{-}}\NormalTok{ StatusPreds }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(model) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{sensitivity}\NormalTok{(}
    \AttributeTok{truth =}\NormalTok{ Status,}
    \AttributeTok{estimate =}\NormalTok{ prediction,}
    \AttributeTok{event\_level =} \StringTok{"second"}
\NormalTok{  )}

\NormalTok{specificity }\OtherTok{\textless{}{-}}\NormalTok{ StatusPreds }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(model) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{specificity}\NormalTok{(}
    \AttributeTok{truth =}\NormalTok{ Status,}
    \AttributeTok{estimate =}\NormalTok{ prediction,}
    \AttributeTok{event\_level =} \StringTok{"second"}
\NormalTok{  )}

\NormalTok{modelMetrics }\OtherTok{\textless{}{-}} \FunctionTok{bind\_rows}\NormalTok{(}
\NormalTok{  accuracy,}
\NormalTok{  sensitivity,}
\NormalTok{  specificity}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

From this, we can get the model's

\begin{itemize}
\item
  True Positive (TP): Predicted Graduated and actually Graduated. In
  this case, 2217
\item
  False Positive (FP): Predicted Graduated but actually Not Graduated.
  In this case, 1000
\item
  False Negative (FN): Predicted Not Graduated but actually Graduated.
  In this case, 244
\item
  True Negative (TN): Predicted Not Graduated and actually Not
  Graduated. In this case, 659
\end{itemize}

With this, we can also calculate the model's metrics.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Make a nice looking table of model metrics {-}{-}{-}{-}}
\NormalTok{modelMetrics }\SpecialCharTok{\%\textgreater{}\%}
\NormalTok{  dplyr}\SpecialCharTok{::}\FunctionTok{select}\NormalTok{(model, .metric, .estimate) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{pivot\_wider}\NormalTok{(}
    \AttributeTok{id\_cols =}\NormalTok{ model,}
    \AttributeTok{names\_from =}\NormalTok{ .metric,}
    \AttributeTok{values\_from =}\NormalTok{ .estimate}
\NormalTok{  ) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{kable}\NormalTok{(}
    \AttributeTok{digits =} \DecValTok{3}\NormalTok{,}
    \AttributeTok{booktabs =} \ConstantTok{TRUE}\NormalTok{,}
    \AttributeTok{align =} \StringTok{"c"}\NormalTok{,}
    \AttributeTok{table.attr =} \StringTok{\textquotesingle{}data{-}quarto{-}disable{-}processing="true"\textquotesingle{}}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

\begin{tabular}{cccc}
\toprule
model & accuracy & sensitivity & specificity\\
\midrule
rpart2\_pred & 0.729 & 0.827 & 0.576\\
\bottomrule
\end{tabular}

\hypertarget{part-ii-exploring-the-predictors-for-the-patients-status-using-logistic-regression}{%
\section{Part II: Exploring the predictors for the patient's status
using logistic
regression}\label{part-ii-exploring-the-predictors-for-the-patients-status-using-logistic-regression}}

Next, let us explore our question with a different type of model and
compare the results with our tree. To do this, we will be using (binary)
logistic regression model.

\hypertarget{introduction-1}{%
\subsection{Introduction}\label{introduction-1}}

Logistic regression is a statistical method used for binary
classification, which predicts the probability of an outcome that can be
either true or false. This is done by understanding the relationship
between a dependent binary variable and one or more independent
variables. Logistic regression is easy to implement and interpret.
However there are also some drawbacks, the model assumes a linear
relationship between the independent variables and the log odds of the
dependent variable, which may not always hold true in complex real-world
scenarios. Furthermore, unlike the decision tree, linear regression
models cannot ignore N/A values.

\begin{figure}
\centering
\includegraphics{46-4.png}
\caption{\ul{Logistic Regression Model
(\href{https://www.spiceworks.com/tech/artificial-intelligence/articles/what-is-logistic-regression/}{Source})}}
\end{figure}

\hypertarget{methodology-1}{%
\subsection{Methodology}\label{methodology-1}}

Similar to the tree, we will start by splitting the data set into
training and testing sets. Note that the data now only contains
instances where the patient was deceased. The training set will be used
to train our model, while the testing set will help evaluate its
performance. We'll use 80\% of the data for training and the remaining
20\% for testing. To allow reproducible code, we have fixed the seed at
380.

With this, we will build two candidate models:

\begin{itemize}
\item
  The first model will test the classification based on just the SGOT
\item
  The second model will use a step wise function using various
  predictors to see the best performance.
\end{itemize}

Another important consideration is the application of prediction
(estimating an outcome based on input variables) and inference
(understanding the relationships between variables). We will evaluate
the inference through the coefficient analysis and prediction through
roc curves and confusion matrix. It is important to note that these
metrics will complement each other in our understanding of the data.
However, the main focus of this analysis will be prediction and we will
work with several metrics to evaluate it.

We will use a variety of tools to understand the model's performance.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cirrhosisRegression }\OtherTok{\textless{}{-}}\NormalTok{ cirrhosis}
\NormalTok{cirrhosisRegression}\SpecialCharTok{$}\NormalTok{Status }\OtherTok{\textless{}{-}} \FunctionTok{ifelse}\NormalTok{(cirrhosisRegression}\SpecialCharTok{$}\NormalTok{Status }\SpecialCharTok{==} \StringTok{"C"} \SpecialCharTok{|}\NormalTok{ cirrhosisRegression}\SpecialCharTok{$}\NormalTok{Status }\SpecialCharTok{==} \StringTok{"CL"}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#model data}
\NormalTok{LRmodelData }\OtherTok{\textless{}{-}}\NormalTok{ cirrhosisRegression }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{drop\_na}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}
    \AttributeTok{tempID =} \FunctionTok{row\_number}\NormalTok{(),}
    \AttributeTok{.before =}\NormalTok{ Status}
\NormalTok{  )}

\CommentTok{\# Set seed for reproducibility and slice}
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{380}\NormalTok{)}
\NormalTok{trainingData }\OtherTok{\textless{}{-}}\NormalTok{ LRmodelData }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(Status) }\SpecialCharTok{\%\textgreater{}\%}  \CommentTok{\# group\_by() function ensures that the data}
  \FunctionTok{slice\_sample}\NormalTok{(}\AttributeTok{prop =} \FloatTok{0.80}\NormalTok{)}

\NormalTok{testingData }\OtherTok{\textless{}{-}}\NormalTok{ LRmodelData }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{filter}\NormalTok{(}\SpecialCharTok{!}\NormalTok{(tempID }\SpecialCharTok{\%in\%}\NormalTok{ trainingData}\SpecialCharTok{$}\NormalTok{tempID))}

\NormalTok{trainingResults }\OtherTok{\textless{}{-}}\NormalTok{ trainingData}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Form Candidate Model 1}
\NormalTok{model1 }\OtherTok{\textless{}{-}} \FunctionTok{glm}\NormalTok{(}
  \AttributeTok{formula =}\NormalTok{ Status }\SpecialCharTok{\textasciitilde{}}\NormalTok{ SGOT,}
  \AttributeTok{data =}\NormalTok{ trainingData,}
  \AttributeTok{family =}\NormalTok{ binomial}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Stepwise results:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Lower bound (Intercept only)}
\NormalTok{lower }\OtherTok{\textless{}{-}} \FunctionTok{glm}\NormalTok{(}
  \AttributeTok{formula =}\NormalTok{ Status }\SpecialCharTok{\textasciitilde{}} \DecValTok{1}\NormalTok{,}
  \AttributeTok{data =}\NormalTok{ trainingData,}
  \AttributeTok{family =}\NormalTok{ binomial}
\NormalTok{)}

\CommentTok{\# Upper bound }
\NormalTok{upper }\OtherTok{\textless{}{-}} \FunctionTok{glm}\NormalTok{(}
  \AttributeTok{formula =}\NormalTok{ Status }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Drug }\SpecialCharTok{+}\NormalTok{ Age }\SpecialCharTok{+}\NormalTok{ Sex }\SpecialCharTok{+}\NormalTok{ Ascites }\SpecialCharTok{+}\NormalTok{ Hepatomegaly }\SpecialCharTok{+}\NormalTok{ Spiders }\SpecialCharTok{+}\NormalTok{ Edema }\SpecialCharTok{+}\NormalTok{ Bilirubin }\SpecialCharTok{+}\NormalTok{ Cholesterol }\SpecialCharTok{+}\NormalTok{ Albumin }\SpecialCharTok{+}\NormalTok{ Copper }\SpecialCharTok{+}\NormalTok{ Alk\_Phos }\SpecialCharTok{+}\NormalTok{ SGOT }\SpecialCharTok{+}\NormalTok{ Platelets }\SpecialCharTok{+}\NormalTok{ Prothrombin }\SpecialCharTok{+}\NormalTok{ Stage,}
  \AttributeTok{data =}\NormalTok{ trainingData,}
  \AttributeTok{family =}\NormalTok{ binomial}
\NormalTok{)}

\CommentTok{\# Stepwise search}
\NormalTok{model2 }\OtherTok{\textless{}{-}} \FunctionTok{step}\NormalTok{(}
  \AttributeTok{object =}\NormalTok{ lower,}
  \AttributeTok{scope =} \FunctionTok{list}\NormalTok{(}
    \AttributeTok{lower =}\NormalTok{ lower,}
    \AttributeTok{upper =}\NormalTok{ upper}
\NormalTok{  ),}
  \AttributeTok{data =}\NormalTok{ trainingData,}
  \AttributeTok{direction =} \StringTok{"both"}\NormalTok{,}
  \AttributeTok{k =} \DecValTok{2}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Start:  AIC=298.13
## Status ~ 1
## 
##                Df Deviance    AIC
## + Bilirubin     1   247.26 251.26
## + Prothrombin   1   257.25 261.25
## + Copper        1   261.84 265.84
## + Edema         2   265.53 271.53
## + Ascites       1   273.57 277.57
## + Stage         3   270.16 278.16
## + Albumin       1   274.21 278.21
## + Alk_Phos      1   274.91 278.91
## + Hepatomegaly  1   275.05 279.05
## + Age           1   275.43 279.43
## + Spiders       1   278.02 282.02
## + SGOT          1   286.47 290.47
## + Platelets     1   291.25 295.25
## + Cholesterol   1   291.70 295.70
## + Sex           1   292.37 296.37
## + Drug          1   293.91 297.91
## <none>              296.12 298.12
## 
## Step:  AIC=251.26
## Status ~ Bilirubin
## 
##                Df Deviance    AIC
## + Age           1   225.30 231.30
## + Prothrombin   1   226.27 232.27
## + Alk_Phos      1   233.06 239.06
## + Edema         2   231.75 239.75
## + Albumin       1   238.01 244.01
## + Ascites       1   238.45 244.45
## + Stage         3   235.74 245.74
## + Copper        1   240.06 246.06
## + Spiders       1   241.63 247.63
## + Hepatomegaly  1   242.06 248.06
## + Platelets     1   243.87 249.87
## + Sex           1   244.21 250.21
## + Drug          1   244.37 250.37
## + Cholesterol   1   245.13 251.13
## <none>              247.26 251.26
## + SGOT          1   247.14 253.14
## - Bilirubin     1   296.12 298.12
## 
## Step:  AIC=231.3
## Status ~ Bilirubin + Age
## 
##                Df Deviance    AIC
## + Alk_Phos      1   209.76 217.76
## + Prothrombin   1   212.33 220.33
## + Edema         2   213.68 223.68
## + Copper        1   218.03 226.03
## + Spiders       1   218.93 226.93
## + Albumin       1   219.94 227.94
## + Ascites       1   220.76 228.76
## + Stage         3   217.58 229.58
## + Hepatomegaly  1   221.76 229.76
## + SGOT          1   222.78 230.78
## <none>              225.30 231.30
## + Platelets     1   223.57 231.57
## + Drug          1   223.83 231.83
## + Cholesterol   1   224.86 232.86
## + Sex           1   225.06 233.06
## - Age           1   247.26 251.26
## - Bilirubin     1   275.43 279.43
## 
## Step:  AIC=217.76
## Status ~ Bilirubin + Age + Alk_Phos
## 
##                Df Deviance    AIC
## + Prothrombin   1   197.16 207.16
## + Edema         2   198.44 210.44
## + Spiders       1   201.18 211.18
## + Stage         3   201.16 215.16
## + Ascites       1   205.18 215.18
## + Platelets     1   205.24 215.24
## + Copper        1   205.94 215.94
## + Hepatomegaly  1   206.62 216.62
## + Albumin       1   206.69 216.69
## <none>              209.76 217.76
## + SGOT          1   208.15 218.15
## + Cholesterol   1   208.82 218.82
## + Drug          1   208.84 218.84
## + Sex           1   209.68 219.68
## - Alk_Phos      1   225.30 231.30
## - Age           1   233.06 239.06
## - Bilirubin     1   251.55 257.55
## 
## Step:  AIC=207.16
## Status ~ Bilirubin + Age + Alk_Phos + Prothrombin
## 
##                Df Deviance    AIC
## + Spiders       1   191.39 203.39
## + Edema         2   189.41 203.41
## + Stage         3   188.90 204.90
## + Copper        1   193.94 205.94
## + Albumin       1   194.23 206.23
## + Hepatomegaly  1   194.37 206.37
## + SGOT          1   194.45 206.45
## + Drug          1   194.58 206.58
## + Platelets     1   194.95 206.95
## + Ascites       1   195.07 207.07
## <none>              197.16 207.16
## + Cholesterol   1   197.12 209.12
## + Sex           1   197.16 209.16
## - Prothrombin   1   209.76 217.76
## - Age           1   211.78 219.78
## - Alk_Phos      1   212.33 220.33
## - Bilirubin     1   225.31 233.31
## 
## Step:  AIC=203.39
## Status ~ Bilirubin + Age + Alk_Phos + Prothrombin + Spiders
## 
##                Df Deviance    AIC
## + Edema         2   185.35 201.35
## + SGOT          1   188.49 202.49
## + Drug          1   188.82 202.82
## + Copper        1   189.22 203.22
## + Albumin       1   189.33 203.33
## + Ascites       1   189.38 203.38
## <none>              191.39 203.39
## + Hepatomegaly  1   189.64 203.64
## + Stage         3   185.74 203.74
## + Platelets     1   189.96 203.96
## + Sex           1   191.13 205.13
## + Cholesterol   1   191.37 205.37
## - Spiders       1   197.16 207.16
## - Prothrombin   1   201.18 211.18
## - Age           1   207.34 217.34
## - Alk_Phos      1   208.03 218.03
## - Bilirubin     1   214.93 224.93
## 
## Step:  AIC=201.35
## Status ~ Bilirubin + Age + Alk_Phos + Prothrombin + Spiders + 
##     Edema
## 
##                Df Deviance    AIC
## + SGOT          1   181.77 199.77
## + Drug          1   182.57 200.57
## <none>              185.35 201.35
## + Copper        1   183.38 201.38
## + Albumin       1   183.48 201.48
## + Hepatomegaly  1   184.08 202.08
## + Stage         3   180.39 202.39
## + Platelets     1   184.40 202.40
## + Sex           1   185.10 203.10
## + Ascites       1   185.32 203.32
## + Cholesterol   1   185.35 203.35
## - Edema         2   191.39 203.39
## - Spiders       1   189.41 203.41
## - Prothrombin   1   192.56 206.56
## - Age           1   199.74 213.74
## - Alk_Phos      1   201.49 215.49
## - Bilirubin     1   206.71 220.71
## 
## Step:  AIC=199.77
## Status ~ Bilirubin + Age + Alk_Phos + Prothrombin + Spiders + 
##     Edema + SGOT
## 
##                Df Deviance    AIC
## + Drug          1   178.92 198.92
## <none>              181.77 199.77
## + Copper        1   179.90 199.90
## + Albumin       1   180.44 200.44
## + Hepatomegaly  1   180.66 200.66
## + Stage         3   177.01 201.01
## + Platelets     1   181.13 201.13
## - SGOT          1   185.35 201.35
## + Sex           1   181.53 201.53
## + Cholesterol   1   181.68 201.68
## + Ascites       1   181.70 201.70
## - Spiders       1   185.80 201.80
## - Edema         2   188.49 202.49
## - Prothrombin   1   189.92 205.92
## - Bilirubin     1   196.19 212.19
## - Alk_Phos      1   196.52 212.52
## - Age           1   198.70 214.70
## 
## Step:  AIC=198.92
## Status ~ Bilirubin + Age + Alk_Phos + Prothrombin + Spiders + 
##     Edema + SGOT + Drug
## 
##                Df Deviance    AIC
## + Copper        1   176.90 198.90
## <none>              178.92 198.92
## + Stage         3   173.10 199.10
## + Hepatomegaly  1   177.38 199.38
## + Albumin       1   177.49 199.49
## - Drug          1   181.77 199.77
## + Platelets     1   178.42 200.42
## - SGOT          1   182.57 200.57
## + Sex           1   178.69 200.69
## + Cholesterol   1   178.73 200.73
## - Spiders       1   182.83 200.83
## + Ascites       1   178.89 200.89
## - Edema         2   185.85 201.85
## - Prothrombin   1   188.64 206.64
## - Alk_Phos      1   193.00 211.00
## - Bilirubin     1   193.77 211.77
## - Age           1   194.21 212.21
## 
## Step:  AIC=198.9
## Status ~ Bilirubin + Age + Alk_Phos + Prothrombin + Spiders + 
##     Edema + SGOT + Drug + Copper
## 
##                Df Deviance    AIC
## <none>              176.90 198.90
## - Copper        1   178.92 198.92
## + Hepatomegaly  1   175.19 199.19
## + Albumin       1   175.31 199.31
## + Stage         3   171.76 199.76
## - Drug          1   179.90 199.90
## - Spiders       1   180.12 200.12
## + Platelets     1   176.38 200.38
## - SGOT          1   180.52 200.52
## + Ascites       1   176.82 200.82
## + Cholesterol   1   176.87 200.87
## + Sex           1   176.90 200.90
## - Edema         2   183.63 201.63
## - Bilirubin     1   184.57 204.57
## - Prothrombin   1   186.72 206.72
## - Alk_Phos      1   188.66 208.66
## - Age           1   191.54 211.54
\end{verbatim}

\hypertarget{results}{%
\section{Results}\label{results}}

Initially, we'll delve into the two preliminary models independently to
understand where they stand. Following that, we'll be deploying the best
candidate model on our test data. Regarding confusion matrices, we'll
employ a basic rule: if the predicted probability of a the patient's
status being ``deceased'' exceeds 0.5, we'll categorize them as deceased
(naïve rule).

\hypertarget{model-1}{%
\subsection{Model 1}\label{model-1}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Model 1 Coefficient Table}
\FunctionTok{as.data.frame}\NormalTok{(}\FunctionTok{summary}\NormalTok{(model1)}\SpecialCharTok{$}\NormalTok{coefficients) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{rownames\_to\_column}\NormalTok{(}\AttributeTok{var =} \StringTok{"X"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{rename}\NormalTok{(}\AttributeTok{coefficient =}\NormalTok{ Estimate) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate}\NormalTok{(}
    \AttributeTok{prob\_odds =} \FunctionTok{case\_when}\NormalTok{(}
\NormalTok{      coefficient }\SpecialCharTok{==} \StringTok{"(Intercept)"} \SpecialCharTok{\textasciitilde{}} \FunctionTok{exp}\NormalTok{(coefficient)}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{(coefficient)),}
      \AttributeTok{.default =} \FunctionTok{exp}\NormalTok{(coefficient)}
\NormalTok{    ),}
    \AttributeTok{.after =}\NormalTok{ coefficient}
\NormalTok{  ) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}
    \StringTok{\textasciigrave{}}\AttributeTok{Pr(\textgreater{}|z|)}\StringTok{\textasciigrave{}} \OtherTok{=} \FunctionTok{ifelse}\NormalTok{(}
      \AttributeTok{test =} \StringTok{\textasciigrave{}}\AttributeTok{Pr(\textgreater{}|z|)}\StringTok{\textasciigrave{}} \SpecialCharTok{\textless{}} \FloatTok{0.001}\NormalTok{,}
      \AttributeTok{yes =} \FunctionTok{paste}\NormalTok{(}\StringTok{"\textless{} 0.001"}\NormalTok{),}
      \AttributeTok{no =} \StringTok{\textasciigrave{}}\AttributeTok{Pr(\textgreater{}|z|)}\StringTok{\textasciigrave{}}
\NormalTok{    ),}
    \AttributeTok{X =} \FunctionTok{case\_when}\NormalTok{(}
\NormalTok{      X }\SpecialCharTok{==} \StringTok{"(Intercept)"} \SpecialCharTok{\textasciitilde{}} \StringTok{"Intercept"}\NormalTok{,}
      \FunctionTok{grepl}\NormalTok{(}\AttributeTok{x =}\NormalTok{ X, }\AttributeTok{pattern =} \StringTok{"SGOT"}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \StringTok{"SGOT"}
\NormalTok{    )}
\NormalTok{  ) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{kable}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{tabular}{l|r|r|r|r|l}
\hline
X & coefficient & prob\_odds & Std. Error & z value & Pr(>|z|)\\
\hline
Intercept & 1.3728804 & 3.9467022 & 0.3566274 & 3.849622 & < 0.001\\
\hline
SGOT & -0.0077204 & 0.9923094 & 0.0026073 & -2.961056 & 0.00306586722502071\\
\hline
\end{tabular}

This table shows us the results of our first model. We can see that,
holding other variables constant, a one-unit increase in SGOT is
associated with a decrease in the log-odds of the response variable by
0.0076446. Furthermore, the odds-ratio indicates that for each one-unit
increase in SGOT, the odds of the event occurring decrease by about
0.76\%.

We can also plot the confusion matrix for this model:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(janitor)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
## Attaching package: 'janitor'
\end{verbatim}

\begin{verbatim}
## The following objects are masked from 'package:stats':
## 
##     chisq.test, fisher.test
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Building confidence intervals for Model 1 coefficients}
\NormalTok{model1CI }\OtherTok{\textless{}{-}} \FunctionTok{confint}\NormalTok{(}
  \AttributeTok{object =}\NormalTok{ model1,}
  \AttributeTok{parm =} \StringTok{"SGOT"}\NormalTok{,}
  \AttributeTok{level =} \FloatTok{0.9}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Waiting for profiling to be done...
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{trainingResults }\OtherTok{\textless{}{-}}\NormalTok{ trainingData }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{ungroup}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{model1Pred =} \FunctionTok{predict}\NormalTok{(}\AttributeTok{object =}\NormalTok{ model1, }\AttributeTok{newdata =}\NormalTok{ ., }\AttributeTok{type =} \StringTok{"response"}\NormalTok{))}

\CommentTok{\# Apply naïve rule {-}{-}{-}{-}}
\NormalTok{trainingResults }\OtherTok{\textless{}{-}}\NormalTok{ trainingResults }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}
    \AttributeTok{model1Class =} \FunctionTok{case\_when}\NormalTok{(}
\NormalTok{      model1Pred }\SpecialCharTok{\textgreater{}} \FloatTok{0.5} \SpecialCharTok{\textasciitilde{}} \StringTok{"Censored"}\NormalTok{,}
      \AttributeTok{.default =} \StringTok{"Deceased"}
\NormalTok{    )}
\NormalTok{  )}

\CommentTok{\#Confusion Matrix for Model 1}
\NormalTok{trainingResults }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Patient\_status =} \FunctionTok{ifelse}\NormalTok{(Status }\SpecialCharTok{==} \DecValTok{1}\NormalTok{, }\StringTok{"Censored"}\NormalTok{, }\StringTok{"Deceased"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{tabyl}\NormalTok{(}\AttributeTok{var1 =}\NormalTok{ model1Class, }\AttributeTok{var2 =}\NormalTok{ Patient\_status) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{adorn\_title}\NormalTok{(}
    \AttributeTok{placement =} \StringTok{"combined"}\NormalTok{,}
    \AttributeTok{row\_name =} \StringTok{"Predicted"}\NormalTok{,}
    \AttributeTok{col\_name =} \StringTok{"Actual"}
\NormalTok{  ) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{kable}\NormalTok{(}
    \AttributeTok{booktabs =} \ConstantTok{TRUE}\NormalTok{,}
    \AttributeTok{align =} \StringTok{"c"}\NormalTok{,}
    \AttributeTok{caption =} \StringTok{"Model 1 Confusion Matrix"}
\NormalTok{  )}\SpecialCharTok{\%\textgreater{}\%}\FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options =} \StringTok{"HOLD\_position"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{table}[H]

\caption{\label{tab:unnamed-chunk-36}Model 1 Confusion Matrix}
\centering
\begin{tabular}[t]{ccc}
\toprule
Predicted/Actual & Censored & Deceased\\
\midrule
Censored & 117 & 68\\
Deceased & 15 & 20\\
\bottomrule
\end{tabular}
\end{table}

We can see that this model tends to over predict censored values. This
shows the need of bringing in more factors. Next let us look at our
Model 2, which has multiple factors as discussed earlier.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#Coeff for model 2}
\FunctionTok{as.data.frame}\NormalTok{(}\FunctionTok{summary}\NormalTok{(model2)}\SpecialCharTok{$}\NormalTok{coefficients) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{rename}\NormalTok{(}\AttributeTok{coefficient =}\NormalTok{ Estimate) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate}\NormalTok{(}
    \AttributeTok{prob\_odds =} \FunctionTok{case\_when}\NormalTok{(}
\NormalTok{      coefficient }\SpecialCharTok{==} \StringTok{"(Intercept)"} \SpecialCharTok{\textasciitilde{}} \FunctionTok{exp}\NormalTok{(coefficient)}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{(coefficient)),}
      \ConstantTok{TRUE} \SpecialCharTok{\textasciitilde{}} \FunctionTok{exp}\NormalTok{(coefficient)}
\NormalTok{    ),}
    \AttributeTok{.after =}\NormalTok{ coefficient}
\NormalTok{  ) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{kable}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{tabular}{l|r|r|r|r|r}
\hline
  & coefficient & prob\_odds & Std. Error & z value & Pr(>|z|)\\
\hline
(Intercept) & 13.1738171 & 5.264002e+05 & 2.5758130 & 5.1144306 & 0.0000003\\
\hline
Bilirubin & -0.2146715 & 8.068064e-01 & 0.0865790 & -2.4794883 & 0.0131571\\
\hline
Age & -0.0002039 & 9.997962e-01 & 0.0000571 & -3.5727423 & 0.0003533\\
\hline
Alk\_Phos & -0.0002988 & 9.997013e-01 & 0.0000968 & -3.0870015 & 0.0020219\\
\hline
Prothrombin & -0.6086022 & 5.441109e-01 & 0.2125486 & -2.8633552 & 0.0041918\\
\hline
Spiders & -0.7375877 & 4.782662e-01 & 0.4110588 & -1.7943606 & 0.0727556\\
\hline
EdemaS & -0.0940912 & 9.101997e-01 & 0.6115311 & -0.1538617 & 0.8777188\\
\hline
EdemaY & -16.3353205 & 1.000000e-07 & 857.1377366 & -0.0190580 & 0.9847948\\
\hline
SGOT & -0.0064630 & 9.935579e-01 & 0.0033599 & -1.9235635 & 0.0544093\\
\hline
DrugPlacebo & 0.6736154 & 1.961316e+00 & 0.3939801 & 1.7097702 & 0.0873084\\
\hline
Copper & -0.0038998 & 9.961078e-01 & 0.0027552 & -1.4154168 & 0.1569463\\
\hline
\end{tabular}

This analysis suggests significant associations between several
predictors (Bilirubin, Age, Alk\_Phos, Prothrombin, Spiders) and the
status of the patient. Each predictor, including the Intercept, shows a
statistically significant effect, as indicated by their small p-values
(Pr(\textgreater\textbar z\textbar)). The negative coefficients for
Bilirubin, Age, Alk\_Phos, Prothrombin, and Spiders imply that increases
in these variables are associated with a decrease in the log-odds of the
outcome. Bilirubin and Spiders are showing notably strong effects in the
model.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#do the Tukey{-}Anscombe plot}
\FunctionTok{ggplot}\NormalTok{(}
  \AttributeTok{data =} \FunctionTok{data.frame}\NormalTok{(}
    \AttributeTok{residuals =} \FunctionTok{residuals}\NormalTok{(model2, }\AttributeTok{type =} \StringTok{"pearson"}\NormalTok{),}
    \AttributeTok{fitted =} \FunctionTok{fitted}\NormalTok{(model2)}
\NormalTok{  ),}
  \AttributeTok{mapping =} \FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ fitted, }\AttributeTok{y =}\NormalTok{ residuals)}
\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{geom\_point}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{geom\_smooth}\NormalTok{(}
    \AttributeTok{formula =}\NormalTok{ y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x,}
    \AttributeTok{method =}\NormalTok{ stats}\SpecialCharTok{::}\NormalTok{loess,}
    \AttributeTok{method.args =} \FunctionTok{list}\NormalTok{(}\AttributeTok{degree =} \DecValTok{1}\NormalTok{),}
    \AttributeTok{se =} \ConstantTok{FALSE}\NormalTok{,}
    \AttributeTok{linewidth =} \FloatTok{0.5}
\NormalTok{  ) }\SpecialCharTok{+}
  \FunctionTok{theme\_bw}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}
    \AttributeTok{x =} \StringTok{"Fitted"}\NormalTok{,}
    \AttributeTok{y =} \StringTok{"Pearson Residuals"}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-38-1.pdf}

This figure shows us the Tukey-Anscombe plot using Pearson residuals for
Model 2. In an ideal fit, the residuals should be evenly distributed
about zero with constant mean and variance. The shape of the line
suggests that the model is not capturing someunder lying structure in
the data, especially in extreme cases.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#plot the gvif}
\FunctionTok{as.data.frame}\NormalTok{(car}\SpecialCharTok{::}\FunctionTok{vif}\NormalTok{(model2)) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{kable}\NormalTok{(}
    \AttributeTok{digits =} \DecValTok{3}\NormalTok{,}
    \AttributeTok{align =} \StringTok{"lcccc"}\NormalTok{,}
    \AttributeTok{booktab =} \ConstantTok{TRUE}\NormalTok{,}
    \AttributeTok{format.args =} \FunctionTok{list}\NormalTok{(}\AttributeTok{big.mark =} \StringTok{","}\NormalTok{),}
    \AttributeTok{table.attr =} \StringTok{\textquotesingle{}data{-}quarto{-}disable{-}processing="true"\textquotesingle{}}\NormalTok{,}
    \AttributeTok{label =} \StringTok{"GVIF analsyis"}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

\begin{tabular}{llcc}
\toprule
  & GVIF & Df & GVIF\textasciicircum{}(1/(2*Df))\\
\midrule
Bilirubin & 1.419 & 1 & 1.191\\
Age & 1.232 & 1 & 1.110\\
Alk\_Phos & 1.033 & 1 & 1.017\\
Prothrombin & 1.117 & 1 & 1.057\\
Spiders & 1.038 & 1 & 1.019\\
\addlinespace
Edema & 1.119 & 2 & 1.029\\
SGOT & 1.252 & 1 & 1.119\\
Drug & 1.091 & 1 & 1.044\\
Copper & 1.301 & 1 & 1.140\\
\bottomrule
\end{tabular}

The Variance Inflation Factor (VIF) values for the variables in the
model (Bilirubin, Age, Alk\_Phos, Prothrombin, Spiders) are all close to
1, indicating minimal multicollinearity. This means that these
predictors are relatively independent of each other, enhancing the
reliability of the model.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#Store the predicted and actual values for Model 2}
\NormalTok{trainingResults}\SpecialCharTok{$}\NormalTok{model2Pred }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(model2, }\AttributeTok{type =} \StringTok{"response"}\NormalTok{)}
\NormalTok{trainingResults}\SpecialCharTok{$}\NormalTok{model2Class }\OtherTok{\textless{}{-}} \FunctionTok{ifelse}\NormalTok{(trainingResults}\SpecialCharTok{$}\NormalTok{model2Pred }\SpecialCharTok{\textgreater{}} \FloatTok{0.5}\NormalTok{, }\StringTok{"Censored"}\NormalTok{, }\StringTok{"Deceased"}\NormalTok{)}
\NormalTok{trainingResults}\SpecialCharTok{$}\NormalTok{Actual }\OtherTok{\textless{}{-}} \FunctionTok{ifelse}\NormalTok{(trainingData}\SpecialCharTok{$}\NormalTok{Status }\SpecialCharTok{==} \DecValTok{1}\NormalTok{, }\StringTok{"Censored"}\NormalTok{, }\StringTok{"Deceased"}\NormalTok{)}

\CommentTok{\# Create confusion matrix using table}
\NormalTok{confusionMatrixRegression }\OtherTok{\textless{}{-}} \FunctionTok{table}\NormalTok{(}\AttributeTok{Predicted =}\NormalTok{ trainingResults}\SpecialCharTok{$}\NormalTok{model2Class, }\AttributeTok{Actual =}\NormalTok{ trainingResults}\SpecialCharTok{$}\NormalTok{Actual)}

\FunctionTok{kable}\NormalTok{(confusionMatrixRegression, }\AttributeTok{caption =} \StringTok{"Confusion matrix for Model 2"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{kable\_classic}\NormalTok{(}\AttributeTok{latex\_options =} \StringTok{"HOLD\_position"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{table}[H]

\caption{\label{tab:unnamed-chunk-40}Confusion matrix for Model 2}
\centering
\begin{tabular}[t]{l|r|r}
\hline
  & Censored & Deceased\\
\hline
Censored & 115 & 24\\
\hline
Deceased & 17 & 64\\
\hline
\end{tabular}
\end{table}

From this confusion matrix we can see the relationships between the True
Positive, True Negative, False Positive and False Negative values. From
this we can calculate:

\begin{itemize}
\item
  \textbf{Accuracy}: Approximately 81.36\%
\item
  \textbf{Recall}: Approximately 79.01\%
\item
  \textbf{Precision}: Approximately 72.73\%
\item
  \textbf{F1 Score}: Approximately 75.74\%
\end{itemize}

Lastly, let us look at the separation plots for each of the models.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(pROC)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Type 'citation("pROC")' for a citation.
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'pROC'
\end{verbatim}

\begin{verbatim}
## The following objects are masked from 'package:stats':
## 
##     cov, smooth, var
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(separationplot)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Loading required package: RColorBrewer
\end{verbatim}

\begin{verbatim}
## Loading required package: Hmisc
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'Hmisc'
\end{verbatim}

\begin{verbatim}
## The following objects are masked from 'package:dplyr':
## 
##     src, summarize
\end{verbatim}

\begin{verbatim}
## The following objects are masked from 'package:base':
## 
##     format.pval, units
\end{verbatim}

\begin{verbatim}
## Loading required package: MASS
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'MASS'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:dplyr':
## 
##     select
\end{verbatim}

\begin{verbatim}
## Loading required package: foreign
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Fit ROC Curves for later}
\DocumentationTok{\#\# Model 1}
\NormalTok{model1ROC }\OtherTok{\textless{}{-}} \FunctionTok{roc}\NormalTok{(}
  \AttributeTok{formula =}\NormalTok{ Status }\SpecialCharTok{\textasciitilde{}}\NormalTok{ model1Pred,}
  \AttributeTok{data =}\NormalTok{ trainingResults}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Setting levels: control = 0, case = 1
\end{verbatim}

\begin{verbatim}
## Setting direction: controls < cases
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{model1ROC\_df }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
  \AttributeTok{threshold =}\NormalTok{ model1ROC}\SpecialCharTok{$}\NormalTok{thresholds,}
  \AttributeTok{sensitivity =}\NormalTok{ model1ROC}\SpecialCharTok{$}\NormalTok{sensitivities,}
  \AttributeTok{specificity =}\NormalTok{ model1ROC}\SpecialCharTok{$}\NormalTok{specificities,}
  \AttributeTok{model =} \StringTok{"Model 1"}
\NormalTok{)}
\DocumentationTok{\#\# Model 2}
\NormalTok{model2ROC }\OtherTok{\textless{}{-}} \FunctionTok{roc}\NormalTok{(}
  \AttributeTok{formula =}\NormalTok{ Status }\SpecialCharTok{\textasciitilde{}}\NormalTok{ model2Pred,}
  \AttributeTok{data =}\NormalTok{ trainingResults}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Setting levels: control = 0, case = 1
## Setting direction: controls < cases
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{model2ROC\_df }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
  \AttributeTok{threshold =}\NormalTok{ model2ROC}\SpecialCharTok{$}\NormalTok{thresholds,}
  \AttributeTok{sensitivity =}\NormalTok{ model2ROC}\SpecialCharTok{$}\NormalTok{sensitivities,}
  \AttributeTok{specificity =}\NormalTok{ model2ROC}\SpecialCharTok{$}\NormalTok{specificities,}
  \AttributeTok{model =} \StringTok{"Model 2"}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Convert \textquotesingle{}Actual\textquotesingle{} column to numeric 0/1}
\NormalTok{trainingResults }\OtherTok{\textless{}{-}}\NormalTok{ trainingResults }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}
    \AttributeTok{actualNum =} \FunctionTok{if\_else}\NormalTok{(Actual }\SpecialCharTok{==} \StringTok{"Deceased"}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{)}
\NormalTok{  )}

\CommentTok{\#Sepeation Plot}
\FunctionTok{par}\NormalTok{(}\AttributeTok{mar =} \FunctionTok{c}\NormalTok{(}\DecValTok{4}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{))}
\FunctionTok{separationplot}\NormalTok{(}
  \AttributeTok{pred =}\NormalTok{ trainingResults}\SpecialCharTok{$}\NormalTok{model1Pred, }
  \AttributeTok{actual =}\NormalTok{ trainingResults}\SpecialCharTok{$}\NormalTok{actualNum, }
  \AttributeTok{type =} \StringTok{"rect"}\NormalTok{,}
  \AttributeTok{line =} \ConstantTok{TRUE}\NormalTok{, }
  \AttributeTok{lwd2 =} \DecValTok{2}\NormalTok{,}
  \AttributeTok{show.expected =} \ConstantTok{TRUE}\NormalTok{, }
  \AttributeTok{newplot =} \ConstantTok{FALSE}\NormalTok{,}
  \AttributeTok{heading =} \StringTok{"Model 1"}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-42-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#Sepeation Plot}
\FunctionTok{par}\NormalTok{(}\AttributeTok{mar =} \FunctionTok{c}\NormalTok{(}\DecValTok{4}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{))}
\FunctionTok{separationplot}\NormalTok{(}
  \AttributeTok{pred =}\NormalTok{ trainingResults}\SpecialCharTok{$}\NormalTok{model2Pred, }
  \AttributeTok{actual =}\NormalTok{ trainingResults}\SpecialCharTok{$}\NormalTok{actualNum, }
  \AttributeTok{type =} \StringTok{"rect"}\NormalTok{,}
  \AttributeTok{line =} \ConstantTok{TRUE}\NormalTok{, }
  \AttributeTok{lwd2 =} \DecValTok{2}\NormalTok{,}
  \AttributeTok{show.expected =} \ConstantTok{TRUE}\NormalTok{, }
  \AttributeTok{newplot =} \ConstantTok{FALSE}\NormalTok{,}
  \AttributeTok{heading =} \StringTok{"Model 2"}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-43-1.pdf}

The separation plot assesses the the fit of the model by providing the
model's ability to predict occurrences with a high probability and
non-occurrences with low probability. The separation plot above
separation plot suggests that Model 2 has a reasonably good performance
in predicting the patient's status, especially for the observations on
the left-most side of the plot compared to Model 1 with the training
data. We will later compare this graph with the testing data.

Lastly, let us look at the ROC curves for bothe the models.

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\# Merge into existing data frame}
\NormalTok{rocData }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(model1ROC\_df, model2ROC\_df)}

\DocumentationTok{\#\# AUC Data}
\NormalTok{aucData }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
  \AttributeTok{model =} \FunctionTok{c}\NormalTok{(}\StringTok{"Model 1"}\NormalTok{, }\StringTok{"Model 2"}\NormalTok{),}
  \AttributeTok{auc =} \FunctionTok{c}\NormalTok{(model1ROC}\SpecialCharTok{$}\NormalTok{auc, model2ROC}\SpecialCharTok{$}\NormalTok{auc)}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#ROC plot}
\FunctionTok{ggplot}\NormalTok{(}
  \AttributeTok{data =}\NormalTok{ rocData,}
  \AttributeTok{mapping =} \FunctionTok{aes}\NormalTok{(}\AttributeTok{x =} \DecValTok{1} \SpecialCharTok{{-}}\NormalTok{ specificity, }\AttributeTok{y =}\NormalTok{ sensitivity, }\AttributeTok{color =}\NormalTok{ model)}
\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{geom\_path}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{geom\_abline}\NormalTok{(}
    \AttributeTok{slope =} \DecValTok{1}\NormalTok{,}
    \AttributeTok{intercept =} \DecValTok{0}\NormalTok{,}
    \AttributeTok{linetype =} \StringTok{"dotted"}
\NormalTok{  ) }\SpecialCharTok{+}
  \FunctionTok{geom\_text}\NormalTok{(}
  \AttributeTok{inherit.aes =} \ConstantTok{FALSE}\NormalTok{,}
  \AttributeTok{data =}\NormalTok{ aucData,}
  \AttributeTok{mapping =} \FunctionTok{aes}\NormalTok{(}\AttributeTok{label =} \FunctionTok{paste}\NormalTok{(model, }\StringTok{"AUC: }\SpecialCharTok{\textbackslash{}n}\StringTok{"}\NormalTok{, }\FunctionTok{round}\NormalTok{(auc, }\DecValTok{3}\NormalTok{))),}
  \AttributeTok{x =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.25}\NormalTok{, }\FloatTok{0.15}\NormalTok{),}
  \AttributeTok{y =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.4}\NormalTok{, }\FloatTok{1.05}\NormalTok{)}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-45-1.pdf}

From the graphs we can interpret that:

Model 1: - Its ROC curve is above the line of no discrimination,
indicating that the model has some predictive capabilities. - The AUC is
0.64, which is better than random guessing but suggests there's room for
improvement.

Model 2: - The ROC curve for Model 2 is significantly above that of
Model 1, and much closer to the top-left corner, indicating better
predictive performance. - The AUC is 0.899, which suggests a good
classification performance, and it's notably better than Model 1. - Its
ability to discriminate between positive and negative classes is
superior to that of Model 1.

Lastly, let us plot our influence plot.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Influence Plot for Model 2}
\NormalTok{idCases }\OtherTok{\textless{}{-}}\NormalTok{ car}\SpecialCharTok{::}\FunctionTok{influencePlot}\NormalTok{(model2)}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-46-1.pdf}

The influence plot shows several data points with high leverage and
large residuals, indicating potential outliers. Some observations,
notably those labeled like ``149,'' have significant influence on the
regression model due to their Cook's D values.

\hypertarget{testing-our-model}{%
\section{Testing our model}\label{testing-our-model}}

Using our final model, we now turn to see how well this classifier does
on our testing data. Recall that we initally set the test data during
the train test split.

The confusion matrix below shows the performance of our model using the
naïve decision rule.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Set up testing data results}
\NormalTok{testingData }\OtherTok{\textless{}{-}}\NormalTok{ testingData }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}
    \AttributeTok{gradNum =} \FunctionTok{case\_when}\NormalTok{(}
\NormalTok{      Status }\SpecialCharTok{==} \DecValTok{0} \SpecialCharTok{\textasciitilde{}} \DecValTok{0}\NormalTok{,}
\NormalTok{      Status }\SpecialCharTok{==} \DecValTok{1} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}
\NormalTok{    ),}
    \AttributeTok{.after =}\NormalTok{ Status}
\NormalTok{  )}
\NormalTok{testingData}\SpecialCharTok{$}\NormalTok{predict }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(}
  \AttributeTok{object =}\NormalTok{ model2,}
  \AttributeTok{newdata =}\NormalTok{ testingData,}
  \AttributeTok{type =} \StringTok{"response"}
\NormalTok{)}
\NormalTok{testingData }\OtherTok{\textless{}{-}}\NormalTok{ testingData }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}
    \AttributeTok{model2Class =} \FunctionTok{case\_when}\NormalTok{(}
\NormalTok{      predict }\SpecialCharTok{\textgreater{}} \FloatTok{0.5} \SpecialCharTok{\textasciitilde{}} \StringTok{"Censored"}\NormalTok{,}
      \AttributeTok{.default =} \StringTok{"Deceased"}
\NormalTok{    )}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{testingData}\SpecialCharTok{$}\NormalTok{Status }\OtherTok{\textless{}{-}} \FunctionTok{ifelse}\NormalTok{(testingData}\SpecialCharTok{$}\NormalTok{Status }\SpecialCharTok{==} \DecValTok{1}\NormalTok{, }\StringTok{"Censored"}\NormalTok{, }\StringTok{"Deceased"}\NormalTok{)}

\CommentTok{\# Build Confusion Matrix for Testing Data}
\NormalTok{testingData }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{tabyl}\NormalTok{(}\AttributeTok{var1 =}\NormalTok{ model2Class, }\AttributeTok{var2 =}\NormalTok{ Status) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{adorn\_title}\NormalTok{(}
    \AttributeTok{placement =} \StringTok{"combined"}\NormalTok{,}
    \AttributeTok{row\_name =} \StringTok{"Predicted"}\NormalTok{,}
    \AttributeTok{col\_name =} \StringTok{"Actual"}
\NormalTok{  ) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{kable}\NormalTok{(}
    \AttributeTok{caption =} \StringTok{"Confusion Matrix for Test data"}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

\begin{table}

\caption{\label{tab:unnamed-chunk-48}Confusion Matrix for Test data}
\centering
\begin{tabular}[t]{l|r|r}
\hline
Predicted/Actual & Censored & Deceased\\
\hline
Censored & 30 & 13\\
\hline
Deceased & 3 & 10\\
\hline
\end{tabular}
\end{table}

\begin{itemize}
\item
  \textbf{Accuracy}: Approximately 71.43\%
\item
  \textbf{Recall}: Approximately 76.92\%
\item
  \textbf{Precision}: Approximately 43.48\%
\item
  \textbf{F1 Score}: Approximately 55.56\%
\end{itemize}

This shows that our model is has struggled with overfitting, with an
especially low precision score. We will discuss this in the comparison
with the tree model, but this is a limitation of our data size. Lastly,
we can plot the separation plot.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#Sepeation Plot}
\FunctionTok{par}\NormalTok{(}\AttributeTok{mar =} \FunctionTok{c}\NormalTok{(}\DecValTok{4}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{))}
\FunctionTok{separationplot}\NormalTok{(}
  \AttributeTok{pred =}\NormalTok{ testingData}\SpecialCharTok{$}\NormalTok{predict, }
  \AttributeTok{actual =}\NormalTok{ testingData}\SpecialCharTok{$}\NormalTok{gradNum, }
  \AttributeTok{type =} \StringTok{"rect"}\NormalTok{,}
  \AttributeTok{line =} \ConstantTok{TRUE}\NormalTok{, }
  \AttributeTok{lwd2 =} \DecValTok{2}\NormalTok{,}
  \AttributeTok{show.expected =} \ConstantTok{TRUE}\NormalTok{, }
  \AttributeTok{newplot =} \ConstantTok{FALSE}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-49-1.pdf}

\hypertarget{part-iii-exploring-the-sub-clusters-of-the-cirrhosis-data-using-unsupervised-learning}{%
\section{Part III: Exploring the sub-clusters of the cirrhosis data
using unsupervised
learning}\label{part-iii-exploring-the-sub-clusters-of-the-cirrhosis-data-using-unsupervised-learning}}

In this section, we sought to if there are sub-collections of patients.
We expect sub-collections to mimic medical definitions of the stages of
Biliary Cirrhosis but are curious to see if clustering can reveal some
newer observations.

\hypertarget{introduction-2}{%
\subsection{Introduction}\label{introduction-2}}

Biliary Cirrhosis patients are typically clustered in four main clusters
(\href{https://www.healthline.com/health/primary-biliary-cirrhosis\#stages}{Source}):

\begin{itemize}
\item
  Stage 1: There's inflammation and damage to the walls of medium-sized
  bile ducts.
\item
  Stage 2: There's blockage of the small bile ducts.
\item
  Stage 3: This stage marks the beginning of scarring.
\item
  Stage 4: Cirrhosis has developed. This permanent, severe, scarring and
  damage to the liver.
\end{itemize}

In this section we will be using clustering on our data. K-means
Clustering is an unsupervised learning technique where data points are
grouped based on their similarities. It's commonly used to identify
patterns and structures within datasets without prior knowledge of the
groups. The main advantage of clustering is its ability to discover
hidden patterns in data. However, a significant drawback is the
subjectivity in defining the `similarity' criteria, which can lead to
varying results and interpretations.

\href{https://medium.datadriveninvestor.com/k-means-clustering-4a700d4a4720}{\includegraphics{1*fz-rjYPPRlGEMdTI-RLbDg.png}}

An important motivation for this part of the study is explore how does
``Stage'' which is typically defined based on medical professionals'
observations compares to the data that we have in this data set.

\hypertarget{pre-processing-the-data-1}{%
\subsection{Pre-processing the data}\label{pre-processing-the-data-1}}

We had to pre-process the data for the best performance. First, we
removed the `Stage' column to avoid using outcome-related features in
the unsupervised learning. The rows with N/A values were also removed.
Lastly, we transformed various categorical variables into numeric
formats, which is necessary for clustering algorithms.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#do the ml, take about observational stages vs now quantify, look into the different variables and what they mean.}
\NormalTok{cirrhosisCluster }\OtherTok{\textless{}{-}}\NormalTok{ cirrhosis}

\NormalTok{cirrhosisCluster }\OtherTok{\textless{}{-}}\NormalTok{ cirrhosisCluster }\SpecialCharTok{\%\textgreater{}\%}                 
\NormalTok{                                  dplyr}\SpecialCharTok{::}\FunctionTok{select}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{Stage)     }

\NormalTok{cirrhosisCluster }\OtherTok{\textless{}{-}} \FunctionTok{na.omit}\NormalTok{(cirrhosisCluster) }\CommentTok{\#cannot have NA values in clustering}

\NormalTok{cirrhosisCluster}\SpecialCharTok{$}\NormalTok{Age\_Years }\OtherTok{\textless{}{-}} \FunctionTok{round}\NormalTok{(cirrhosisCluster}\SpecialCharTok{$}\NormalTok{Age\_Years) }\CommentTok{\#round age to whole numbers}

\CommentTok{\#reode sex... recode others later if need be \#female is 0}

\CommentTok{\#cirrhosisCluster  \textless{}{-} cirrhosisCluster \%\textgreater{}\%}
\CommentTok{\#  select( {-}c(Status, Drug, Edema)) \%\textgreater{}\%}

\CommentTok{\# mutate(Sex = ifelse(Sex == \textquotesingle{}Female\textquotesingle{}, 0, 1))}

\NormalTok{  cirrhosisCluster  }\OtherTok{\textless{}{-}}\NormalTok{ cirrhosisCluster }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Sex =} \FunctionTok{ifelse}\NormalTok{(Sex }\SpecialCharTok{==} \StringTok{\textquotesingle{}Female\textquotesingle{}}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
          \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Transplant =} \FunctionTok{ifelse}\NormalTok{(Status }\SpecialCharTok{==} \StringTok{"CL"}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
              \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Status =} \FunctionTok{ifelse}\NormalTok{(Status }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\StringTok{\textquotesingle{}C\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}CL\textquotesingle{}}\NormalTok{), }\DecValTok{0}\NormalTok{ , }\DecValTok{1}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
                    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Drug =} \FunctionTok{ifelse}\NormalTok{(Drug }\SpecialCharTok{==} \StringTok{"D{-}penicillamine"}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
                        \FunctionTok{mutate}\NormalTok{(}\AttributeTok{EdemaDiurectics =} \FunctionTok{ifelse}\NormalTok{(Edema }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\StringTok{\textquotesingle{}S\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}Y\textquotesingle{}}\NormalTok{), }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
                              \FunctionTok{mutate}\NormalTok{(}\AttributeTok{NoEdemaORD =} \FunctionTok{ifelse}\NormalTok{(Edema }\SpecialCharTok{==} \StringTok{\textquotesingle{}N\textquotesingle{}}\NormalTok{ , }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
                                        \FunctionTok{mutate}\NormalTok{(}\AttributeTok{EdemaANDD =} \FunctionTok{ifelse}\NormalTok{(Edema }\SpecialCharTok{==} \StringTok{"Y"}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
                                                \FunctionTok{mutate}\NormalTok{(}\AttributeTok{EdemaORD =} \FunctionTok{ifelse}\NormalTok{(Edema }\SpecialCharTok{==} \StringTok{"S"}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
\NormalTok{                                                                              dplyr}\SpecialCharTok{::}\FunctionTok{select}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{Edema)}

\CommentTok{\#Data is already factored}

\CommentTok{\#mutate(Sex = ifelse(Sex == \textquotesingle{}Female\textquotesingle{}, 0, 1)) \%\textgreater{}\%}

   \CommentTok{\#mutate(Status = ifelse(Status \%in\% c(\textquotesingle{}C\textquotesingle{}, \textquotesingle{}CL\textquotesingle{}), 0 , 1)) \%\textgreater{}\%}
          \CommentTok{\#    mutate(Drug = ifelse(Drug == "D{-}penicillamine", 0, 1))}

\CommentTok{\#???:}

\CommentTok{\#make column yes no edema \#then yes no under treatment}

\CommentTok{\#same for transplant stuff}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{distCirrhosis }\OtherTok{\textless{}{-}} \FunctionTok{dist}\NormalTok{(}
  \AttributeTok{x =}\NormalTok{ cirrhosisCluster,}
  \AttributeTok{method =} \StringTok{"euclidean"}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{methodology-2}{%
\subsection{Methodology}\label{methodology-2}}

To apply the k-means algorithm. A mixed hierarchical and
non-hierarchical was applied. This was done using the hkmeansm module,
with k = 4 as our initial value. Based on the data properties, the
euclidean (L2) distance works best.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hybridCirrhosis }\OtherTok{\textless{}{-}} \FunctionTok{hkmeans}\NormalTok{(}
  \AttributeTok{x =}\NormalTok{ cirrhosisCluster,}
  \AttributeTok{k =} \DecValTok{4}\NormalTok{,}
  \AttributeTok{hc.metric =} \StringTok{"euclidean"}\NormalTok{,}
  \AttributeTok{hc.method =} \StringTok{"ward.D"}\NormalTok{,}
  \AttributeTok{iter.max =} \DecValTok{10}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{results-1}{%
\subsection{Results}\label{results-1}}

We were able to visualize the results of the k-means algorithm. The
first step was making a color palette and plotting the dendrogram tree.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{StagesPalette }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"\#AA336A"}\NormalTok{, }\StringTok{"\#770737"}\NormalTok{, }\StringTok{"\#40B5AD"}\NormalTok{, }\StringTok{"\#009E60"}\NormalTok{, }\StringTok{"\#9FE2BF"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\# MAKE A NEW PALLETE TO VISUALIZE}
\CommentTok{\# Plot the initial dendrogram for hybrid approach {-}{-}{-}{-}}
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{380}\NormalTok{)}
\FunctionTok{hkmeans\_tree}\NormalTok{(}
  \AttributeTok{hkmeans =}\NormalTok{ hybridCirrhosis,}
  \AttributeTok{rect.col =}\NormalTok{ StagesPalette,}
  \AttributeTok{cex =} \FloatTok{0.4}\NormalTok{,}
  \AttributeTok{main =} \StringTok{"Initial Hierarchical Clusters"}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-54-1.pdf}

As, you can see, the model was able to create clear clusters for the
data. However, it is important to find the best value of k to get the
optimal clusters. To do this we can use a scree plot. Here,
we\textquotesingle re looking for the number of clusters that
corresponds to the "elbow".

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Create scree plot for choosing k {-}{-}{-}{-}}
\FunctionTok{library}\NormalTok{(factoextra)}
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{380}\NormalTok{)}
\FunctionTok{fviz\_nbclust}\NormalTok{(}
  \AttributeTok{x =}\NormalTok{ cirrhosisCluster,}
  \AttributeTok{diss =} \ConstantTok{NULL}\NormalTok{,}
  \AttributeTok{FUNcluster =}\NormalTok{ kmeans,}
  \AttributeTok{method =} \StringTok{"wss"}\NormalTok{,}
  \AttributeTok{k.max =} \DecValTok{10}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-55-1.pdf}

From this, we identified that 5 was the ideal value of k, where the
Total Within {[}Cluster{]} Sums of Squares begins leveling off. Let us
create a new k-means model with k = 5. We can plot this refined model,
with a format more easy to visualize.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hybridCirrhosis2 }\OtherTok{\textless{}{-}} \FunctionTok{hkmeans}\NormalTok{(}
  \AttributeTok{x =}\NormalTok{ cirrhosisCluster,}
  \AttributeTok{k =} \DecValTok{5}\NormalTok{,}
  \AttributeTok{hc.metric =} \StringTok{"euclidean"}\NormalTok{,}
  \AttributeTok{hc.method =} \StringTok{"ward.D"}\NormalTok{,}
  \AttributeTok{iter.max =} \DecValTok{10}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Plot the final dendrogram for hybrid approach {-}{-}{-}{-}}
\CommentTok{\# library(factoextra)}
\FunctionTok{fviz\_dend}\NormalTok{(}
  \AttributeTok{x =}\NormalTok{ hybridCirrhosis2,}
  \AttributeTok{cex =} \FloatTok{0.4}\NormalTok{,}
  \AttributeTok{palette =}\NormalTok{ StagesPalette,}
  \AttributeTok{rect =} \ConstantTok{FALSE}\NormalTok{,}
  \AttributeTok{horiz =} \ConstantTok{TRUE}\NormalTok{,}
  \AttributeTok{repel =} \ConstantTok{TRUE}\NormalTok{,}
  \AttributeTok{main =} \StringTok{"Final Dendrogram"}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: The `<scale>` argument of `guides()` cannot be `FALSE`. Use "none" instead as
## of ggplot2 3.3.4.
## i The deprecated feature was likely used in the factoextra package.
##   Please report the issue at <https://github.com/kassambara/factoextra/issues>.
## This warning is displayed once every 8 hours.
## Call `lifecycle::last_lifecycle_warnings()` to see where this warning was
## generated.
\end{verbatim}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-57-1.pdf}

As you can see, the 5 clusters were identified, which are color coded in
our updated diagram.

Lastly, let us plot these clusters. Here is a plot of the initial model
where k = 4

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Plot the final clustering for hybrid approach {-}{-}{-}{-}}
\CommentTok{\# library(factoextra)}
\FunctionTok{fviz\_cluster}\NormalTok{(}
  \AttributeTok{object =}\NormalTok{ hybridCirrhosis,}
  \AttributeTok{stand =} \ConstantTok{FALSE}\NormalTok{,}
  \AttributeTok{geom =} \StringTok{"point"}\NormalTok{,}
  \AttributeTok{main =} \StringTok{"Hybrid Cluster Plot {-} Initial model"}
\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ StagesPalette) }\SpecialCharTok{+}
  \FunctionTok{scale\_fill\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ StagesPalette) }\SpecialCharTok{+}
  \FunctionTok{theme\_bw}\NormalTok{() }
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-58-1.pdf}

We can also plot our refined model, as you can see the plot below
identifies 5 clear clusters.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Plot the final clustering for hybrid approach {-}{-}{-}{-}}
\CommentTok{\# library(factoextra)}
\FunctionTok{fviz\_cluster}\NormalTok{(}
  \AttributeTok{object =}\NormalTok{ hybridCirrhosis2,}
  \AttributeTok{stand =} \ConstantTok{FALSE}\NormalTok{,}
  \AttributeTok{geom =} \StringTok{"point"}\NormalTok{,}
  \AttributeTok{main =} \StringTok{"Hybrid Cluster Plot {-} Refined model"}
\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ StagesPalette) }\SpecialCharTok{+}
  \FunctionTok{scale\_fill\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ StagesPalette) }\SpecialCharTok{+}
  \FunctionTok{theme\_bw}\NormalTok{() }
\end{Highlighting}
\end{Shaded}

\includegraphics{Submission_files/figure-latex/unnamed-chunk-59-1.pdf}

\end{document}

### 2023-12-15 Unsupervised.Rmd
Added comparison to variables in unspervised

@@
 cirrhosisCluster <- cirrhosis
+cirrhosisCluster2 <- na.omit(cirrhosis)
 
 cirrhosisCluster <- cirrhosisCluster %>%                 
-                                  select(-Stage)
+                                  dplyr::select(-Stage)
 
@@
                                                 mutate(EdemaORD = ifelse(Edema == "S", 1, 0)) %>%
-                                                                              select(-Edema)
+                                                                              dplyr::select(-Edema)
 
@@
 ```
+
+Now we can go back to our initial goal: comparing how the custers compare to the Stage variable
+
+```{r}
+cirrhosisCluster2$cluster <- hybridCirrhosis$cluster
+# Calculate mean (or median, etc.) for each variable in each cluster
+library(dplyr)
+cluster_summary <- cirrhosisCluster2 %>%
+  group_by(cluster) %>%
+  summarise_all(funs(mean(., na.rm = TRUE))) # Replace mean with median or any other function as necessary
+kable(cluster_summary %>% dplyr::select(-Status, -Drug, - Sex, -Edema, -Stage), label = "Cluster Summary")
+ggplot(cirrhosisCluster2, aes(x = factor(cluster), fill = factor(Stage))) +
+  geom_bar(position = "dodge") +
+  labs(title = "Distribution of Stages within Clusters",
+       x = "Cluster",
+       y = "Count",
+       fill = "Stage") +
+  theme_minimal()
+
+```

### 2023-12-15 Final Project.Rmd
Wrote the introduction for the data

@@
 
+## Why is the data important to us?
+
+Both of us have been very interetsted in the intersection of medicine and machine learning.
+The idea that the concepts we are learning in our DS majors could have such an impact on the health industry was really special.
+Another important aspect was that there is so much potential for advancement of medicine, we both agreed that this area is MLs most promising aspect.
+The cirrhosis data set perticularly stood out to us because the data set included so many interesting attributes and we anted to see how they are connected with the outcomes of the patient.
+
+The data set also includes some important limitations.
+Most importantly, there are only 418 cases, greatly affecting the types of analysis we could do.
+This shows realm world problems data scientists face.
+Thus, our goal is to use our Statistical DS skillset to approach this problem, despite the limitations.
+
 ## Objectives
 
-### Q1: What predictors, if any, increases an individual's chance of receiving a liver transplant?
+Exploring the study, we concluded on two core questions we hope to address in this report:
 
-### Q2: Are there factors not currently included in medical definition that can help distinguish different stages of Biliary Cholangitis?
+### CRQ1: What predictors, if any affected the status of the patient at the end of the study?
+
+### CRQ2: Are there factors not currently included in medical definition that can help distinguish different stages of Biliary Cholangitis?
 
@@
 'Status' was either Dead, Censored, or Censored due to liver transplant.
-
 A link to the original study is here: <https://faculty.washington.edu/abansal/ShortCourse_DynamicDecisionMaking/Dickson1989_MayoPBCOriginalArticle.pdf>
 
-### Load in Packages
+The data contains the following attributes ([Data Source](https://www.kaggle.com/datasets/fedesoriano/cirrhosis-prediction-dataset/data)):
 
-```{r}
+1) **ID**: unique identifier\
+2) **N_Days**: number of days between registration and the earlier of death, transplantation, or study analysis time in July 1986\
+3) **Status**: status of the patient C (censored), CL (censored due to liver tx), or D (death)\
+4) **Drug**: type of drug D-penicillamine or placebo\
+5) **Age**: age in [days]\
+6) **Sex**: M (male) or F (female)\
+7) **Ascites**: presence of ascites N (No) or Y (Yes)\
+8) **Hepatomegaly**: presence of hepatomegaly N (No) or Y (Yes)\
+9) **Spiders**: presence of spiders N (No) or Y (Yes)\
+10) **Edema**: presence of edema N (no edema and no diuretic therapy for edema), S (edema present without diuretics, or edema resolved by diuretics), or Y (edema despite diuretic therapy)\
+11) **Bilirubin**: serum bilirubin in [mg/dl]\
+12) **Cholesterol**: serum cholesterol in [mg/dl]\
+13) **Albumin**: albumin in [gm/dl]\
+14) **Copper**: urine copper in [ug/day]\
+15) **Alk_Phos**: alkaline phosphatase in [U/liter]\
+16) **SGOT**: SGOT in [U/ml]\
+17) **Triglycerides**: triglicerides in [mg/dl]\
+18) **Platelets**: platelets per cubic [ml/1000]\
+19) **Prothrombin**: prothrombin time in seconds [s]\
+20) **Stage**: histologic stage of disease (1, 2, 3, or 4)
+
+For the basic pre-processing, we binary-encoded the categorical variables in the data set and set them as factors.
+
+```{r load packages}
 
@@
 
-### Data Processing
-
 ```{r}
@@
 library(tidyverse)
+library(kable)
 #load data
 cirrhosis <- read_csv("cirrhosis.csv")
-head(cirrhosis)
 ```
-
-### Pre-process the data
 

### 2023-12-15 EDA.Rmd
Added comparison to variables in unspervised
idk

@@
 
-# EDA
+# Exploratory Data Analysis
 
-## Exploring the demographics of the patients
+## Part 1: Exploring the demographics of the patients
 
@@
 
-## Looking at the data itself
+## Part 2: Looking at the data itself
 
@@
 
-An observation is that many of the variables have 106 NAs. This indicates that a good fraction [106 patients] were equally tracked and measured in a less extensive way.
+An observation is that many of the variables have 106 NAs. This indicates that a good fraction [106 patients] could have been equally tracked and measured in a less extensive way.
 
@@
 
-## Exploring some interactions of variables
+## Part 3: Exploring some interactions of variables
 
@@
 
-### mihir choose your preffered graph and explain
-
 ### Survival Analysis Based on Treatment and Stage: How does the survival rate differ among patients at different stages of PBC who received D-penicillamine versus placebo?
@@
 This information is vital for clinicians to make informed decisions about treatment plans and for researchers to understand the drug's impact.
-(LOLL this is what the study acc sought to discover)
 

### 2023-12-15 Final Project.Rmd
idk

@@
 
--   Engage in Exploratory Data Analysis
-
--   Use Machine Learning Algorithms:
-
-    -   You must use at least one of each: a Regression-based algorithm, a Tree-based algorithm, and an Unsupervised Learning algorithm.
-    -   Each algorithm must be explained for a general audience.
-    -   All results should be explained.
-    -   Where appropriate, comparisons between methods should be discussed. This includes highlighting the strengths and weaknesses of the various approaches as well as when each is appropriate.
-
--   A discussion of what you've learned from your work about the context and recommendations for future work.
-
--   Author Contributions--list out who did what in the project.
-
--   References--Citation style is your choice, but all sources should be documented (both in text and in the References section). This includes where you got your data.
-
--   A commented Code Appendix
-
 ```{r}

### 2023-12-15 Final Submission.Rmd
idk

---
title: "Final Project Report: An exploration of Biliary Cirrhosis and Treatments"
output: html_notebook
authors: "Mihir Kulkarni, Nithika Menon"
---

# Background

## What is Biliary Cholangitis?

Primary Biliary Cholangitis (previously called Primary Biliary Cirrhosis) is an autoimmune disease where bile ducts become swollen and inflamed and block the flow of bile. Bile is a substance that aids with digestion. The bile ducts carry bile from the liver to the small intestine. The swelling and inflammation can lead to scarring of the liver which is cirrhosis. Advanced cirrhosis can lead to liver failure or liver cancer. Medication can slow progression. There is no definite cure at this time.

![[Primary Biliary Cholangitis ([Source](https://www.google.com/search?client=safari&sca_esv=591276386&rls=en&sxsrf=AM9HkKmtJVOC2Jrf2BSYVB2O-QcnTSgx6g:1702664705100&q=Primary+Biliary+Cholangitis&tbm=isch&source=lnms&sa=X&ved=2ahUKEwid0qyniJKDAxX9FFkFHVH3CcoQ0pQJegQIDxAB&biw=1440&bih=848&dpr=2#imgrc=UeV_fhP-8vPOlM))]{.underline}](17715-primary-biliary-cholangitis.webp)

## Why is the data important to us?

Both of us have been very interetsted in the intersection of medicine and machine learning. The idea that the concepts we are learning in our DS majors could have such an impact on the health industry was really special. Another important aspect was that there is so much potential for advancement of medicine, we both agreed that this area is MLs most promising aspect. The cirrhosis data set perticularly stood out to us because the data set included so many interesting attributes and we anted to see how they are connected with the outcomes of the patient.

The data set also includes some important limitations. Most importantly, there are only 418 cases, greatly affecting the types of analysis we could do. This shows realm world problems data scientists face. Thus, our goal is to use our Statistical DS skillset to approach this problem, despite the limitations.

## Objectives

Exploring the study, we concluded on two core questions we hope to address in this report:

### CRQ1: What predictors, if any affected the status of the patient at the end of the study?

### CRQ2: Are there factors not currently included in medical definition that can help distinguish different stages of Biliary Cholangitis?

## Clinical Study + Dataset Background :

The dataset is produced from a clinical study by Mayo Clinic run from 1974-1984. The final 'Status' of each patient was observed in 1986. 'Status' was either Dead, Censored, or Censored due to liver transplant. A link to the original study is here: <https://faculty.washington.edu/abansal/ShortCourse_DynamicDecisionMaking/Dickson1989_MayoPBCOriginalArticle.pdf>

The data contains the following attributes ([Data Source](https://www.kaggle.com/datasets/fedesoriano/cirrhosis-prediction-dataset/data)):

1) **ID**: unique identifier\
2) **N_Days**: number of days between registration and the earlier of death, transplantation, or study analysis time in July 1986\
3) **Status**: status of the patient C (censored), CL (censored due to liver tx), or D (death)\
4) **Drug**: type of drug D-penicillamine or placebo\
5) **Age**: age in [days]\
6) **Sex**: M (male) or F (female)\
7) **Ascites**: presence of ascites N (No) or Y (Yes)\
8) **Hepatomegaly**: presence of hepatomegaly N (No) or Y (Yes)\
9) **Spiders**: presence of spiders N (No) or Y (Yes)\
10) **Edema**: presence of edema N (no edema and no diuretic therapy for edema), S (edema present without diuretics, or edema resolved by diuretics), or Y (edema despite diuretic therapy)\
11) **Bilirubin**: serum bilirubin in [mg/dl]\
12) **Cholesterol**: serum cholesterol in [mg/dl]\
13) **Albumin**: albumin in [gm/dl]\
14) **Copper**: urine copper in [ug/day]\
15) **Alk_Phos**: alkaline phosphatase in [U/liter]\
16) **SGOT**: SGOT in [U/ml]\
17) **Triglycerides**: triglicerides in [mg/dl]\
18) **Platelets**: platelets per cubic [ml/1000]\
19) **Prothrombin**: prothrombin time in seconds [s]\
20) **Stage**: histologic stage of disease (1, 2, 3, or 4)

For the basic pre-processing, we binary-encoded the categorical variables in the data set and set them as factors.

```{r load packages}

library(survival)
library(ggplot2)
library(survival)
library(survminer)
library(kableExtra)
library(factoextra)

```

```{r}
#load in the data
library(tidyverse)
#load data
cirrhosis <- read_csv("cirrhosis.csv")
```

```{r}
#encode the sex, ascites, hepatomegaly, spiders, edema into binary variables
cirrhosis$Sex <- ifelse(cirrhosis$Sex == "F", 0, 1)
cirrhosis$Ascites <- ifelse(cirrhosis$Ascites == "Y", 1, 0)
cirrhosis$Hepatomegaly <- ifelse(cirrhosis$Hepatomegaly == "Y", 1, 0)
cirrhosis$Spiders <- ifelse(cirrhosis$Spiders == "Y", 1, 0)
#factorize the stage
cirrhosis$Stage <- factor(cirrhosis$Stage)
cirrhosis$Status <- factor(cirrhosis$Status)
cirrhosis$Drug <- factor(cirrhosis$Drug)
cirrhosis$Sex <- factor(cirrhosis$Sex)
#cirrhosis$Ascites <- factor(cirrhosis$Ascites)
#cirrhosis$Hepatomegaly <- factor(cirrhosis$Hepatomegaly)
#cirrhosis$Spiders <- factor(cirrhosis$Spiders)
cirrhosis$Edema <- factor(cirrhosis$Edema)
cirrhosis$Sex <- factor(ifelse(cirrhosis$Sex == 0, "Female", "Male"))

#Drop the first column
cirrhosis <- cirrhosis[,-1]
```

```{r}
#?????
cirrhosis %>% 
  ggplot(aes(x = N_Days)) + geom_histogram() +
  facet_wrap(facets = vars(Status))

```

```{r}
#Required libraries
library(survival)
library(ggplot2)
library(survival)
library(survminer)

```

# Exploratory Data Analysis

## Part 1: Exploring the demographics of the patients

### Sex Ratio of Patients

```{r}

#cirrhosis$Sex <- factor(ifelse(cirrhosis$Sex == 0, "Female", "Male"))

# Preparing the data for pie chart
sex_data <- cirrhosis %>% 
            count(Sex) %>% 
            mutate(Percentage = n / sum(n) * 100)

# Pie Chart
ggplot(sex_data, aes(x="", y=Percentage, fill=Sex)) +
    geom_bar(stat="identity", width=1) +
    coord_polar("y", start=0) +
    scale_fill_brewer(palette="Pastel1") +
    theme_minimal() +
    theme(axis.line = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(),
          plot.title = element_text(size=14, face="bold")) +
    labs(fill="Sex", title=" Sex Amongst Patients")

```

Primary Biliary Cholangitis affects women much more than men at a 10-1 ratio. The data of this clinical study mimics the greater population. Investigating gender-based differences in disease progression can uncover any gender-specific patterns in PBC. This could lead to gender-tailored treatment approaches and a better understanding of the disease's biology, which might differ between males and females.

### Age Distribution of Patients

```{r}
# Converting age from days to years for better readability
cirrhosis$Age_Years <- cirrhosis$Age / 365.25

# Histogram
ggplot(cirrhosis, aes(x=Age_Years)) +
    geom_histogram(binwidth=5, fill="#69b3a2", color="#e9ecef") +
    theme_minimal() +
    labs(title="Age Distribution of PBC Patients", x="Age (years)", y="Count") +
    theme(plot.title = element_text(size=14, face="bold"))

```

The distribution of ages is quite normal.

## Part 2: Looking at the data itself

### NA counts amongst our variables

```{r}
count <-  data.frame(NA_CountsPerVar = colSums(is.na(cirrhosis)))
kable(count)
```

An observation is that many of the variables have 106 NAs. This indicates that a good fraction [106 patients] could have been equally tracked and measured in a less extensive way.

## Distribution of Stages of Biliary Cholangitis

```{r}

cirrhosis %>%
      ggplot(aes(x = Stage)) + geom_bar(fill = "#097969") # need color

```

This distribution is left skewed and not symmetric. Most patients have stage 3 and then 4 of Biliary Cholangitis.

## Part 3: Exploring some interactions of variables

### How does prescense of a D-penicillamine with stage of Biliary Cholangitis impact number of days till death?

```{r}

ggplot(data = cirrhosis, aes (Drug, Stage, fill = N_Days))+ geom_tile() + scale_fill_distiller(palette = "RdPu", trans = "reverse")

```

This heat map shows that being a placebo in stage 1 gives you a greater amount of days till death in this sample of patients.

### Individuals with Edema, and if they were on the drug

```{r}

hi = c("#40B5AD", "#009E60", "#9FE2BF")
library(ggmosaic)
cirrhosis %>%
  ggplot() +
  geom_mosaic(aes( x = product(Edema), fill = Drug)) + scale_fill_manual(values = hi)
```

Edema is swelling due to too much liquid trapped in the body's tissues. It's common complication in liver diseases and can significantly impact patient quality of life and survival. Understanding how different levels of edema (none, controlled by diuretics, or persistent despite treatment) affect survival can inform patient management strategies and highlight the need for aggressive interventions in certain cases. From the graph we can see a even split between the placebo and drug for Edema status. There is no NA values for Edema persistent despite diuretics. We cannot directly compare it the splits for other statuses given their NA values if known could change the look of the graph.

### Survival Analysis Based on Treatment and Stage: How does the survival rate differ among patients at different stages of PBC who received D-penicillamine versus placebo?

This question investigates the effectiveness of the drug D-penicillamine compared to a placebo, considering the stage of PBC. By analyzing survival rates across different disease stages and treatment types, we can assess the drug's effectiveness at various disease stages. This information is vital for clinicians to make informed decisions about treatment plans and for researchers to understand the drug's impact.

```{r}
surv_object <- Surv(cirrhosis$N_Days, cirrhosis$Status == 'D')
surv_fit <- survfit(surv_object ~ cirrhosis$Drug + cirrhosis$Stage, data = cirrhosis)
ggsurvplot(surv_fit, 
           data = cirrhosis,
           conf.int = TRUE, 
           #palette = c("#00AFBB", "#E7B800", "#FC4E07"),
           xlab = "Days", 
           ylab = "Survival Probability", 
           title = "Survival Curves by Treatment and Stage")
```

### Impact of Liver Complications on Survival: How do the presence of ascites, hepatomegaly, and spiders relate to survival time?

Exploring how liver-related symptoms (ascites, hepatomegaly, and spiders) affect patient survival provides insights into the severity of these complications in PBC progression. This can help in identifying high-risk patients and understanding the disease's impact on liver function.

```{r}
surv_fit_complications <- survfit(surv_object ~ cirrhosis$Ascites + cirrhosis$Hepatomegaly + cirrhosis$Spiders, data = cirrhosis)
ggsurvplot(surv_fit_complications, 
           data = cirrhosis,
           conf.int = TRUE, 
           #palette = c("#2E9FDF", "#FC4E07", "#6ACC65"),
           xlab = "Days", 
           ylab = "Survival Probability", 
           title = "Survival Curves by Liver Complications")
```

### Correlation between Biochemical Markers and Disease Stage: What is the correlation between biochemical markers (Bilirubin, Cholesterol, Albumin, Copper, Alk_Phos, SGOT, Triglycerides) and the histologic stage of the disease?

This analysis is crucial to understand how different biochemical markers correlate with the disease's progression. Such correlations can aid in the early detection of disease severity, help in monitoring the disease progression, and potentially guide treatment adjustments.

# Part I: Classification Algorithm for the Status of the Patient using CART

To address the first sub-goal of this project, we will be exploring the prediction of the status of a patient at the end of the study. Our main objective through this is to gain a stronger understanding of what factors played in the role of the death of the patient. To do this, we will be using Decision Trees. Let us start by exploring what these are and why we chose to use them.

## Introduction

Decision trees are a type of model used in statistics for making predictions based on data. They work by breaking down a dataset into smaller subsets through "splits," resembling a tree with branches. Each branch represents a possible decision or outcome, leading to a final prediction or classification.

The main advantages of decision trees include their simplicity and interpretability, as they are easy to visualize and understand. Another important advantage is that they are able to learn with data with N/A values, something alternatives such as Logistic Regression. Considering the limitation we have with our data size and number of N/A values, this an important advantage. However, there are some downsides such as a tendency to overfit the data, which is something we need to be careful about.

![[Decision Tree Example (]{.underline}[Source](https://data-flair.training/blogs/r-decision-trees/))](DecTreeEg.png)

## Pre-processing the data

For this data, we will be dealing with the Status variable. The Status variable was divided into 3 classes:

-   Class 0 (D): The patient didn't survive by the end of the observation
-   Class 1 (C): The patient is censored, meaning that the observation period ended without the death being recorded
-   Class 2 (CL): Similar to Class 1, the patient is censored due to liver transplantation

Thus, we can group Class 1 and 2.

```{r}
#Combine C and CL status into one variable and binarize
cirrhosisTreeData <- cirrhosis
cirrhosisTreeData$Status <- ifelse(cirrhosisTreeData$Status == "C" | cirrhosisTreeData$Status == "CL", 1, 0)
```

## Methodology

Our objective is to create a classifier capable of predicting a patient's outcome. To achieve this, we will be testing our data with the cart algorithm. To ensure model validation, we'll be using a 80% training and 20% testing data division. I will also be stratifying the data based on the status.

To guarantee consistency and reproducibility in our results, we have fixed the seed for our 80/20 data split at 380. With these steps, we are now well-positioned to finalize our training and testing data sets.

The predictors in these models will be guided by the results from the EDA. We will also use a variety of tools to understand the model's performance.

```{r}
# Wrangle the Graduate data to set up training and testing datasets
modelCirrhosis <- cirrhosisTreeData %>%
  #drop_na() %>%
  mutate(
    tempID = row_number(),
    .before = Status
  )

## Set seed for reproducibility and slice ----
set.seed(380)
trainingData <- modelCirrhosis %>%
  group_by(Status) %>%
  slice_sample(prop = 0.8)

testingData <- modelCirrhosis %>%
  filter(!(tempID %in% trainingData$tempID))
```

### Step 1: Growing the tree

There are useful packages to build decision trees in R: the {tree} and the {rpart} (recursive partitioning) packages.

In this report, we have decided to use {rpart} because it provides more flexibility for surrogate splits and the trees are a bit easier to make attractive looking.

```{r}
# Grow Graduate tree via rpart package
library(rpart)
rPartStatus <- rpart(
  formula = Status ~ Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage + Tryglicerides, 
  data = trainingData,
  method = "class",
  parms = list(split = "information")
  # We did not need to use the control parameters
)

```

### Part 2: Visualizing the tree

With the tree grown, we can now visualize it for an easy understanding of its functioning. This is an important advantage for CART over logistic regression.

The following is a basic diagram for the tree that was just grown.

```{r}
# Display rpart.plot ----
 library(rpart.plot)
rpart.plot(
  x = rPartStatus,
  type = 2,
  extra = 101
)
```

To gain a further understanding of the data, we can plot a tree yielding Collection Node style trees. This can help us understand how the data is split.

```{r}
# Using the rattle package to visualize the tree ----
library(rattle)

fancyRpartPlot(
  model = rPartStatus,
  main = NULL,
  sub = NULL
)
```

The tree shows us the splits that were done on Age, Bilirubin and Prothrombin. Interestingly, Stage did not contribute in the tree.

### Part 3: Pruning the tree

Pruning reduces the size of decision trees by removing parts of the tree that do not provide power to classify instances. The first step of pruning a tree is understanding the complexity parameter used. The complexity parameter (cp) in rpart is the minimum improvement in the model needed at each node. This is used when building the tree. We can see the results based on the cross validation from the table below.

```{r}
invisible(capture.output({cpTable <- printcp(rPartStatus)}))

library(kableExtra)

kable(
  x = cpTable,
  col.names = c("CP", "Num. of splits", "Rel. Error",
                "Mean Error", "Std. Deviation of Error"),
  digits = 3,
  booktabs = TRUE,
  align = "c",
  table.attr = 'data-quarto-disable-processing="true"'
)
```

This can also be visualized in a graph to gain a better understanding of the data. The graph below shows the connection between the cp, size of tree and the x-val relative error.

```{r}
plotcp(
  x = rPartStatus,
  minline = TRUE,
  upper = "size"
)
```

From the graph we can see that a cp of 0.039 is ideal as it is under the horizontal (dotted) reference line. We can prune the tree with this CP value.

```{r}
# Prune the rpart Graduate Tree ----
rPartStatus2 <- prune(
  tree = rPartStatus,
  cp = 0.029
)
```

We can plot the pruned tree

```{r}
fancyRpartPlot(
  model = rPartStatus2,
  main = NULL,
  sub = NULL
)
```

We ca see the pruned tree has cut out some leaf nodes. This would help in avoiding overfitting the model.

### Part 4: Results

Now, we can evaluate the results of the tree on the testing data from the initial 80-20 split. As is true whenever we use validation approaches, we need to test out our model on the testing data set. This will give us a more accurate understanding of how well the model fits the context we're seeking to build our understanding of.

An important part of our results is understanding the role of prediction and inference. In a broad sense, prediction refers to the process of making forecasts about future events or unknown values based on a model while inference generally refers to the process of drawing conclusions from data. For the basic tree, I will be mainly focusing on prediction aspects of the results. However, later in the report, we will also be exploring inference findings.

```{r}
pred_StatusRpart2 <- predict(
  object = rPartStatus2,
  newdata = testingData,
  type = "prob"
)

# Data Wrangling the predictions ----
StatusPrediction <- data.frame(
  rpart2_non_death = pred_StatusRpart2[, 1],
  rpart2_death = pred_StatusRpart2[, 2]
) %>%
  mutate(
    rpart2_pred = ifelse(
      test = rpart2_death > rpart2_non_death,
      yes = 1,
      no = 0
    )
  )

## Set predictions as factors
StatusPrediction$rpart2_pred <- as.factor(StatusPrediction$rpart2_pred)

# Merge supervision column into predictions data frame ----
StatusPrediction <- cbind(
  tempID = testingData$tempID,
  Status = testingData$Status,
  StatusPrediction
)
```

We can evaluate the results of this through a confusion matrix.

```{r}
StatusPrediction$Status <- factor(StatusPrediction$Status)

library(yardstick)

# Build confusion matrix for second tree model
conf_matrix <- conf_mat(
  data = StatusPrediction,
  truth = Status,
  estimate = rpart2_pred
)$table

kable(
  conf_matrix,
  col.names = c("Prediction/Supervision", "0", "1"),
  digits = 3,
  booktabs = TRUE,
  caption = "Model 1: Confusion Matrix (0=Deceased, 1=Censored)",
  align = "c"
) %>%
kable_styling(latex_options = "HOLD_position")

accuracy <- accuracy(StatusPrediction, Status, rpart2_pred)
specificity <- specificity(StatusPrediction, Status, rpart2_pred)
sensitivity <- sensitivity(StatusPrediction, Status, rpart2_pred)
```

```{r}
# Build a data frame with model metrics ----
StatusPreds <- StatusPrediction %>%
  dplyr::select(tempID, Status, contains("_pred")) %>%
  pivot_longer(
    cols = !c(tempID, Status),
    names_to = "model",
    values_to = "prediction"
  )

accuracy <- StatusPreds %>%
  group_by(model) %>%
  accuracy(
    truth = Status,
    estimate = prediction
  )

sensitivity <- StatusPreds %>%
  group_by(model) %>%
  sensitivity(
    truth = Status,
    estimate = prediction,
    event_level = "second"
  )

specificity <- StatusPreds %>%
  group_by(model) %>%
  specificity(
    truth = Status,
    estimate = prediction,
    event_level = "second"
  )

modelMetrics <- bind_rows(
  accuracy,
  sensitivity,
  specificity
)
```

With this, we can also calculate the model's metrics on the test data.

```{r}
# Make a nice looking table of model metrics ----
modelMetrics %>%
  dplyr::select(model, .metric, .estimate) %>%
  pivot_wider(
    id_cols = model,
    names_from = .metric,
    values_from = .estimate
  ) %>%
  kable(
    digits = 3,
    booktabs = TRUE,
    align = "c",
    table.attr = 'data-quarto-disable-processing="true"'
  )
```

As you can see the model shows good accuracy with 72.9%. However, the specificity is an issue with 57.6%. Now let us compare this with a different type of model: logistic regression.

---
title: "Regression-based"
author: "Mihir Kulkarni, Nithika Menon"
date: "2023-12-12"
output: html_document
---

# Part II: Exploring the predictors for the patient's status using logistic regression

Next, let us explore our question with a different type of model and compare the results with our tree. To do this, we will be using (binary) logistic regression model.

## Introduction

Logistic regression is a statistical method used for binary classification, which predicts the probability of an outcome that can be either true or false. This is done by understanding the relationship between a dependent binary variable and one or more independent variables. Logistic regression is easy to implement and interpret. However there are also some drawbacks, the model assumes a linear relationship between the independent variables and the log odds of the dependent variable, which may not always hold true in complex real-world scenarios. Furthermore, unlike the decision tree, linear regression models cannot ignore N/A values.

![[Logistic Regression Model ([Source](https://www.spiceworks.com/tech/artificial-intelligence/articles/what-is-logistic-regression/))]{.underline}](46-4.png)

## Methodology

Similar to the tree, we will start by splitting the data set into training and testing sets. Note that the data now only contains instances where the patient was deceased. The training set will be used to train our model, while the testing set will help evaluate its performance. We'll use 80% of the data for training and the remaining 20% for testing. To allow reproducible code, we have fixed the seed at 380.

With this, we will build two candidate models:

-   The first model will test the classification based on just the SGOT

-   The second model will use a step wise function using various predictors to see the best performance.

Another important consideration is the application of prediction (estimating an outcome based on input variables) and inference (understanding the relationships between variables). We will evaluate the inference through the coefficient analysis and prediction through roc curves and confusion matrix. It is important to note that these metrics will complement each other in our understanding of the data. However, the main focus of this analysis will be prediction and we will work with several metrics to evaluate it.

We will use a variety of tools to understand the model's performance.

```{r}
cirrhosisRegression <- cirrhosis
cirrhosisRegression$Status <- ifelse(cirrhosisRegression$Status == "C" | cirrhosisRegression$Status == "CL", 1, 0)
```

```{r}
#model data
LRmodelData <- cirrhosisRegression %>%
  drop_na() %>%
  mutate(
    tempID = row_number(),
    .before = Status
  )

# Set seed for reproducibility and slice
set.seed(380)
trainingData <- LRmodelData %>%
  group_by(Status) %>%  # group_by() function ensures that the data
  slice_sample(prop = 0.80)

testingData <- LRmodelData %>%
  filter(!(tempID %in% trainingData$tempID))

trainingResults <- trainingData
```

```{r}
# Form Candidate Model 1
model1 <- glm(
  formula = Status ~ SGOT,
  data = trainingData,
  family = binomial
)
```

Stepwise results:

```{r models}
# Lower bound (Intercept only)
lower <- glm(
  formula = Status ~ 1,
  data = trainingData,
  family = binomial
)

# Upper bound 
upper <- glm(
  formula = Status ~ Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage,
  data = trainingData,
  family = binomial
)

# Stepwise search
model2 <- step(
  object = lower,
  scope = list(
    lower = lower,
    upper = upper
  ),
  data = trainingData,
  direction = "both",
  k = 2
)
```

# Results

Initially, we'll delve into the two preliminary models independently to understand where they stand. Following that, we'll be deploying the best candidate model on our test data. Regarding confusion matrices, we'll employ a basic rule: if the predicted probability of a the patient's status being "deceased" exceeds 0.5, we'll categorize them as deceased (naïve rule).

## Model 1

```{r}
# Model 1 Coefficient Table
as.data.frame(summary(model1)$coefficients) %>%
  rownames_to_column(var = "X") %>%
  rename(coefficient = Estimate) %>% 
  mutate(
    prob_odds = case_when(
      coefficient == "(Intercept)" ~ exp(coefficient)/(1 + exp(coefficient)),
      .default = exp(coefficient)
    ),
    .after = coefficient
  ) %>%
  mutate(
    `Pr(>|z|)` = ifelse(
      test = `Pr(>|z|)` < 0.001,
      yes = paste("< 0.001"),
      no = `Pr(>|z|)`
    ),
    X = case_when(
      X == "(Intercept)" ~ "Intercept",
      grepl(x = X, pattern = "SGOT") ~ "SGOT"
    )
  ) %>%
  kable()

```

This table shows us the results of our first model. We can see that, holding other variables constant, a one-unit increase in SGOT is associated with a decrease in the log-odds of the response variable by 0.0076446. Furthermore, the odds-ratio indicates that for each one-unit increase in SGOT, the odds of the event occurring decrease by about 0.76%.

We can also plot the confusion matrix for this model:

```{r}
library(janitor)
# Building confidence intervals for Model 1 coefficients
model1CI <- confint(
  object = model1,
  parm = "SGOT",
  level = 0.9
)

trainingResults <- trainingData %>%
  ungroup() %>%
  mutate(model1Pred = predict(object = model1, newdata = ., type = "response"))

# Apply naïve rule ----
trainingResults <- trainingResults %>%
  mutate(
    model1Class = case_when(
      model1Pred > 0.5 ~ "Censored",
      .default = "Deceased"
    )
  )

#Confusion Matrix for Model 1
trainingResults %>%
  mutate(Patient_status = ifelse(Status == 1, "Censored", "Deceased")) %>%
  tabyl(var1 = model1Class, var2 = Patient_status) %>%
  adorn_title(
    placement = "combined",
    row_name = "Predicted",
    col_name = "Actual"
  ) %>%
  kable(
    booktabs = TRUE,
    align = "c",
    caption = "Model 1 Confusion Matrix"
  )%>%kable_styling(latex_options = "HOLD_position")

```

We can see that this model tends to over predict censored values. This shows the need of bringing in more factors. Next let us look at our Model 2, which has multiple factors as discussed earlier.

```{r}
#Coeff for model 2
as.data.frame(summary(model2)$coefficients) %>%
  rename(coefficient = Estimate) %>% 
  mutate(
    prob_odds = case_when(
      coefficient == "(Intercept)" ~ exp(coefficient)/(1 + exp(coefficient)),
      TRUE ~ exp(coefficient)
    ),
    .after = coefficient
  ) %>%
  kable()

```

This is the role of inference in evaluating our model. The most notable predictors are Bilirubin, Age, Alk_Phos, and Prothrombin, each showing a statistically significant relationship (p \< 0.05) with the dependent variable. The Intercept and EdemaY have extremely significant p-values, but the practical significance of EdemaY is questionable due to its large standard error. Other variables like Spiders, SGOT, DrugPlacebo, and Copper, while contributing to the model, do not reach conventional levels of statistical significance (p \< 0.05).

```{r}
#do the Tukey-Anscombe plot
ggplot(
  data = data.frame(
    residuals = residuals(model2, type = "pearson"),
    fitted = fitted(model2)
  ),
  mapping = aes(x = fitted, y = residuals)
) +
  geom_point() +
  geom_smooth(
    formula = y ~ x,
    method = stats::loess,
    method.args = list(degree = 1),
    se = FALSE,
    linewidth = 0.5
  ) +
  theme_bw() +
  labs(
    x = "Fitted",
    y = "Pearson Residuals"
  )
```

This figure shows us the Tukey-Anscombe plot using Pearson residuals for Model 2. In an ideal fit, the residuals should be evenly distributed about zero with constant mean and variance. The shape of the line suggests that the model is not capturing someunder lying structure in the datain extreme cases.

```{r}
#plot the gvif
as.data.frame(car::vif(model2)) %>%
  kable(
    digits = 3,
    align = "lcccc",
    booktab = TRUE,
    format.args = list(big.mark = ","),
    table.attr = 'data-quarto-disable-processing="true"',
    label = "GVIF analsyis"
  )

```

The Variance Inflation Factor (VIF) values for the variables in the model (Bilirubin, Age, Alk_Phos, Prothrombin, Spiders) are all close to 1, indicating minimal multicollinearity. This means that these predictors are relatively independent of each other, enhancing the reliability of the model.

```{r}
#Store the predicted and actual values for Model 2
trainingResults$model2Pred <- predict(model2, type = "response")
trainingResults$model2Class <- ifelse(trainingResults$model2Pred > 0.5, "Censored", "Deceased")
trainingResults$Actual <- ifelse(trainingData$Status == 1, "Censored", "Deceased")

# Create confusion matrix using table
confusionMatrixRegression <- table(Predicted = trainingResults$model2Class, Actual = trainingResults$Actual)

kable(confusionMatrixRegression, caption = "Confusion matrix for Model 2") %>%
  kable_classic(latex_options = "HOLD_position")

```

From this confusion matrix we can see the relationships between the True Positive, True Negative, False Positive and False Negative values. From this we can calculate:

-   **Accuracy**: Approximately 81.36%

-   **Recall**: Approximately 79.01%

-   **Precision**: Approximately 72.73%

-   **F1 Score**: Approximately 75.74%

Lastly, let us look at the separation plots for each of the models.

```{r}
library(pROC)
library(separationplot)
# Fit ROC Curves for later
## Model 1
model1ROC <- roc(
  formula = Status ~ model1Pred,
  data = trainingResults
)
model1ROC_df <- data.frame(
  threshold = model1ROC$thresholds,
  sensitivity = model1ROC$sensitivities,
  specificity = model1ROC$specificities,
  model = "Model 1"
)
## Model 2
model2ROC <- roc(
  formula = Status ~ model2Pred,
  data = trainingResults
)
model2ROC_df <- data.frame(
  threshold = model2ROC$thresholds,
  sensitivity = model2ROC$sensitivities,
  specificity = model2ROC$specificities,
  model = "Model 2"
)
```

```{r}
# Convert 'Actual' column to numeric 0/1
trainingResults <- trainingResults %>%
  mutate(
    actualNum = if_else(Actual == "Deceased", 0, 1)
  )

#Sepeation Plot
par(mar = c(4,0,0,0))
separationplot(
  pred = trainingResults$model1Pred, 
  actual = trainingResults$actualNum, 
  type = "rect",
  line = TRUE, 
  lwd2 = 2,
  show.expected = TRUE, 
  newplot = FALSE,
  heading = "Model 1"
)

```

```{r}
#Sepeation Plot
par(mar = c(4,0,0,0))
separationplot(
  pred = trainingResults$model2Pred, 
  actual = trainingResults$actualNum, 
  type = "rect",
  line = TRUE, 
  lwd2 = 2,
  show.expected = TRUE, 
  newplot = FALSE,
  heading = "Model 2"
)
```

The separation plot assesses the the fit of the model by providing the model's ability to predict occurrences with a high probability and non-occurrences with low probability. The separation plot above separation plot suggests that Model 2 has a reasonably good performance in predicting the patient's status, especially for the observations on the left-most side of the plot compared to Model 1 with the training data. We will later compare this graph with the testing data.

Lastly, let us look at the ROC curves for both the models.

```{r}
## Merge into existing data frame
rocData <- rbind(model1ROC_df, model2ROC_df)

## AUC Data
aucData <- data.frame(
  model = c("Model 1", "Model 2"),
  auc = c(model1ROC$auc, model2ROC$auc)
)
```

```{r}
#ROC plot
ggplot(
  data = rocData,
  mapping = aes(x = 1 - specificity, y = sensitivity, color = model)
) +
  geom_path() +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dotted"
  ) +
  geom_text(
  inherit.aes = FALSE,
  data = aucData,
  mapping = aes(label = paste(model, "AUC: \n", round(auc, 3))),
  x = c(0.25, 0.15),
  y = c(0.4, 1.05)
)
```

From the graphs we can interpret that:

Model 1:

-   Its ROC curve is above the line of no discrimination, indicating that the model has some predictive capabilities

-   The AUC is 0.638, which is better than random guessing but suggests there's room for improvement.

Model 2:

-   The ROC curve for Model 2 is significantly above that of Model 1, and much closer to the top-left corner, indicating better predictive performance.

-   The AUC is 0.899, which suggests a good classification performance, and it's notably better than Model 1.

-   Its ability to discriminate between positive and negative classes is superior to that of Model 1.

Lastly, let us plot our influence plot.

```{r}
# Influence Plot for Model 2
idCases <- car::influencePlot(model2)
```

The influence plot shows several data points with high leverage and large residuals, indicating potential outliers. Some observations, notably those labeled like "149," have significant influence on the regression model due to their Cook's D values.

# Testing our model

Using our final model, we now turn to see how well this classifier does on our testing data. Recall that we initially set the test data during the train test split.

The confusion matrix below shows the performance of our model using the naïve decision rule.

```{r}
# Set up testing data results
testingData <- testingData %>%
  mutate(
    gradNum = case_when(
      Status == 0 ~ 0,
      Status == 1 ~ 1
    ),
    .after = Status
  )
testingData$predict <- predict(
  object = model2,
  newdata = testingData,
  type = "response"
)
testingData <- testingData %>%
  mutate(
    model2Class = case_when(
      predict > 0.5 ~ "Censored",
      .default = "Deceased"
    )
  )
```

```{r}
testingData$Status <- ifelse(testingData$Status == 1, "Censored", "Deceased")

# Build Confusion Matrix for Testing Data
testingData %>%
  tabyl(var1 = model2Class, var2 = Status) %>%
  adorn_title(
    placement = "combined",
    row_name = "Predicted",
    col_name = "Actual"
  ) %>%
  kable(
    caption = "Confusion Matrix for Test data"
  )
```

-   **Accuracy**: Approximately 71.43%

-   **Recall**: Approximately 76.92%

-   **Precision**: Approximately 43.48%

-   **F1 Score**: Approximately 55.56%

This shows that our model is has struggled with overfitting, with an especially low precision score. We will discuss this in the comparison with the tree model, but this is an importnat limitation of our data size as we discussed in the introduction. Lastly, we can plot the separation plot. We can see the separation has increased due to the over fitting we discussed.

```{r}
#Sepeation Plot
par(mar = c(4,0,0,0))
separationplot(
  pred = testingData$predict, 
  actual = testingData$gradNum, 
  type = "rect",
  line = TRUE, 
  lwd2 = 2,
  show.expected = TRUE, 
  newplot = FALSE
)

```

#Comparison between models

Now let us compare the performance of our logistic regression model with a decision tree model. We will use the same train test split as before.

# Part III: Exploring the sub-clusters of the cirrhosis data using unsupervised learning

In this section, we sought to if there are sub-collections of patients. We expect sub-collections to mimic medical definitions of the stages of Biliary Cirrhosis but are curious to see if clustering can reveal some newer observations.

## Introduction

Biliary Cirrhosis patients are typically clustered in four main clusters ([Source](https://www.healthline.com/health/primary-biliary-cirrhosis#stages){.uri}):

-   Stage 1: There's inflammation and damage to the walls of medium-sized bile ducts.

-   Stage 2: There's blockage of the small bile ducts.

-   Stage 3: This stage marks the beginning of scarring.

-   Stage 4: Cirrhosis has developed. This permanent, severe, scarring and damage to the liver.

In this section we will be using clustering on our data. K-means Clustering is an unsupervised learning technique where data points are grouped based on their similarities. It's commonly used to identify patterns and structures within datasets without prior knowledge of the groups. The main advantage of clustering is its ability to discover hidden patterns in data. However, a significant drawback is the subjectivity in defining the 'similarity' criteria, which can lead to varying results and interpretations.

[![K-means clustering ([Source](https://medium.datadriveninvestor.com/k-means-clustering-4a700d4a4720))](1*fz-rjYPPRlGEMdTI-RLbDg.png)](https://medium.datadriveninvestor.com/k-means-clustering-4a700d4a4720)

An important motivation for this part of the study is explore how does "Stage" which is typically defined based on medical professionals' observations compares to the data that we have in this data set.

## Pre-processing the data

We had to pre-process the data for the best performance. First, we removed the 'Stage' column to avoid using outcome-related features in the unsupervised learning. The rows with N/A values were also removed. Lastly, we transformed various categorical variables into numeric formats, which is necessary for clustering algorithms.

```{r}
#do the ml, take about observational stages vs now quantify, look into the different variables and what they mean.
cirrhosisCluster <- cirrhosis
cirrhosisCluster2 <- na.omit(cirrhosis)

cirrhosisCluster <- cirrhosisCluster %>%                 
                                  dplyr::select(-Stage)

cirrhosisCluster <- na.omit(cirrhosisCluster) #cannot have NA values in clustering

cirrhosisCluster$Age_Years <- round(cirrhosisCluster$Age_Years) #round age to whole numbers

#reode sex... recode others later if need be #female is 0

#cirrhosisCluster  <- cirrhosisCluster %>%
#  select( -c(Status, Drug, Edema)) %>%

# mutate(Sex = ifelse(Sex == 'Female', 0, 1))

  cirrhosisCluster  <- cirrhosisCluster %>%
    mutate(Sex = ifelse(Sex == 'Female', 0, 1)) %>%
          mutate(Transplant = ifelse(Status == "CL", 1, 0)) %>%
              mutate(Status = ifelse(Status %in% c('C', 'CL'), 0 , 1)) %>%
                    mutate(Drug = ifelse(Drug == "D-penicillamine", 0, 1)) %>%
                        mutate(EdemaDiurectics = ifelse(Edema %in% c('S', 'Y'), 1, 0)) %>%
                              mutate(NoEdemaORD = ifelse(Edema == 'N' , 1, 0)) %>%
                                        mutate(EdemaANDD = ifelse(Edema == "Y", 1, 0)) %>%
                                                mutate(EdemaORD = ifelse(Edema == "S", 1, 0)) %>%
                                                                              dplyr::select(-Edema)

#Data is already factored

#mutate(Sex = ifelse(Sex == 'Female', 0, 1)) %>%

   #mutate(Status = ifelse(Status %in% c('C', 'CL'), 0 , 1)) %>%
          #    mutate(Drug = ifelse(Drug == "D-penicillamine", 0, 1))

#???:

#make column yes no edema #then yes no under treatment

#same for transplant stuff
```

```{r}
distCirrhosis <- dist(
  x = cirrhosisCluster,
  method = "euclidean"
)

```

## Methodology

To apply the k-means algorithm. A mixed hierarchical and non-hierarchical was applied. This was done using the hkmeansm module, with k = 4 as our initial value. Based on the data properties, the euclidean (L2) distance works best.

```{r}

hybridCirrhosis <- hkmeans(
  x = cirrhosisCluster,
  k = 4,
  hc.metric = "euclidean",
  hc.method = "ward.D",
  iter.max = 10
)

```

## Results

We were able to visualize the results of the k-means algorithm. The first step was making a color palette and plotting the dendrogram tree.

```{r}

StagesPalette <- c("#AA336A", "#770737", "#40B5AD", "#009E60", "#9FE2BF")

```

```{r}
## MAKE A NEW PALLETE TO VISUALIZE
# Plot the initial dendrogram for hybrid approach ----
set.seed(380)
hkmeans_tree(
  hkmeans = hybridCirrhosis,
  rect.col = StagesPalette,
  cex = 0.4,
  main = "Initial Hierarchical Clusters"
)

```

As, you can see, the model was able to create clear clusters for the data. However, it is important to find the best value of k to get the optimal clusters. To do this we can use a scree plot. Here, we\'re looking for the number of clusters that corresponds to the \"elbow\".

```{r}
# Create scree plot for choosing k ----
library(factoextra)
set.seed(380)
fviz_nbclust(
  x = cirrhosisCluster,
  diss = NULL,
  FUNcluster = kmeans,
  method = "wss",
  k.max = 10
)

```

From this, we identified that 5 was the ideal value of k, where the Total Within [Cluster] Sums of Squares begins leveling off. Let us create a new k-means model with k = 5. We can plot this refined model, with a format more easy to visualize.

```{r}

hybridCirrhosis2 <- hkmeans(
  x = cirrhosisCluster,
  k = 5,
  hc.metric = "euclidean",
  hc.method = "ward.D",
  iter.max = 10
)

```

```{r}
# Plot the final dendrogram for hybrid approach ----
# library(factoextra)
fviz_dend(
  x = hybridCirrhosis2,
  cex = 0.4,
  palette = StagesPalette,
  rect = FALSE,
  horiz = TRUE,
  repel = TRUE,
  main = "Final Dendrogram"
)

```

As you can see, the 5 clusters were identified, which are color coded in our updated diagram.

Lastly, let us plot these clusters. Here is a plot of the initial model where k = 4

```{r}
# Plot the final clustering for hybrid approach ----
# library(factoextra)
fviz_cluster(
  object = hybridCirrhosis,
  stand = FALSE,
  geom = "point",
  main = "Hybrid Cluster Plot - Initial model"
) +
  scale_color_manual(values = StagesPalette) +
  scale_fill_manual(values = StagesPalette) +
  theme_bw()

```

We can also plot our refined model, as you can see the plot below identifies 5 clear clusters.

```{r}

# Plot the final clustering for hybrid approach ----
# library(factoextra)
fviz_cluster(
  object = hybridCirrhosis2,
  stand = FALSE,
  geom = "point",
  main = "Hybrid Cluster Plot - Refined model"
) +
  scale_color_manual(values = StagesPalette) +
  scale_fill_manual(values = StagesPalette) +
  theme_bw()

```

Now we can go back to our initial goal: comparing how the custers compare to the Stage variable

```{r}
cirrhosisCluster2$cluster <- hybridCirrhosis$cluster
# Calculate mean (or median, etc.) for each variable in each cluster
library(dplyr)
cluster_summary <- cirrhosisCluster2 %>%
  group_by(cluster) %>%
  summarise_all(funs(mean(., na.rm = TRUE))) # Replace mean with median or any other function as necessary
kable(cluster_summary %>% dplyr::select(-Status, -Drug, - Sex, -Edema, -Stage), label = "Cluster Summary")
ggplot(cirrhosisCluster2, aes(x = factor(cluster), fill = factor(Stage))) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Stages within Clusters",
       x = "Cluster",
       y = "Count",
       fill = "Stage") +
  theme_minimal()

```

We can see that the clusters are not evenly distributed. This is a good sign, as it means that the clusters are not just a random grouping of the data.

# Discussion

# Author Contributions

# References--Citation style is your choice, but all sources should be documented (both in text and in the References section). This includes where you got your data.

# Code Appendix

### 2023-12-15 Regression.Rmd
Wrote the introduction for the data
idk

@@
 
-This analysis suggests significant associations between several predictors (Bilirubin, Age, Alk_Phos, Prothrombin, Spiders) and the status of the patient.
-Each predictor, including the Intercept, shows a statistically significant effect, as indicated by their small p-values (Pr(\>\|z\|)).
-The negative coefficients for Bilirubin, Age, Alk_Phos, Prothrombin, and Spiders imply that increases in these variables are associated with a decrease in the log-odds of the outcome.
-Bilirubin and Spiders are showing notably strong effects in the model.
+This is the role of inference in evaluating our model.
+The most notable predictors are Bilirubin, Age, Alk_Phos, and Prothrombin, each showing a statistically significant relationship (p \< 0.05) with the dependent variable.
+The Intercept and EdemaY have extremely significant p-values, but the practical significance of EdemaY is questionable due to its large standard error.
+Other variables like Spiders, SGOT, DrugPlacebo, and Copper, while contributing to the model, do not reach conventional levels of statistical significance (p \< 0.05).
 
@@
 In an ideal fit, the residuals should be evenly distributed about zero with constant mean and variance.
-The shape of the line suggests that the model is not capturing someunder lying structure in the data, especially in extreme cases.
+The shape of the line suggests that the model is not capturing someunder lying structure in the datain extreme cases.
 
@@
 
-Lastly, let us look at the ROC curves for bothe the models.
+Lastly, let us look at the ROC curves for both the models.
 
@@
 
-Model 1: - Its ROC curve is above the line of no discrimination, indicating that the model has some predictive capabilities.
-- The AUC is 0.64, which is better than random guessing but suggests there's room for improvement.
+Model 1:
 
-Model 2: - The ROC curve for Model 2 is significantly above that of Model 1, and much closer to the top-left corner, indicating better predictive performance.
+-   Its ROC curve is above the line of no discrimination, indicating that the model has some predictive capabilities
+
+-   The AUC is 0.638, which is better than random guessing but suggests there's room for improvement.
+
+Model 2:
+
+-   The ROC curve for Model 2 is significantly above that of Model 1, and much closer to the top-left corner, indicating better predictive performance.
+
 -   The AUC is 0.899, which suggests a good classification performance, and it's notably better than Model 1.
+
 -   Its ability to discriminate between positive and negative classes is superior to that of Model 1.
@@
 
-Using our final model, we now turn to see how well this classifier does on our testing data. Recall that we initally set the test data during the train test split.
+Using our final model, we now turn to see how well this classifier does on our testing data. Recall that we initially set the test data during the train test split.
 
@@
 This shows that our model is has struggled with overfitting, with an especially low precision score.
-We will discuss this in the comparison with the tree model, but this is a limitation of our data size.
+We will discuss this in the comparison with the tree model, but this is an importnat limitation of our data size as we discussed in the introduction.
 Lastly, we can plot the separation plot.
+We can see the separation has increased due to the over fitting we discussed.
 

### 2023-12-15 Tree.Rmd
Wrote the introduction for the data
idk

@@
 
-# Part I: Classification Algorithm for the Status of the Patient
+# Part I: Classification Algorithm for the Status of the Patient using CART
 
@@
 
-In this report, I have decided to use {rpart} because it provides more flexibility for surrogate splits and the trees are a bit easier to make attractive looking.
+In this report, we have decided to use {rpart} because it provides more flexibility for surrogate splits and the trees are a bit easier to make attractive looking.
 
@@
 
-(Explanation of above) To gain a further understanding of the data, we can plot a tree yielding Collection Node style trees. This can help us understand how the data is split.
+To gain a further understanding of the data, we can plot a tree yielding Collection Node style trees. This can help us understand how the data is split.
 
@@
 ```
+
+The tree shows us the splits that were done on Age, Bilirubin and Prothrombin. Interestingly, Stage did not contribute in the tree.
 
@@
 
-From the graph we can see that a cp of 0.011 is ideal as it is under the horizontal (dotted) reference line. We can prune the tree with this CP value.
+From the graph we can see that a cp of 0.039 is ideal as it is under the horizontal (dotted) reference line. We can prune the tree with this CP value.
 
@@
 ```
+
+We ca see the pruned tree has cut out some leaf nodes. This would help in avoiding overfitting the model.
 
@@
   booktabs = TRUE,
-  caption = "Model 1: Confusion Matrix",
+  caption = "Model 1: Confusion Matrix (0=Deceased, 1=Censored)",
   align = "c"
@@
 
-From this, we can get the model's
-
--   True Positive (TP): Predicted Graduated and actually Graduated. In this case, 2217
-
--   False Positive (FP): Predicted Graduated but actually Not Graduated. In this case, 1000
-
--   False Negative (FN): Predicted Not Graduated but actually Graduated. In this case, 244
-
--   True Negative (TN): Predicted Not Graduated and actually Not Graduated. In this case, 659
-
-With this, we can also calculate the model's metrics.
+With this, we can also calculate the model's metrics on the test data.
 
@@
 ```
+
+As you can see the model shows good accuracy with 72.9%. However, the specificity is an issue with 57.6%. Now let us compare this with a different type of model: logistic regression.

### 2023-12-16 Final Submission.Rmd
SUBMISSION is our final file

@@
 ---
-title: "Final Project Report: An exploration of Biliary Cirrhosis and Treatments"
-output: html_notebook
-authors: "Mihir Kulkarni, Nithika Menon"
+title: 'Final Project Report - A statistical exploration of Biliary Cirrhosis and Treatments'
+output:
+  pdf_document: default
+authors: Mihir Kulkarni+Nithika Menon
 ---
 
+```{r setupFiles, include=FALSE}
+knitr::opts_chunk$set(
+  echo = FALSE,
+  warning = FALSE,
+  message = FALSE,
+  fig.align = "center"
+)
+```
+
 # Background
+
+![[Primary biliary cholangitis (Source:]{.underline} <https://www.mayoclinic.org/diseases-conditions/primary-biliary-cholangitis/symptoms-causes/syc-20376874>)](ds00604_-ds00918_im02271_r7_biliarysystemthu_jpg.jpg)
 
@@
 There is no definite cure at this time.
-
-![[Primary Biliary Cholangitis ([Source](https://www.google.com/search?client=safari&sca_esv=591276386&rls=en&sxsrf=AM9HkKmtJVOC2Jrf2BSYVB2O-QcnTSgx6g:1702664705100&q=Primary+Biliary+Cholangitis&tbm=isch&source=lnms&sa=X&ved=2ahUKEwid0qyniJKDAxX9FFkFHVH3CcoQ0pQJegQIDxAB&biw=1440&bih=848&dpr=2#imgrc=UeV_fhP-8vPOlM))]{.underline}](17715-primary-biliary-cholangitis.webp)
 
@@
 To do this we can use a scree plot.
-Here, we\'re looking for the number of clusters that corresponds to the \"elbow\".
+Here, we're looking for the number of clusters that corresponds to the "elbow".
 
@@
 
+\newpage
+
 # Code Appendix
+
+```{r codeAppendix}
+#| ref.label = knitr::all_labels(),
+#| echo = TRUE,
+#| eval = FALSE
+```

### 2023-12-16 Final-Submission.tex
SUBMISSION is our final file

--- 2023-12-15 Submission.tex
@@
 \hypersetup{
-  pdftitle={Submission},
+  pdftitle={Regression-based},
+  pdfauthor={Mihir Kulkarni, Nithika Menon},
   hidelinks,
@@
 
-\title{Submission}
-\author{}
-\date{\vspace{-2.5em}}
+\title{Regression-based}
+\author{Mihir Kulkarni, Nithika Menon}
+\date{2023-12-12}
 
@@
 
-{
-\setcounter{tocdepth}{2}
-\tableofcontents
-}
+\#Final Project Report-An exploration of Biliary Cirrhosis and
+Treatments
+
 \hypertarget{background}{%
@@
 
+\hypertarget{why-is-the-data-important-to-us}{%
+\subsection{Why is the data important to
+us?}\label{why-is-the-data-important-to-us}}
+
+Both of us have been very interetsted in the intersection of medicine
+and machine learning. The idea that the concepts we are learning in our
+DS majors could have such an impact on the health industry was really
+special. Another important aspect was that there is so much potential
+for advancement of medicine, we both agreed that this area is MLs most
+promising aspect. The cirrhosis data set perticularly stood out to us
+because the data set included so many interesting attributes and we
+anted to see how they are connected with the outcomes of the patient.
+
+The data set also includes some important limitations. Most importantly,
+there are only 418 cases, greatly affecting the types of analysis we
+could do. This shows realm world problems data scientists face. Thus,
+our goal is to use our Statistical DS skillset to approach this problem,
+despite the limitations.
+
 \hypertarget{objectives}{%
@@
 
-\hypertarget{q1-what-predictors-if-any-increases-an-individuals-chance-of-receiving-a-liver-transplant}{%
-\subsubsection{Q1: What predictors, if any, increases an individual's
-chance of receiving a liver
-transplant?}\label{q1-what-predictors-if-any-increases-an-individuals-chance-of-receiving-a-liver-transplant}}
-
-\hypertarget{q2-are-there-factors-not-currently-included-in-medical-definition-that-can-help-distinguish-different-stages-of-biliary-cholangitis}{%
-\subsubsection{Q2: Are there factors not currently included in medical
+Exploring the study, we concluded on two core questions we hope to
+address in this report:
+
+\hypertarget{crq1-what-predictors-if-any-affected-the-status-of-the-patient-at-the-end-of-the-study}{%
+\subsubsection{CRQ1: What predictors, if any affected the status of the
+patient at the end of the
+study?}\label{crq1-what-predictors-if-any-affected-the-status-of-the-patient-at-the-end-of-the-study}}
+
+\hypertarget{crq2-are-there-factors-not-currently-included-in-medical-definition-that-can-help-distinguish-different-stages-of-biliary-cholangitis}{%
+\subsubsection{CRQ2: Are there factors not currently included in medical
 definition that can help distinguish different stages of Biliary
-Cholangitis?}\label{q2-are-there-factors-not-currently-included-in-medical-definition-that-can-help-distinguish-different-stages-of-biliary-cholangitis}}
+Cholangitis?}\label{crq2-are-there-factors-not-currently-included-in-medical-definition-that-can-help-distinguish-different-stages-of-biliary-cholangitis}}
 
@@
 `Status' was either Dead, Censored, or Censored due to liver transplant.
-
 A link to the original study is here:
@@
 
-\hypertarget{load-in-packages}{%
-\subsubsection{Load in Packages}\label{load-in-packages}}
+The data contains the following attributes
+(\href{https://www.kaggle.com/datasets/fedesoriano/cirrhosis-prediction-dataset/data}{Data
+Source}):
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi})}
+\tightlist
+\item
+  \textbf{ID}: unique identifier\\
+\item
+  \textbf{N\_Days}: number of days between registration and the earlier
+  of death, transplantation, or study analysis time in July 1986\\
+\item
+  \textbf{Status}: status of the patient C (censored), CL (censored due
+  to liver tx), or D (death)\\
+\item
+  \textbf{Drug}: type of drug D-penicillamine or placebo\\
+\item
+  \textbf{Age}: age in {[}days{]}\\
+\item
+  \textbf{Sex}: M (male) or F (female)\\
+\item
+  \textbf{Ascites}: presence of ascites N (No) or Y (Yes)\\
+\item
+  \textbf{Hepatomegaly}: presence of hepatomegaly N (No) or Y (Yes)\\
+\item
+  \textbf{Spiders}: presence of spiders N (No) or Y (Yes)\\
+\item
+  \textbf{Edema}: presence of edema N (no edema and no diuretic therapy
+  for edema), S (edema present without diuretics, or edema resolved by
+  diuretics), or Y (edema despite diuretic therapy)\\
+\item
+  \textbf{Bilirubin}: serum bilirubin in {[}mg/dl{]}\\
+\item
+  \textbf{Cholesterol}: serum cholesterol in {[}mg/dl{]}\\
+\item
+  \textbf{Albumin}: albumin in {[}gm/dl{]}\\
+\item
+  \textbf{Copper}: urine copper in {[}ug/day{]}\\
+\item
+  \textbf{Alk\_Phos}: alkaline phosphatase in {[}U/liter{]}\\
+\item
+  \textbf{SGOT}: SGOT in {[}U/ml{]}\\
+\item
+  \textbf{Triglycerides}: triglicerides in {[}mg/dl{]}\\
+\item
+  \textbf{Platelets}: platelets per cubic {[}ml/1000{]}\\
+\item
+  \textbf{Prothrombin}: prothrombin time in seconds {[}s{]}\\
+\item
+  \textbf{Stage}: histologic stage of disease (1, 2, 3, or 4)
+\end{enumerate}
+
+For the basic pre-processing, we binary-encoded the categorical
+variables in the data set and set them as factors.
 
@@
 \end{verbatim}
-
-\hypertarget{data-processing}{%
-\subsubsection{Data Processing}\label{data-processing}}
 
@@
 \end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{head}\NormalTok{(cirrhosis)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## # A tibble: 6 x 20
-##      ID N_Days Status Drug          Age Sex   Ascites Hepatomegaly Spiders Edema
-##   <dbl>  <dbl> <chr>  <chr>       <dbl> <chr> <chr>   <chr>        <chr>   <chr>
-## 1     1    400 D      D-penicill~ 21464 F     Y       Y            Y       Y    
-## 2     2   4500 C      D-penicill~ 20617 F     N       Y            Y       N    
-## 3     3   1012 D      D-penicill~ 25594 M     N       N            N       S    
-## 4     4   1925 D      D-penicill~ 19994 F     N       Y            Y       S    
-## 5     5   1504 CL     Placebo     13918 F     N       Y            Y       N    
-## 6     6   2503 D      Placebo     24201 F     N       Y            N       N    
-## # i 10 more variables: Bilirubin <dbl>, Cholesterol <dbl>, Albumin <dbl>,
-## #   Copper <dbl>, Alk_Phos <dbl>, SGOT <dbl>, Tryglicerides <dbl>,
-## #   Platelets <dbl>, Prothrombin <dbl>, Stage <dbl>
-\end{verbatim}
-
-\hypertarget{pre-process-the-data}{%
-\subsubsection{Pre-process the data}\label{pre-process-the-data}}
 
@@
 
-\begin{itemize}
-\item
-  Engage in Exploratory Data Analysis
-\item
-  Use Machine Learning Algorithms:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    You must use at least one of each: a Regression-based algorithm, a
-    Tree-based algorithm, and an Unsupervised Learning algorithm.
-  \item
-    Each algorithm must be explained for a general audience.
-  \item
-    All results should be explained.
-  \item
-    Where appropriate, comparisons between methods should be discussed.
-    This includes highlighting the strengths and weaknesses of the
-    various approaches as well as when each is appropriate.
-  \end{itemize}
-\item
-  A discussion of what you've learned from your work about the context
-  and recommendations for future work.
-\item
-  Author Contributions--list out who did what in the project.
-\item
-  References--Citation style is your choice, but all sources should be
-  documented (both in text and in the References section). This includes
-  where you got your data.
-\item
-  A commented Code Appendix
-\end{itemize}
-
-\begin{Shaded}
-\begin{Highlighting}[]
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#?????}
 \NormalTok{cirrhosis }\SpecialCharTok{\%\textgreater{}\%} 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-9-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-3-1.pdf}
 
@@
 
-\hypertarget{eda}{%
-\section{EDA}\label{eda}}
-
-\hypertarget{exploring-the-demographics-of-the-patients}{%
-\subsection{Exploring the demographics of the
-patients}\label{exploring-the-demographics-of-the-patients}}
+\hypertarget{exploratory-data-analysis}{%
+\section{Exploratory Data Analysis}\label{exploratory-data-analysis}}
+
+\hypertarget{part-1-exploring-the-demographics-of-the-patients}{%
+\subsection{Part 1: Exploring the demographics of the
+patients}\label{part-1-exploring-the-demographics-of-the-patients}}
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-11-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-5-1.pdf}
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-12-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-6-1.pdf}
 
@@
 
-\hypertarget{looking-at-the-data-itself}{%
-\subsection{Looking at the data
-itself}\label{looking-at-the-data-itself}}
+\hypertarget{part-2-looking-at-the-data-itself}{%
+\subsection{Part 2: Looking at the data
+itself}\label{part-2-looking-at-the-data-itself}}
 
@@
 An observation is that many of the variables have 106 NAs. This
-indicates that a good fraction {[}106 patients{]} were equally tracked
-and measured in a less extensive way.
+indicates that a good fraction {[}106 patients{]} could have been
+equally tracked and measured in a less extensive way.
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-14-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-8-1.pdf}
 
@@
 
-\hypertarget{exploring-some-interactions-of-variables}{%
-\subsection{Exploring some interactions of
-variables}\label{exploring-some-interactions-of-variables}}
+\hypertarget{part-3-exploring-some-interactions-of-variables}{%
+\subsection{Part 3: Exploring some interactions of
+variables}\label{part-3-exploring-some-interactions-of-variables}}
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-15-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-9-1.pdf}
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-16-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-10-1.pdf}
 
@@
 
-\hypertarget{mihir-choose-your-preffered-graph-and-explain}{%
-\subsubsection{mihir choose your preffered graph and
-explain}\label{mihir-choose-your-preffered-graph-and-explain}}
-
 \hypertarget{survival-analysis-based-on-treatment-and-stage-how-does-the-survival-rate-differ-among-patients-at-different-stages-of-pbc-who-received-d-penicillamine-versus-placebo}{%
@@
 treatment plans and for researchers to understand the drug's impact.
-(LOLL this is what the study acc sought to discover)
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-17-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-11-1.pdf}
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-18-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-12-1.pdf}
 
@@
 
-\hypertarget{part-i-classification-algorithm-for-the-status-of-the-patient}{%
-\section{Part I: Classification Algorithm for the Status of the
-Patient}\label{part-i-classification-algorithm-for-the-status-of-the-patient}}
+\hypertarget{part-i-classification-algorithm-for-the-status-of-the-patient-using-cart}{%
+\section{Part I: Classification Algorithm for the Status of the Patient
+using
+CART}\label{part-i-classification-algorithm-for-the-status-of-the-patient-using-cart}}
 
@@
 
-In this report, I have decided to use \{rpart\} because it provides more
-flexibility for surrogate splits and the trees are a bit easier to make
-attractive looking.
+In this report, we have decided to use \{rpart\} because it provides
+more flexibility for surrogate splits and the trees are a bit easier to
+make attractive looking.
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-22-1.pdf}
-
-(Explanation of above) To gain a further understanding of the data, we
-can plot a tree yielding Collection Node style trees. This can help us
-understand how the data is split.
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-16-1.pdf}
+
+To gain a further understanding of the data, we can plot a tree yielding
+Collection Node style trees. This can help us understand how the data is
+split.
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-23-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-17-1.pdf}
+
+The tree shows us the splits that were done on Age, Bilirubin and
+Prothrombin. Interestingly, Stage did not contribute in the tree.
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-25-1.pdf}
-
-From the graph we can see that a cp of 0.011 is ideal as it is under the
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-19-1.pdf}
+
+From the graph we can see that a cp of 0.039 is ideal as it is under the
 horizontal (dotted) reference line. We can prune the tree with this CP
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-27-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-21-1.pdf}
+
+We ca see the pruned tree has cut out some leaf nodes. This would help
+in avoiding overfitting the model.
 
@@
   \AttributeTok{booktabs =} \ConstantTok{TRUE}\NormalTok{,}
-  \AttributeTok{caption =} \StringTok{"Model 1: Confusion Matrix"}\NormalTok{,}
+  \AttributeTok{caption =} \StringTok{"Model 1: Confusion Matrix (0=Deceased, 1=Censored)"}\NormalTok{,}
   \AttributeTok{align =} \StringTok{"c"}
@@
 
-\caption{\label{tab:unnamed-chunk-29}Model 1: Confusion Matrix}
+\caption{\label{tab:unnamed-chunk-23}Model 1: Confusion Matrix (0=Deceased, 1=Censored)}
 \centering
@@
 
-From this, we can get the model's
-
-\begin{itemize}
-\item
-  True Positive (TP): Predicted Graduated and actually Graduated. In
-  this case, 2217
-\item
-  False Positive (FP): Predicted Graduated but actually Not Graduated.
-  In this case, 1000
-\item
-  False Negative (FN): Predicted Not Graduated but actually Graduated.
-  In this case, 244
-\item
-  True Negative (TN): Predicted Not Graduated and actually Not
-  Graduated. In this case, 659
-\end{itemize}
-
-With this, we can also calculate the model's metrics.
+With this, we can also calculate the model's metrics on the test data.
 
@@
 \end{tabular}
+
+As you can see the model shows good accuracy with 72.9\%. However, the
+specificity is an issue with 57.6\%. Now let us compare this with a
+different type of model: logistic regression.
 
@@
 
-\caption{\label{tab:unnamed-chunk-36}Model 1 Confusion Matrix}
+\caption{\label{tab:unnamed-chunk-30}Model 1 Confusion Matrix}
 \centering
@@
 
-This analysis suggests significant associations between several
-predictors (Bilirubin, Age, Alk\_Phos, Prothrombin, Spiders) and the
-status of the patient. Each predictor, including the Intercept, shows a
-statistically significant effect, as indicated by their small p-values
-(Pr(\textgreater\textbar z\textbar)). The negative coefficients for
-Bilirubin, Age, Alk\_Phos, Prothrombin, and Spiders imply that increases
-in these variables are associated with a decrease in the log-odds of the
-outcome. Bilirubin and Spiders are showing notably strong effects in the
-model.
+This is the role of inference in evaluating our model. The most notable
+predictors are Bilirubin, Age, Alk\_Phos, and Prothrombin, each showing
+a statistically significant relationship (p \textless{} 0.05) with the
+dependent variable. The Intercept and EdemaY have extremely significant
+p-values, but the practical significance of EdemaY is questionable due
+to its large standard error. Other variables like Spiders, SGOT,
+DrugPlacebo, and Copper, while contributing to the model, do not reach
+conventional levels of statistical significance (p \textless{} 0.05).
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-38-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-32-1.pdf}
 
@@
 suggests that the model is not capturing someunder lying structure in
-the data, especially in extreme cases.
+the datain extreme cases.
 
@@
 
-\caption{\label{tab:unnamed-chunk-40}Confusion matrix for Model 2}
+\caption{\label{tab:unnamed-chunk-34}Confusion matrix for Model 2}
 \centering
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-42-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-36-1.pdf}
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-43-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-37-1.pdf}
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-45-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-39-1.pdf}
 
@@
 
-Model 1: - Its ROC curve is above the line of no discrimination,
-indicating that the model has some predictive capabilities. - The AUC is
-0.64, which is better than random guessing but suggests there's room for
-improvement.
-
-Model 2: - The ROC curve for Model 2 is significantly above that of
-Model 1, and much closer to the top-left corner, indicating better
-predictive performance. - The AUC is 0.899, which suggests a good
-classification performance, and it's notably better than Model 1. - Its
-ability to discriminate between positive and negative classes is
+Model 1:
+
+\begin{itemize}
+\item
+  Its ROC curve is above the line of no discrimination, indicating that
+  the model has some predictive capabilities
+\item
+  The AUC is 0.638, which is better than random guessing but suggests
+  there's room for improvement.
+\end{itemize}
+
+Model 2:
+
+\begin{itemize}
+\item
+  The ROC curve for Model 2 is significantly above that of Model 1, and
+  much closer to the top-left corner, indicating better predictive
+  performance.
+\item
+  The AUC is 0.899, which suggests a good classification performance,
+  and it's notably better than Model 1.
+\item
+  Its ability to discriminate between positive and negative classes is
   superior to that of Model 1.
+\end{itemize}
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-46-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-40-1.pdf}
 
@@
 Using our final model, we now turn to see how well this classifier does
-on our testing data. Recall that we initally set the test data during
+on our testing data. Recall that we initially set the test data during
 the train test split.
@@
 
-\caption{\label{tab:unnamed-chunk-48}Confusion Matrix for Test data}
+\caption{\label{tab:unnamed-chunk-42}Confusion Matrix for Test data}
 \centering
@@
 especially low precision score. We will discuss this in the comparison
-with the tree model, but this is a limitation of our data size. Lastly,
-we can plot the separation plot.
+with the tree model, but this is an importnat limitation of our data
+size as we discussed in the introduction. Lastly, we can plot the
+separation plot. We can see the separation has increased due to the over
+fitting we discussed.
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-49-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-43-1.pdf}
+
+\#Comparison between models
+
+Now let us compare the performance of our logistic regression model with
+a decision tree model. We will use the same train test split as before.
 
@@
 \NormalTok{cirrhosisCluster }\OtherTok{\textless{}{-}}\NormalTok{ cirrhosis}
+\NormalTok{cirrhosisCluster2 }\OtherTok{\textless{}{-}} \FunctionTok{na.omit}\NormalTok{(cirrhosis)}
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-54-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-48-1.pdf}
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-55-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-49-1.pdf}
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-57-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-51-1.pdf}
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-58-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-52-1.pdf}
 
@@
 
-\includegraphics{Submission_files/figure-latex/unnamed-chunk-59-1.pdf}
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-53-1.pdf}
+
+Now we can go back to our initial goal: comparing how the custers
+compare to the Stage variable
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{cirrhosisCluster2}\SpecialCharTok{$}\NormalTok{cluster }\OtherTok{\textless{}{-}}\NormalTok{ hybridCirrhosis}\SpecialCharTok{$}\NormalTok{cluster}
+\CommentTok{\# Calculate mean (or median, etc.) for each variable in each cluster}
+\FunctionTok{library}\NormalTok{(dplyr)}
+\NormalTok{cluster\_summary }\OtherTok{\textless{}{-}}\NormalTok{ cirrhosisCluster2 }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{group\_by}\NormalTok{(cluster) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{summarise\_all}\NormalTok{(}\FunctionTok{funs}\NormalTok{(}\FunctionTok{mean}\NormalTok{(., }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{))) }\CommentTok{\# Replace mean with median or any other function as necessary}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: `funs()` was deprecated in dplyr 0.8.0.
+## i Please use a list of either functions or lambdas:
+## 
+## # Simple named list: list(mean = mean, median = median)
+## 
+## # Auto named with `tibble::lst()`: tibble::lst(mean, median)
+## 
+## # Using lambdas list(~ mean(., trim = .2), ~ median(., na.rm = TRUE))
+## Call `lifecycle::last_lifecycle_warnings()` to see where this warning was
+## generated.
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: There were 20 warnings in `summarise()`.
+## The first warning was:
+## i In argument: `Status = mean(Status, na.rm = TRUE)`.
+## i In group 1: `cluster = 1`.
+## Caused by warning in `mean.default()`:
+## ! argument is not numeric or logical: returning NA
+## i Run `dplyr::last_dplyr_warnings()` to see the 19 remaining warnings.
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{kable}\NormalTok{(cluster\_summary }\SpecialCharTok{\%\textgreater{}\%}\NormalTok{ dplyr}\SpecialCharTok{::}\FunctionTok{select}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{Status, }\SpecialCharTok{{-}}\NormalTok{Drug, }\SpecialCharTok{{-}}\NormalTok{ Sex, }\SpecialCharTok{{-}}\NormalTok{Edema, }\SpecialCharTok{{-}}\NormalTok{Stage), }\AttributeTok{label =} \StringTok{"Cluster Summary"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{tabular}{r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r}
+\hline
+cluster & N\_Days & Age & Ascites & Hepatomegaly & Spiders & Bilirubin & Cholesterol & Albumin & Copper & Alk\_Phos & SGOT & Tryglicerides & Platelets & Prothrombin & Age\_Years\\
+\hline
+1 & 1475.862 & 23503.95 & 0.1724138 & 0.5517241 & 0.3103448 & 4.034483 & 327.0862 & 3.337069 & 107.77586 & 1492.241 & 120.0178 & 126.7414 & 248.3276 & 11.15517 & 64.35030\\
+\hline
+2 & 2579.950 & 18505.65 & 0.1000000 & 0.6500000 & 0.3500000 & 4.275000 & 376.3500 & 3.372000 & 149.15000 & 8394.200 & 136.8590 & 145.4000 & 263.0500 & 10.98500 & 50.66571\\
+\hline
+3 & 2074.659 & 13933.03 & 0.0227273 & 0.4204545 & 0.2954545 & 2.842045 & 426.8977 & 3.629773 & 93.68182 & 1582.877 & 130.9384 & 119.4205 & 281.6591 & 10.46591 & 38.14657\\
+\hline
+4 & 2058.918 & 18734.59 & 0.0454545 & 0.5454545 & 0.2636364 & 3.186364 & 349.1182 & 3.547546 & 93.94545 & 1430.342 & 118.5102 & 124.7818 & 252.7182 & 10.68455 & 51.29251\\
+\hline
+\end{tabular}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{ggplot}\NormalTok{(cirrhosisCluster2, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =} \FunctionTok{factor}\NormalTok{(cluster), }\AttributeTok{fill =} \FunctionTok{factor}\NormalTok{(Stage))) }\SpecialCharTok{+}
+  \FunctionTok{geom\_bar}\NormalTok{(}\AttributeTok{position =} \StringTok{"dodge"}\NormalTok{) }\SpecialCharTok{+}
+  \FunctionTok{labs}\NormalTok{(}\AttributeTok{title =} \StringTok{"Distribution of Stages within Clusters"}\NormalTok{,}
+       \AttributeTok{x =} \StringTok{"Cluster"}\NormalTok{,}
+       \AttributeTok{y =} \StringTok{"Count"}\NormalTok{,}
+       \AttributeTok{fill =} \StringTok{"Stage"}\NormalTok{) }\SpecialCharTok{+}
+  \FunctionTok{theme\_minimal}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{Final-Submission_files/figure-latex/unnamed-chunk-54-1.pdf}
+
+We can see that the clusters are not evenly distributed. This is a good
+sign, as it means that the clusters are not just a random grouping of
+the data.
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+\hypertarget{author-contributions}{%
+\section{Author Contributions}\label{author-contributions}}
+
+\hypertarget{referencescitation-style-is-your-choice-but-all-sources-should-be-documented-both-in-text-and-in-the-references-section.-this-includes-where-you-got-your-data.}{%
+\section{References--Citation style is your choice, but all sources
+should be documented (both in text and in the References section). This
+includes where you got your
+data.}\label{referencescitation-style-is-your-choice-but-all-sources-should-be-documented-both-in-text-and-in-the-references-section.-this-includes-where-you-got-your-data.}}
+
+\hypertarget{code-appendix}{%
+\section{Code Appendix}\label{code-appendix}}
 

### 2023-12-16 Submission.Rmd
SUBMISSION is our final file

@@
 ---
-title: "Submission"
+title: 'Final Project Report - An exploration of Biliary Cirrhosis and Treatments'
 output:
-  pdf_document:
-    toc: yes
-  html_document:
-    toc: yes
-    df_print: paged
+  pdf_document: default
+authors: Mihir Kulkarni+Nithika Menon
 ---
 
-```{r child = 'Final Project.Rmd'}
+```{r setupFiles, include=FALSE}
+knitr::opts_chunk$set(
+  echo = FALSE,
+  warning = FALSE,
+  message = FALSE,
+  fig.align = "center"
+)
 ```
 
-```{r child = 'EDA.Rmd'}
+# Background
+
+## What is Biliary Cholangitis?
+
+Primary Biliary Cholangitis (previously called Primary Biliary Cirrhosis) is an autoimmune disease where bile ducts become swollen and inflamed and block the flow of bile.
+Bile is a substance that aids with digestion.
+The bile ducts carry bile from the liver to the small intestine.
+The swelling and inflammation can lead to scarring of the liver which is cirrhosis.
+Advanced cirrhosis can lead to liver failure or liver cancer.
+Medication can slow progression.
+There is no definite cure at this time.
+
+![Primary biliary cholangitis (https://www.mayoclinic.org/diseases-conditions/primary-biliary-cholangitis/symptoms-causes/syc-20376874)](diagram.jpeg)
+
+## Why is the data important to us?
+
+Both of us have been very interetsted in the intersection of medicine and machine learning.
+The idea that the concepts we are learning in our DS majors could have such an impact on the health industry was really special.
+Another important aspect was that there is so much potential for advancement of medicine, we both agreed that this area is MLs most promising aspect.
+The cirrhosis data set perticularly stood out to us because the data set included so many interesting attributes and we anted to see how they are connected with the outcomes of the patient.
+
+The data set also includes some important limitations.
+Most importantly, there are only 418 cases, greatly affecting the types of analysis we could do.
+This shows realm world problems data scientists face.
+Thus, our goal is to use our Statistical DS skillset to approach this problem, despite the limitations.
+
+## Objectives
+
+Exploring the study, we concluded on two core questions we hope to address in this report:
+
+### CRQ1: What predictors, if any affected the status of the patient at the end of the study?
+
+### CRQ2: Are there factors not currently included in medical definition that can help distinguish different stages of Biliary Cholangitis?
+
+## Clinical Study + Dataset Background :
+
+The dataset is produced from a clinical study by Mayo Clinic run from 1974-1984.
+The final 'Status' of each patient was observed in 1986.
+'Status' was either Dead, Censored, or Censored due to liver transplant.
+A link to the original study is here: <https://faculty.washington.edu/abansal/ShortCourse_DynamicDecisionMaking/Dickson1989_MayoPBCOriginalArticle.pdf>
+
+The data contains the following attributes ([Data Source](https://www.kaggle.com/datasets/fedesoriano/cirrhosis-prediction-dataset/data)):
+
+1)  **ID**: unique identifier\
+2)  **N_Days**: number of days between registration and the earlier of death, transplantation, or study analysis time in July 1986\
+3)  **Status**: status of the patient C (censored), CL (censored due to liver tx), or D (death)\
+4)  **Drug**: type of drug D-penicillamine or placebo\
+5)  **Age**: age in [days]\
+6)  **Sex**: M (male) or F (female)\
+7)  **Ascites**: presence of ascites N (No) or Y (Yes)\
+8)  **Hepatomegaly**: presence of hepatomegaly N (No) or Y (Yes)\
+9)  **Spiders**: presence of spiders N (No) or Y (Yes)\
+10) **Edema**: presence of edema N (no edema and no diuretic therapy for edema), S (edema present without diuretics, or edema resolved by diuretics), or Y (edema despite diuretic therapy)\
+11) **Bilirubin**: serum bilirubin in [mg/dl]\
+12) **Cholesterol**: serum cholesterol in [mg/dl]\
+13) **Albumin**: albumin in [gm/dl]\
+14) **Copper**: urine copper in [ug/day]\
+15) **Alk_Phos**: alkaline phosphatase in [U/liter]\
+16) **SGOT**: SGOT in [U/ml]\
+17) **Triglycerides**: triglicerides in [mg/dl]\
+18) **Platelets**: platelets per cubic [ml/1000]\
+19) **Prothrombin**: prothrombin time in seconds [s]\
+20) **Stage**: histologic stage of disease (1, 2, 3, or 4)
+
+For the basic pre-processing, we binary-encoded the categorical variables in the data set and set them as factors.
+
+```{r load packages}
+
+library(survival)
+library(ggplot2)
+library(survival)
+library(survminer)
+library(kableExtra)
+library(factoextra)
+
 ```
 
-```{r child = 'Tree.Rmd'}
+```{r}
+#load in the data
+library(tidyverse)
+#load data
+cirrhosis <- read_csv("cirrhosis.csv")
 ```
 
-```{r child = 'Regression.Rmd'}
+```{r}
+#encode the sex, ascites, hepatomegaly, spiders, edema into binary variables
+cirrhosis$Sex <- ifelse(cirrhosis$Sex == "F", 0, 1)
+cirrhosis$Ascites <- ifelse(cirrhosis$Ascites == "Y", 1, 0)
+cirrhosis$Hepatomegaly <- ifelse(cirrhosis$Hepatomegaly == "Y", 1, 0)
+cirrhosis$Spiders <- ifelse(cirrhosis$Spiders == "Y", 1, 0)
+#factorize the stage
+cirrhosis$Stage <- factor(cirrhosis$Stage)
+cirrhosis$Status <- factor(cirrhosis$Status)
+cirrhosis$Drug <- factor(cirrhosis$Drug)
+cirrhosis$Sex <- factor(cirrhosis$Sex)
+#cirrhosis$Ascites <- factor(cirrhosis$Ascites)
+#cirrhosis$Hepatomegaly <- factor(cirrhosis$Hepatomegaly)
+#cirrhosis$Spiders <- factor(cirrhosis$Spiders)
+cirrhosis$Edema <- factor(cirrhosis$Edema)
+cirrhosis$Sex <- factor(ifelse(cirrhosis$Sex == 0, "Female", "Male"))
+
+#Drop the first column
+cirrhosis <- cirrhosis[,-1]
 ```
 
-```{r child = 'Unsupervised.Rmd'}
+```{r}
+#?????
+cirrhosis %>% 
+  ggplot(aes(x = N_Days)) + geom_histogram() +
+  facet_wrap(facets = vars(Status))
+
 ```
+
+```{r}
+#Required libraries
+library(survival)
+library(ggplot2)
+library(survival)
+library(survminer)
+
+```
+
+# Exploratory Data Analysis
+
+## Part 1: Exploring the demographics of the patients
+
+### Sex Ratio of Patients
+
+```{r}
+
+#cirrhosis$Sex <- factor(ifelse(cirrhosis$Sex == 0, "Female", "Male"))
+
+# Preparing the data for pie chart
+sex_data <- cirrhosis %>% 
+            count(Sex) %>% 
+            mutate(Percentage = n / sum(n) * 100)
+
+# Pie Chart
+ggplot(sex_data, aes(x="", y=Percentage, fill=Sex)) +
+    geom_bar(stat="identity", width=1) +
+    coord_polar("y", start=0) +
+    scale_fill_brewer(palette="Pastel1") +
+    theme_minimal() +
+    theme(axis.line = element_blank(),
+          axis.text = element_blank(),
+          axis.ticks = element_blank(),
+          panel.grid = element_blank(),
+          plot.title = element_text(size=14, face="bold")) +
+    labs(fill="Sex", title=" Sex Amongst Patients")
+
+```
+
+Primary Biliary Cholangitis affects women much more than men at a 10-1 ratio.
+The data of this clinical study mimics the greater population.
+Investigating gender-based differences in disease progression can uncover any gender-specific patterns in PBC.
+This could lead to gender-tailored treatment approaches and a better understanding of the disease's biology, which might differ between males and females.
+
+### Age Distribution of Patients
+
+```{r}
+# Converting age from days to years for better readability
+cirrhosis$Age_Years <- cirrhosis$Age / 365.25
+
+# Histogram
+ggplot(cirrhosis, aes(x=Age_Years)) +
+    geom_histogram(binwidth=5, fill="#69b3a2", color="#e9ecef") +
+    theme_minimal() +
+    labs(title="Age Distribution of PBC Patients", x="Age (years)", y="Count") +
+    theme(plot.title = element_text(size=14, face="bold"))
+
+```
+
+The distribution of ages is quite normal.
+
+## Part 2: Looking at the data itself
+
+### NA counts amongst our variables
+
+```{r}
+count <-  data.frame(NA_CountsPerVar = colSums(is.na(cirrhosis)))
+kable(count)
+```
+
+An observation is that many of the variables have 106 NAs. This indicates that a good fraction [106 patients] could have been equally tracked and measured in a less extensive way.
+
+## Distribution of Stages of Biliary Cholangitis
+
+```{r}
+
+cirrhosis %>%
+      ggplot(aes(x = Stage)) + geom_bar(fill = "#097969") # need color
+
+```
+
+This distribution is left skewed and not symmetric. Most patients have stage 3 and then 4 of Biliary Cholangitis.
+
+## Part 3: Exploring some interactions of variables
+
+### How does prescense of a D-penicillamine with stage of Biliary Cholangitis impact number of days till death?
+
+```{r}
+
+ggplot(data = cirrhosis, aes (Drug, Stage, fill = N_Days))+ geom_tile() + scale_fill_distiller(palette = "RdPu", trans = "reverse")
+
+```
+
+This heat map shows that being a placebo in stage 1 gives you a greater amount of days till death in this sample of patients.
+
+### Individuals with Edema, and if they were on the drug
+
+```{r}
+
+hi = c("#40B5AD", "#009E60", "#9FE2BF")
+library(ggmosaic)
+cirrhosis %>%
+  ggplot() +
+  geom_mosaic(aes( x = product(Edema), fill = Drug)) + scale_fill_manual(values = hi)
+```
+
+Edema is swelling due to too much liquid trapped in the body's tissues.
+It's common complication in liver diseases and can significantly impact patient quality of life and survival.
+Understanding how different levels of edema (none, controlled by diuretics, or persistent despite treatment) affect survival can inform patient management strategies and highlight the need for aggressive interventions in certain cases.
+From the graph we can see a even split between the placebo and drug for Edema status.
+There is no NA values for Edema persistent despite diuretics.
+We cannot directly compare it the splits for other statuses given their NA values if known could change the look of the graph.
+
+### Survival Analysis Based on Treatment and Stage: How does the survival rate differ among patients at different stages of PBC who received D-penicillamine versus placebo?
+
+This question investigates the effectiveness of the drug D-penicillamine compared to a placebo, considering the stage of PBC.
+By analyzing survival rates across different disease stages and treatment types, we can assess the drug's effectiveness at various disease stages.
+This information is vital for clinicians to make informed decisions about treatment plans and for researchers to understand the drug's impact.
+
+```{r}
+surv_object <- Surv(cirrhosis$N_Days, cirrhosis$Status == 'D')
+surv_fit <- survfit(surv_object ~ cirrhosis$Drug + cirrhosis$Stage, data = cirrhosis)
+ggsurvplot(surv_fit, 
+           data = cirrhosis,
+           conf.int = TRUE, 
+           #palette = c("#00AFBB", "#E7B800", "#FC4E07"),
+           xlab = "Days", 
+           ylab = "Survival Probability", 
+           title = "Survival Curves by Treatment and Stage")
+```
+
+### Impact of Liver Complications on Survival: How do the presence of ascites, hepatomegaly, and spiders relate to survival time?
+
+Exploring how liver-related symptoms (ascites, hepatomegaly, and spiders) affect patient survival provides insights into the severity of these complications in PBC progression.
+This can help in identifying high-risk patients and understanding the disease's impact on liver function.
+
+```{r}
+surv_fit_complications <- survfit(surv_object ~ cirrhosis$Ascites + cirrhosis$Hepatomegaly + cirrhosis$Spiders, data = cirrhosis)
+ggsurvplot(surv_fit_complications, 
+           data = cirrhosis,
+           conf.int = TRUE, 
+           #palette = c("#2E9FDF", "#FC4E07", "#6ACC65"),
+           xlab = "Days", 
+           ylab = "Survival Probability", 
+           title = "Survival Curves by Liver Complications")
+```
+
+### Correlation between Biochemical Markers and Disease Stage: What is the correlation between biochemical markers (Bilirubin, Cholesterol, Albumin, Copper, Alk_Phos, SGOT, Triglycerides) and the histologic stage of the disease?
+
+This analysis is crucial to understand how different biochemical markers correlate with the disease's progression.
+Such correlations can aid in the early detection of disease severity, help in monitoring the disease progression, and potentially guide treatment adjustments.
+
+# Part I: Classification Algorithm for the Status of the Patient using CART
+
+To address the first sub-goal of this project, we will be exploring the prediction of the status of a patient at the end of the study.
+Our main objective through this is to gain a stronger understanding of what factors played in the role of the death of the patient.
+To do this, we will be using Decision Trees.
+Let us start by exploring what these are and why we chose to use them.
+
+## Introduction
+
+Decision trees are a type of model used in statistics for making predictions based on data.
+They work by breaking down a dataset into smaller subsets through "splits," resembling a tree with branches.
+Each branch represents a possible decision or outcome, leading to a final prediction or classification.
+
+The main advantages of decision trees include their simplicity and interpretability, as they are easy to visualize and understand.
+Another important advantage is that they are able to learn with data with N/A values, something alternatives such as Logistic Regression.
+Considering the limitation we have with our data size and number of N/A values, this an important advantage.
+However, there are some downsides such as a tendency to overfit the data, which is something we need to be careful about.
+
+![[Decision Tree Example (]{.underline}[Source](https://data-flair.training/blogs/r-decision-trees/))](DecTreeEg.png)
+
+## Pre-processing the data
+
+For this data, we will be dealing with the Status variable. The Status variable was divided into 3 classes:
+
+-   Class 0 (D): The patient didn't survive by the end of the observation
+-   Class 1 (C): The patient is censored, meaning that the observation period ended without the death being recorded
+-   Class 2 (CL): Similar to Class 1, the patient is censored due to liver transplantation
+
+Thus, we can group Class 1 and 2.
+
+```{r}
+#Combine C and CL status into one variable and binarize
+cirrhosisTreeData <- cirrhosis
+cirrhosisTreeData$Status <- ifelse(cirrhosisTreeData$Status == "C" | cirrhosisTreeData$Status == "CL", 1, 0)
+```
+
+## Methodology
+
+Our objective is to create a classifier capable of predicting a patient's outcome.
+To achieve this, we will be testing our data with the cart algorithm.
+To ensure model validation, we'll be using a 80% training and 20% testing data division.
+I will also be stratifying the data based on the status.
+
+To guarantee consistency and reproducibility in our results, we have fixed the seed for our 80/20 data split at 380.
+With these steps, we are now well-positioned to finalize our training and testing data sets.
+
+The predictors in these models will be guided by the results from the EDA. We will also use a variety of tools to understand the model's performance.
+
+```{r}
+# Wrangle the Graduate data to set up training and testing datasets
+modelCirrhosis <- cirrhosisTreeData %>%
+  #drop_na() %>%
+  mutate(
+    tempID = row_number(),
+    .before = Status
+  )
+
+## Set seed for reproducibility and slice ----
+set.seed(380)
+trainingData <- modelCirrhosis %>%
+  group_by(Status) %>%
+  slice_sample(prop = 0.8)
+
+testingData <- modelCirrhosis %>%
+  filter(!(tempID %in% trainingData$tempID))
+```
+
+### Step 1: Growing the tree
+
+There are useful packages to build decision trees in R: the {tree} and the {rpart} (recursive partitioning) packages.
+
+In this report, we have decided to use {rpart} because it provides more flexibility for surrogate splits and the trees are a bit easier to make attractive looking.
+
+```{r}
+# Grow Graduate tree via rpart package
+library(rpart)
+rPartStatus <- rpart(
+  formula = Status ~ Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage + Tryglicerides, 
+  data = trainingData,
+  method = "class",
+  parms = list(split = "information")
+  # We did not need to use the control parameters
+)
+
+```
+
+### Part 2: Visualizing the tree
+
+With the tree grown, we can now visualize it for an easy understanding of its functioning. This is an important advantage for CART over logistic regression.
+
+The following is a basic diagram for the tree that was just grown.
+
+```{r}
+# Display rpart.plot ----
+ library(rpart.plot)
+rpart.plot(
+  x = rPartStatus,
+  type = 2,
+  extra = 101
+)
+```
+
+To gain a further understanding of the data, we can plot a tree yielding Collection Node style trees. This can help us understand how the data is split.
+
+```{r}
+# Using the rattle package to visualize the tree ----
+library(rattle)
+
+fancyRpartPlot(
+  model = rPartStatus,
+  main = NULL,
+  sub = NULL
+)
+```
+
+The tree shows us the splits that were done on Age, Bilirubin and Prothrombin. Interestingly, Stage did not contribute in the tree.
+
+### Part 3: Pruning the tree
+
+Pruning reduces the size of decision trees by removing parts of the tree that do not provide power to classify instances.
+The first step of pruning a tree is understanding the complexity parameter used.
+The complexity parameter (cp) in rpart is the minimum improvement in the model needed at each node.
+This is used when building the tree.
+We can see the results based on the cross validation from the table below.
+
+```{r}
+invisible(capture.output({cpTable <- printcp(rPartStatus)}))
+
+library(kableExtra)
+
+kable(
+  x = cpTable,
+  col.names = c("CP", "Num. of splits", "Rel. Error",
+                "Mean Error", "Std. Deviation of Error"),
+  digits = 3,
+  booktabs = TRUE,
+  align = "c",
+  table.attr = 'data-quarto-disable-processing="true"'
+)
+```
+
+This can also be visualized in a graph to gain a better understanding of the data. The graph below shows the connection between the cp, size of tree and the x-val relative error.
+
+```{r}
+plotcp(
+  x = rPartStatus,
+  minline = TRUE,
+  upper = "size"
+)
+```
+
+From the graph we can see that a cp of 0.039 is ideal as it is under the horizontal (dotted) reference line. We can prune the tree with this CP value.
+
+```{r}
+# Prune the rpart Graduate Tree ----
+rPartStatus2 <- prune(
+  tree = rPartStatus,
+  cp = 0.029
+)
+```
+
+We can plot the pruned tree
+
+```{r}
+fancyRpartPlot(
+  model = rPartStatus2,
+  main = NULL,
+  sub = NULL
+)
+```
+
+We ca see the pruned tree has cut out some leaf nodes. This would help in avoiding overfitting the model.
+
+### Part 4: Results
+
+Now, we can evaluate the results of the tree on the testing data from the initial 80-20 split.
+As is true whenever we use validation approaches, we need to test out our model on the testing data set.
+This will give us a more accurate understanding of how well the model fits the context we're seeking to build our understanding of.
+
+An important part of our results is understanding the role of prediction and inference.
+In a broad sense, prediction refers to the process of making forecasts about future events or unknown values based on a model while inference generally refers to the process of drawing conclusions from data.
+For the basic tree, I will be mainly focusing on prediction aspects of the results.
+However, later in the report, we will also be exploring inference findings.
+
+```{r}
+pred_StatusRpart2 <- predict(
+  object = rPartStatus2,
+  newdata = testingData,
+  type = "prob"
+)
+
+# Data Wrangling the predictions ----
+StatusPrediction <- data.frame(
+  rpart2_non_death = pred_StatusRpart2[, 1],
+  rpart2_death = pred_StatusRpart2[, 2]
+) %>%
+  mutate(
+    rpart2_pred = ifelse(
+      test = rpart2_death > rpart2_non_death,
+      yes = 1,
+      no = 0
+    )
+  )
+
+## Set predictions as factors
+StatusPrediction$rpart2_pred <- as.factor(StatusPrediction$rpart2_pred)
+
+# Merge supervision column into predictions data frame ----
+StatusPrediction <- cbind(
+  tempID = testingData$tempID,
+  Status = testingData$Status,
+  StatusPrediction
+)
+```
+
+We can evaluate the results of this through a confusion matrix.
+
+```{r}
+StatusPrediction$Status <- factor(StatusPrediction$Status)
+
+library(yardstick)
+
+# Build confusion matrix for second tree model
+conf_matrix <- conf_mat(
+  data = StatusPrediction,
+  truth = Status,
+  estimate = rpart2_pred
+)$table
+
+kable(
+  conf_matrix,
+  col.names = c("Prediction/Supervision", "0", "1"),
+  digits = 3,
+  booktabs = TRUE,
+  caption = "Model 1: Confusion Matrix (0=Deceased, 1=Censored)",
+  align = "c"
+) %>%
+kable_styling(latex_options = "HOLD_position")
+
+accuracy <- accuracy(StatusPrediction, Status, rpart2_pred)
+specificity <- specificity(StatusPrediction, Status, rpart2_pred)
+sensitivity <- sensitivity(StatusPrediction, Status, rpart2_pred)
+```
+
+```{r}
+# Build a data frame with model metrics ----
+StatusPreds <- StatusPrediction %>%
+  dplyr::select(tempID, Status, contains("_pred")) %>%
+  pivot_longer(
+    cols = !c(tempID, Status),
+    names_to = "model",
+    values_to = "prediction"
+  )
+
+accuracy <- StatusPreds %>%
+  group_by(model) %>%
+  accuracy(
+    truth = Status,
+    estimate = prediction
+  )
+
+sensitivity <- StatusPreds %>%
+  group_by(model) %>%
+  sensitivity(
+    truth = Status,
+    estimate = prediction,
+    event_level = "second"
+  )
+
+specificity <- StatusPreds %>%
+  group_by(model) %>%
+  specificity(
+    truth = Status,
+    estimate = prediction,
+    event_level = "second"
+  )
+
+modelMetrics <- bind_rows(
+  accuracy,
+  sensitivity,
+  specificity
+)
+```
+
+With this, we can also calculate the model's metrics on the test data.
+
+```{r}
+# Make a nice looking table of model metrics ----
+modelMetrics %>%
+  dplyr::select(model, .metric, .estimate) %>%
+  pivot_wider(
+    id_cols = model,
+    names_from = .metric,
+    values_from = .estimate
+  ) %>%
+  kable(
+    digits = 3,
+    booktabs = TRUE,
+    align = "c",
+    table.attr = 'data-quarto-disable-processing="true"'
+  )
+```
+
+As you can see the model shows good accuracy with 72.9%. However, the specificity is an issue with 57.6%. Now let us compare this with a different type of model: logistic regression.
+
+# Part II: Exploring the predictors for the patient's status using logistic regression
+
+Next, let us explore our question with a different type of model and compare the results with our tree. To do this, we will be using (binary) logistic regression model.
+
+## Introduction
+
+Logistic regression is a statistical method used for binary classification, which predicts the probability of an outcome that can be either true or false.
+This is done by understanding the relationship between a dependent binary variable and one or more independent variables.
+Logistic regression is easy to implement and interpret.
+However there are also some drawbacks, the model assumes a linear relationship between the independent variables and the log odds of the dependent variable, which may not always hold true in complex real-world scenarios.
+Furthermore, unlike the decision tree, linear regression models cannot ignore N/A values.
+
+![[Logistic Regression Model (Source:]{.underline} <https://www.spiceworks.com/tech/artificial-intelligence/articles/what-is-logistic-regression/>[)]{.underline}](46-4.png)
+
+## Methodology
+
+Similar to the tree, we will start by splitting the data set into training and testing sets.
+Note that the data now only contains instances where the patient was deceased.
+The training set will be used to train our model, while the testing set will help evaluate its performance.
+We'll use 80% of the data for training and the remaining 20% for testing.
+To allow reproducible code, we have fixed the seed at 380.
+
+With this, we will build two candidate models:
+
+-   The first model will test the classification based on just the SGOT
+
+-   The second model will use a step wise function using various predictors to see the best performance.
+
+Another important consideration is the application of prediction (estimating an outcome based on input variables) and inference (understanding the relationships between variables).
+We will evaluate the inference through the coefficient analysis and prediction through roc curves and confusion matrix.
+It is important to note that these metrics will complement each other in our understanding of the data.
+However, the main focus of this analysis will be prediction and we will work with several metrics to evaluate it.
+
+We will use a variety of tools to understand the model's performance.
+
+```{r}
+cirrhosisRegression <- cirrhosis
+cirrhosisRegression$Status <- ifelse(cirrhosisRegression$Status == "C" | cirrhosisRegression$Status == "CL", 1, 0)
+```
+
+```{r}
+#model data
+LRmodelData <- cirrhosisRegression %>%
+  drop_na() %>%
+  mutate(
+    tempID = row_number(),
+    .before = Status
+  )
+
+# Set seed for reproducibility and slice
+set.seed(380)
+trainingData <- LRmodelData %>%
+  group_by(Status) %>%  # group_by() function ensures that the data
+  slice_sample(prop = 0.80)
+
+testingData <- LRmodelData %>%
+  filter(!(tempID %in% trainingData$tempID))
+
+trainingResults <- trainingData
+```
+
+```{r}
+# Form Candidate Model 1
+model1 <- glm(
+  formula = Status ~ SGOT,
+  data = trainingData,
+  family = binomial
+)
+```
+
+Stepwise results:
+
+```{r models}
+# Lower bound (Intercept only)
+lower <- glm(
+  formula = Status ~ 1,
+  data = trainingData,
+  family = binomial
+)
+
+# Upper bound 
+upper <- glm(
+  formula = Status ~ Drug + Age + Sex + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Cholesterol + Albumin + Copper + Alk_Phos + SGOT + Platelets + Prothrombin + Stage,
+  data = trainingData,
+  family = binomial
+)
+
+# Stepwise search
+model2 <- step(
+  object = lower,
+  scope = list(
+    lower = lower,
+    upper = upper
+  ),
+  data = trainingData,
+  direction = "both",
+  k = 2
+)
+```
+
+# Results
+
+Initially, we'll delve into the two preliminary models independently to understand where they stand.
+Following that, we'll be deploying the best candidate model on our test data.
+Regarding confusion matrices, we'll employ a basic rule: if the predicted probability of a the patient's status being "deceased" exceeds 0.5, we'll categorize them as deceased (naïve rule).
+
+## Model 1
+
+```{r}
+# Model 1 Coefficient Table
+as.data.frame(summary(model1)$coefficients) %>%
+  rownames_to_column(var = "X") %>%
+  rename(coefficient = Estimate) %>% 
+  mutate(
+    prob_odds = case_when(
+      coefficient == "(Intercept)" ~ exp(coefficient)/(1 + exp(coefficient)),
+      .default = exp(coefficient)
+    ),
+    .after = coefficient
+  ) %>%
+  mutate(
+    `Pr(>|z|)` = ifelse(
+      test = `Pr(>|z|)` < 0.001,
+      yes = paste("< 0.001"),
+      no = `Pr(>|z|)`
+    ),
+    X = case_when(
+      X == "(Intercept)" ~ "Intercept",
+      grepl(x = X, pattern = "SGOT") ~ "SGOT"
+    )
+  ) %>%
+  kable()
+
+```
+
+This table shows us the results of our first model.
+We can see that, holding other variables constant, a one-unit increase in SGOT is associated with a decrease in the log-odds of the response variable by 0.0076446.
+Furthermore, the odds-ratio indicates that for each one-unit increase in SGOT, the odds of the event occurring decrease by about 0.76%.
+
+We can also plot the confusion matrix for this model:
+
+```{r}
+library(janitor)
+# Building confidence intervals for Model 1 coefficients
+model1CI <- confint(
+  object = model1,
+  parm = "SGOT",
+  level = 0.9
+)
+
+trainingResults <- trainingData %>%
+  ungroup() %>%
+  mutate(model1Pred = predict(object = model1, newdata = ., type = "response"))
+
+# Apply naïve rule ----
+trainingResults <- trainingResults %>%
+  mutate(
+    model1Class = case_when(
+      model1Pred > 0.5 ~ "Censored",
+      .default = "Deceased"
+    )
+  )
+
+#Confusion Matrix for Model 1
+trainingResults %>%
+  mutate(Patient_status = ifelse(Status == 1, "Censored", "Deceased")) %>%
+  tabyl(var1 = model1Class, var2 = Patient_status) %>%
+  adorn_title(
+    placement = "combined",
+    row_name = "Predicted",
+    col_name = "Actual"
+  ) %>%
+  kable(
+    booktabs = TRUE,
+    align = "c",
+    caption = "Model 1 Confusion Matrix"
+  )%>%kable_styling(latex_options = "HOLD_position")
+
+```
+
+We can see that this model tends to over predict censored values. This shows the need of bringing in more factors. Next let us look at our Model 2, which has multiple factors as discussed earlier.
+
+```{r}
+#Coeff for model 2
+as.data.frame(summary(model2)$coefficients) %>%
+  rename(coefficient = Estimate) %>% 
+  mutate(
+    prob_odds = case_when(
+      coefficient == "(Intercept)" ~ exp(coefficient)/(1 + exp(coefficient)),
+      TRUE ~ exp(coefficient)
+    ),
+    .after = coefficient
+  ) %>%
+  kable()
+
+```
+
+This is the role of inference in evaluating our model.
+The most notable predictors are Bilirubin, Age, Alk_Phos, and Prothrombin, each showing a statistically significant relationship (p \< 0.05) with the dependent variable.
+The Intercept and EdemaY have extremely significant p-values, but the practical significance of EdemaY is questionable due to its large standard error.
+Other variables like Spiders, SGOT, DrugPlacebo, and Copper, while contributing to the model, do not reach conventional levels of statistical significance (p \< 0.05).
+
+```{r}
+#do the Tukey-Anscombe plot
+ggplot(
+  data = data.frame(
+    residuals = residuals(model2, type = "pearson"),
+    fitted = fitted(model2)
+  ),
+  mapping = aes(x = fitted, y = residuals)
+) +
+  geom_point() +
+  geom_smooth(
+    formula = y ~ x,
+    method = stats::loess,
+    method.args = list(degree = 1),
+    se = FALSE,
+    linewidth = 0.5
+  ) +
+  theme_bw() +
+  labs(
+    x = "Fitted",
+    y = "Pearson Residuals"
+  )
+```
+
+This figure shows us the Tukey-Anscombe plot using Pearson residuals for Model 2.
+In an ideal fit, the residuals should be evenly distributed about zero with constant mean and variance.
+The shape of the line suggests that the model is not capturing someunder lying structure in the datain extreme cases.
+
+```{r}
+#plot the gvif
+as.data.frame(car::vif(model2)) %>%
+  kable(
+    digits = 3,
+    align = "lcccc",
+    booktab = TRUE,
+    format.args = list(big.mark = ","),
+    table.attr = 'data-quarto-disable-processing="true"',
+    label = "GVIF analsyis"
+  )
+
+```
+
+The Variance Inflation Factor (VIF) values for the variables in the model (Bilirubin, Age, Alk_Phos, Prothrombin, Spiders) are all close to 1, indicating minimal multicollinearity.
+This means that these predictors are relatively independent of each other, enhancing the reliability of the model.
+
+```{r}
+#Store the predicted and actual values for Model 2
+trainingResults$model2Pred <- predict(model2, type = "response")
+trainingResults$model2Class <- ifelse(trainingResults$model2Pred > 0.5, "Censored", "Deceased")
+trainingResults$Actual <- ifelse(trainingData$Status == 1, "Censored", "Deceased")
+
+# Create confusion matrix using table
+confusionMatrixRegression <- table(Predicted = trainingResults$model2Class, Actual = trainingResults$Actual)
+
+kable(confusionMatrixRegression, caption = "Confusion matrix for Model 2") %>%
+  kable_classic(latex_options = "HOLD_position")
+
+```
+
+From this confusion matrix we can see the relationships between the True Positive, True Negative, False Positive and False Negative values. From this we can calculate:
+
+-   **Accuracy**: Approximately 81.36%
+
+-   **Recall**: Approximately 79.01%
+
+-   **Precision**: Approximately 72.73%
+
+-   **F1 Score**: Approximately 75.74%
+
+Lastly, let us look at the separation plots for each of the models.
+
+```{r}
+library(pROC)
+library(separationplot)
+# Fit ROC Curves for later
+## Model 1
+model1ROC <- roc(
+  formula = Status ~ model1Pred,
+  data = trainingResults
+)
+model1ROC_df <- data.frame(
+  threshold = model1ROC$thresholds,
+  sensitivity = model1ROC$sensitivities,
+  specificity = model1ROC$specificities,
+  model = "Model 1"
+)
+## Model 2
+model2ROC <- roc(
+  formula = Status ~ model2Pred,
+  data = trainingResults
+)
+model2ROC_df <- data.frame(
+  threshold = model2ROC$thresholds,
+  sensitivity = model2ROC$sensitivities,
+  specificity = model2ROC$specificities,
+  model = "Model 2"
+)
+```
+
+```{r}
+# Convert 'Actual' column to numeric 0/1
+trainingResults <- trainingResults %>%
+  mutate(
+    actualNum = if_else(Actual == "Deceased", 0, 1)
+  )
+
+#Sepeation Plot
+par(mar = c(4,0,0,0))
+separationplot(
+  pred = trainingResults$model1Pred, 
+  actual = trainingResults$actualNum, 
+  type = "rect",
+  line = TRUE, 
+  lwd2 = 2,
+  show.expected = TRUE, 
+  newplot = FALSE,
+  heading = "Model 1"
+)
+
+```
+
+```{r}
+#Sepeation Plot
+par(mar = c(4,0,0,0))
+separationplot(
+  pred = trainingResults$model2Pred, 
+  actual = trainingResults$actualNum, 
+  type = "rect",
+  line = TRUE, 
+  lwd2 = 2,
+  show.expected = TRUE, 
+  newplot = FALSE,
+  heading = "Model 2"
+)
+```
+
+The separation plot assesses the the fit of the model by providing the model's ability to predict occurrences with a high probability and non-occurrences with low probability.
+The separation plot above separation plot suggests that Model 2 has a reasonably good performance in predicting the patient's status, especially for the observations on the left-most side of the plot compared to Model 1 with the training data.
+We will later compare this graph with the testing data.
+
+Lastly, let us look at the ROC curves for both the models.
+
+```{r}
+## Merge into existing data frame
+rocData <- rbind(model1ROC_df, model2ROC_df)
+
+## AUC Data
+aucData <- data.frame(
+  model = c("Model 1", "Model 2"),
+  auc = c(model1ROC$auc, model2ROC$auc)
+)
+```
+
+```{r}
+#ROC plot
+ggplot(
+  data = rocData,
+  mapping = aes(x = 1 - specificity, y = sensitivity, color = model)
+) +
+  geom_path() +
+  geom_abline(
+    slope = 1,
+    intercept = 0,
+    linetype = "dotted"
+  ) +
+  geom_text(
+  inherit.aes = FALSE,
+  data = aucData,
+  mapping = aes(label = paste(model, "AUC: \n", round(auc, 3))),
+  x = c(0.25, 0.15),
+  y = c(0.4, 1.05)
+)
+```
+
+From the graphs we can interpret that:
+
+Model 1:
+
+-   Its ROC curve is above the line of no discrimination, indicating that the model has some predictive capabilities
+
+-   The AUC is 0.638, which is better than random guessing but suggests there's room for improvement.
+
+Model 2:
+
+-   The ROC curve for Model 2 is significantly above that of Model 1, and much closer to the top-left corner, indicating better predictive performance.
+
+-   The AUC is 0.899, which suggests a good classification performance, and it's notably better than Model 1.
+
+-   Its ability to discriminate between positive and negative classes is superior to that of Model 1.
+
+Lastly, let us plot our influence plot.
+
+```{r}
+# Influence Plot for Model 2
+idCases <- car::influencePlot(model2)
+```
+
+The influence plot shows several data points with high leverage and large residuals, indicating potential outliers.
+Some observations, notably those labeled like "149," have significant influence on the regression model due to their Cook's D values.
+
+# Testing our model
+
+Using our final model, we now turn to see how well this classifier does on our testing data. Recall that we initially set the test data during the train test split.
+
+The confusion matrix below shows the performance of our model using the naïve decision rule.
+
+```{r}
+# Set up testing data results
+testingData <- testingData %>%
+  mutate(
+    gradNum = case_when(
+      Status == 0 ~ 0,
+      Status == 1 ~ 1
+    ),
+    .after = Status
+  )
+testingData$predict <- predict(
+  object = model2,
+  newdata = testingData,
+  type = "response"
+)
+testingData <- testingData %>%
+  mutate(
+    model2Class = case_when(
+      predict > 0.5 ~ "Censored",
+      .default = "Deceased"
+    )
+  )
+```
+
+```{r}
+testingData$Status <- ifelse(testingData$Status == 1, "Censored", "Deceased")
+
+# Build Confusion Matrix for Testing Data
+testingData %>%
+  tabyl(var1 = model2Class, var2 = Status) %>%
+  adorn_title(
+    placement = "combined",
+    row_name = "Predicted",
+    col_name = "Actual"
+  ) %>%
+  kable(
+    caption = "Confusion Matrix for Test data"
+  )%>%
+kable_styling(latex_options = "HOLD_position")
+```
+
+-   **Accuracy**: Approximately 71.43%
+
+-   **Recall**: Approximately 76.92%
+
+-   **Precision**: Approximately 43.48%
+
+-   **F1 Score**: Approximately 55.56%
+
+This shows that our model is has struggled with overfitting, with an especially low precision score.
+We will discuss this in the comparison with the tree model, but this is an importnat limitation of our data size as we discussed in the introduction.
+Lastly, we can plot the separation plot.
+We can see the separation has increased due to the over fitting we discussed.
+
+```{r}
+#Sepeation Plot
+par(mar = c(4,0,0,0))
+separationplot(
+  pred = testingData$predict, 
+  actual = testingData$gradNum, 
+  type = "rect",
+  line = TRUE, 
+  lwd2 = 2,
+  show.expected = TRUE, 
+  newplot = FALSE
+)
+
+```
+
+#Comparison between models
+
+Now let us compare the performance of our logistic regression model with a decision tree model. We will use the same train test split as before.
+
+# Part III: Exploring the sub-clusters of the cirrhosis data using unsupervised learning
+
+In this section, we sought to if there are sub-collections of patients.
+We expect sub-collections to mimic medical definitions of the stages of Biliary Cirrhosis but are curious to see if clustering can reveal some newer observations.
+
+## Introduction
+
+Biliary Cirrhosis patients are typically clustered in four main clusters ([Source](https://www.healthline.com/health/primary-biliary-cirrhosis#stages){.uri}):
+
+-   Stage 1: There's inflammation and damage to the walls of medium-sized bile ducts.
+
+-   Stage 2: There's blockage of the small bile ducts.
+
+-   Stage 3: This stage marks the beginning of scarring.
+
+-   Stage 4: Cirrhosis has developed. This permanent, severe, scarring and damage to the liver.
+
+In this section we will be using clustering on our data.
+K-means Clustering is an unsupervised learning technique where data points are grouped based on their similarities.
+It's commonly used to identify patterns and structures within datasets without prior knowledge of the groups.
+The main advantage of clustering is its ability to discover hidden patterns in data.
+However, a significant drawback is the subjectivity in defining the 'similarity' criteria, which can lead to varying results and interpretations.
+
+[![K-means clustering ([Source](https://medium.datadriveninvestor.com/k-means-clustering-4a700d4a4720))](1*fz-rjYPPRlGEMdTI-RLbDg.png)](https://medium.datadriveninvestor.com/k-means-clustering-4a700d4a4720)
+
+An important motivation for this part of the study is explore how does "Stage" which is typically defined based on medical professionals' observations compares to the data that we have in this data set.
+
+## Pre-processing the data
+
+We had to pre-process the data for the best performance.
+First, we removed the 'Stage' column to avoid using outcome-related features in the unsupervised learning.
+The rows with N/A values were also removed.
+Lastly, we transformed various categorical variables into numeric formats, which is necessary for clustering algorithms.
+
+```{r}
+#do the ml, take about observational stages vs now quantify, look into the different variables and what they mean.
+cirrhosisCluster <- cirrhosis
+cirrhosisCluster2 <- na.omit(cirrhosis)
+
+cirrhosisCluster <- cirrhosisCluster %>%                 
+                                  dplyr::select(-Stage)
+
+cirrhosisCluster <- na.omit(cirrhosisCluster) #cannot have NA values in clustering
+
+cirrhosisCluster$Age_Years <- round(cirrhosisCluster$Age_Years) #round age to whole numbers
+
+#reode sex... recode others later if need be #female is 0
+
+#cirrhosisCluster  <- cirrhosisCluster %>%
+#  select( -c(Status, Drug, Edema)) %>%
+
+# mutate(Sex = ifelse(Sex == 'Female', 0, 1))
+
+  cirrhosisCluster  <- cirrhosisCluster %>%
+    mutate(Sex = ifelse(Sex == 'Female', 0, 1)) %>%
+          mutate(Transplant = ifelse(Status == "CL", 1, 0)) %>%
+              mutate(Status = ifelse(Status %in% c('C', 'CL'), 0 , 1)) %>%
+                    mutate(Drug = ifelse(Drug == "D-penicillamine", 0, 1)) %>%
+                        mutate(EdemaDiurectics = ifelse(Edema %in% c('S', 'Y'), 1, 0)) %>%
+                              mutate(NoEdemaORD = ifelse(Edema == 'N' , 1, 0)) %>%
+                                        mutate(EdemaANDD = ifelse(Edema == "Y", 1, 0)) %>%
+                                                mutate(EdemaORD = ifelse(Edema == "S", 1, 0)) %>%
+                                                                              dplyr::select(-Edema)
+
+#Data is already factored
+
+#mutate(Sex = ifelse(Sex == 'Female', 0, 1)) %>%
+
+   #mutate(Status = ifelse(Status %in% c('C', 'CL'), 0 , 1)) %>%
+          #    mutate(Drug = ifelse(Drug == "D-penicillamine", 0, 1))
+
+#???:
+
+#make column yes no edema #then yes no under treatment
+
+#same for transplant stuff
+```
+
+```{r}
+distCirrhosis <- dist(
+  x = cirrhosisCluster,
+  method = "euclidean"
+)
+
+```
+
+## Methodology
+
+To apply the k-means algorithm.
+A mixed hierarchical and non-hierarchical was applied.
+This was done using the hkmeansm module, with k = 4 as our initial value.
+Based on the data properties, the euclidean (L2) distance works best.
+
+```{r}
+
+hybridCirrhosis <- hkmeans(
+  x = cirrhosisCluster,
+  k = 4,
+  hc.metric = "euclidean",
+  hc.method = "ward.D",
+  iter.max = 10
+)
+
+```
+
+## Results
+
+We were able to visualize the results of the k-means algorithm. The first step was making a color palette and plotting the dendrogram tree.
+
+```{r}
+
+StagesPalette <- c("#AA336A", "#770737", "#40B5AD", "#009E60", "#9FE2BF")
+
+```
+
+```{r}
+## MAKE A NEW PALLETE TO VISUALIZE
+# Plot the initial dendrogram for hybrid approach ----
+set.seed(380)
+hkmeans_tree(
+  hkmeans = hybridCirrhosis,
+  rect.col = StagesPalette,
+  cex = 0.4,
+  main = "Initial Hierarchical Clusters"
+)
+
+```
+
+As, you can see, the model was able to create clear clusters for the data.
+However, it is important to find the best value of k to get the optimal clusters.
+To do this we can use a scree plot.
+Here, we're looking for the number of clusters that corresponds to the "elbow".
+
+```{r}
+# Create scree plot for choosing k ----
+library(factoextra)
+set.seed(380)
+fviz_nbclust(
+  x = cirrhosisCluster,
+  diss = NULL,
+  FUNcluster = kmeans,
+  method = "wss",
+  k.max = 10
+)
+
+```
+
+From this, we identified that 5 was the ideal value of k, where the Total Within [Cluster] Sums of Squares begins leveling off.
+Let us create a new k-means model with k = 5.
+We can plot this refined model, with a format more easy to visualize.
+
+```{r}
+
+hybridCirrhosis2 <- hkmeans(
+  x = cirrhosisCluster,
+  k = 5,
+  hc.metric = "euclidean",
+  hc.method = "ward.D",
+  iter.max = 10
+)
+
+```
+
+```{r}
+# Plot the final dendrogram for hybrid approach ----
+# library(factoextra)
+fviz_dend(
+  x = hybridCirrhosis2,
+  cex = 0.4,
+  palette = StagesPalette,
+  rect = FALSE,
+  horiz = TRUE,
+  repel = TRUE,
+  main = "Final Dendrogram"
+)
+
+```
+
+As you can see, the 5 clusters were identified, which are color coded in our updated diagram.
+
+Lastly, let us plot these clusters. Here is a plot of the initial model where k = 4
+
+```{r}
+# Plot the final clustering for hybrid approach ----
+# library(factoextra)
+fviz_cluster(
+  object = hybridCirrhosis,
+  stand = FALSE,
+  geom = "point",
+  main = "Hybrid Cluster Plot - Initial model"
+) +
+  scale_color_manual(values = StagesPalette) +
+  scale_fill_manual(values = StagesPalette) +
+  theme_bw()
+
+```
+
+We can also plot our refined model, as you can see the plot below identifies 5 clear clusters.
+
+```{r}
+
+# Plot the final clustering for hybrid approach ----
+# library(factoextra)
+fviz_cluster(
+  object = hybridCirrhosis2,
+  stand = FALSE,
+  geom = "point",
+  main = "Hybrid Cluster Plot - Refined model"
+) +
+  scale_color_manual(values = StagesPalette) +
+  scale_fill_manual(values = StagesPalette) +
+  theme_bw()
+
+```
+
+Now we can go back to our initial goal: comparing how the custers compare to the Stage variable
+
+```{r}
+cirrhosisCluster2$cluster <- hybridCirrhosis$cluster
+# Calculate mean (or median, etc.) for each variable in each cluster
+library(dplyr)
+cluster_summary <- cirrhosisCluster2 %>%
+  group_by(cluster) %>%
+  summarise_all(funs(mean(., na.rm = TRUE))) # Replace mean with median or any other function as necessary
+kable(cluster_summary %>% dplyr::select(-Status, -Drug, - Sex, -Edema, -Stage), label = "Cluster Summary")
+ggplot(cirrhosisCluster2, aes(x = factor(cluster), fill = factor(Stage))) +
+  geom_bar(position = "dodge") +
+  labs(title = "Distribution of Stages within Clusters",
+       x = "Cluster",
+       y = "Count",
+       fill = "Stage") +
+  theme_minimal()
+
+```
+
+We can see that the clusters are not evenly distributed. This is a good sign, as it means that the clusters are not just a random grouping of the data.
+
+# Discussion
+
+# Author Contributions
+
+# References--Citation style is your choice, but all sources should be documented (both in text and in the References section). This includes where you got your data.
+
+\newpage
+
+# Code Appendix
+
+```{r codeAppendix}
+#| ref.label = knitr::all_labels(),
+#| echo = TRUE,
+#| eval = FALSE
+```

### 2023-12-16 Submission.Rmd
CHANGES

@@
   pdf_document: default
+  toc: true
 authors: Mihir Kulkarni+Nithika Menon
@@
 
-![Primary biliary cholangitis (https://www.mayoclinic.org/diseases-conditions/primary-biliary-cholangitis/symptoms-causes/syc-20376874)](diagram.jpeg)
+![Primary biliary cholangitis (<https://www.mayoclinic.org/diseases-conditions/primary-biliary-cholangitis/symptoms-causes/syc-20376874>)](diagram.jpeg)
 
@@
 ```{r}
-#?????
-cirrhosis %>% 
-  ggplot(aes(x = N_Days)) + geom_histogram() +
-  facet_wrap(facets = vars(Status))
-
-```
-
-```{r}
 #Required libraries
@@
 
-Stepwise results:
-
-```{r models}
+```{r results = 'hide'}
 # Lower bound (Intercept only)
@@
 
-# Results
+## Results
 
@@
 
-## Model 1
+### Model 1
 
@@
 
+### Model 2
+
 ```{r}
@@
 
-# Testing our model
+## Testing our model
 
@@
 
-#Comparison between models
-
-Now let us compare the performance of our logistic regression model with a decision tree model. We will use the same train test split as before.
+# Discussion: Comparison between models
+
+Let us recall the differences between the two models:
+
+Logistic Regression:
+
+-   Pros: Simple, efficient, good for linear relationships, provides outcome probabilities
+
+-   Cons: Assumes linear decision boundary, struggles with complex relationships, sensitive to imbalanced data
+
+Decision Trees:
+
+-   Pros: Handles non-linear data, easy to interpret, works with mixed data types, non-parametric, N/A values
+
+-   Cons: Prone to overfitting, sensitive to data changes, poor at extrapolation, biased with imbalanced datasets
+
+Now let us compare the performance of our logistic regression model with a decision tree model.
+
+| Testing Data Score | CART-based approach | Regression based Approach |
+|--------------------|---------------------|---------------------------|
+| Accuracy           | 72.9%               | 71.43%                    |
+| Precision          | 57.6%               | 43.48%                    |
+| Recall             | 82.70%              | 76.92%                    |
+
+We can see that for this data, surprisingly, the tree based approach had better results.
+This is an uncommon observation.
+We believe this was because of the sparcity of the data and how the rows N/A values prevented over fitting with the small data size.
 
@@
 
-Now we can go back to our initial goal: comparing how the custers compare to the Stage variable
-
-```{r}
-cirrhosisCluster2$cluster <- hybridCirrhosis$cluster
+Now we can go back to our initial goal: comparing how the clusters compare to the Stage variable.
+
+```{r}
+cirrhosisCluster2$cluster <- hybridCirrhosis2$cluster
 # Calculate mean (or median, etc.) for each variable in each cluster
@@
 
-We can see that the clusters are not evenly distributed. This is a good sign, as it means that the clusters are not just a random grouping of the data.
+We can see that the clusters have a somewhat connection to the Stage, but is not as direct as we hypothesized.
+We were also able to find the cluster summary compared to the rest of the variables and were able to plot that into a table.
 
@@
 
+To conclude, in this report, we were able to explore Biliary Cirrhosis and Treatments through various angles.
+Though our clustering we were able to find new patterns among the data.
+Though the logistic and CART algorithms we were able to build models that were able to predict the chance of a patient's death using various predictors.
+
+Along with the analysis in the report project, we attempted to explore various models such as predicting the likelihood of a liver transplant and the number of days until an event.
+However, due to our data's and report's limitations, such as the limited data quantities, we were unable to get concrete solutions.
+We hope that our results can be used to build even better models, making healthcare more inclusive, efficient and affordable.
+
 # Author Contributions
 
-# References--Citation style is your choice, but all sources should be documented (both in text and in the References section). This includes where you got your data.
+We collaborated throughout the document using GitHub, but here are the main responsibilities:
+
+**Introduction: Nithika**
+
+**EDA: Nithika**
+
+**Tree: Mihir**
+
+**Logistic Regression: Mihir**
+
+**Clustering: Nithika**
+
+**Discussion: Mihir**
+
+# References
+
+Dataset: <https://www.kaggle.com/datasets/fedesoriano/cirrhosis-prediction-dataset/data>
+
+Study: <https://faculty.washington.edu/abansal/ShortCourse_DynamicDecisionMaking/Dickson1989_MayoPBCOriginalArticle.pdf>
 

### 2023-12-16 Submission.Rmd
Adapted code for rubric

@@
 
+## Goals
+
+Our goal is to explore the data through multiple approaches including regression, classification and clustering. Let us frist clear some definitions and differences overall.
+
+[Unsupervised vs Supervised learning]{.underline}
+
+-   **Supervised Learning:** Involves training a model on a labeled dataset, where the correct answers are provided, allowing the model to learn the relationships between the input data and output labels.
+
+-   **Unsupervised Learning:** Involves training a model on data without labeled responses, enabling the model to identify patterns and structures in the data on its own.
+
+[Regression vs Clustering]{.underline}
+
+-   **Regression Learning**: A type of machine learning where the model predicts a continuous output. For example, predicting house prices based on features like size and location.
+
+-   **Classification Learning**: A type of machine learning where the model categorizes data into discrete classes. For example, determining if an email is spam or not based on its content.
+
 # Exploratory Data Analysis
@@
 
-We ca see the pruned tree has cut out some leaf nodes. This would help in avoiding overfitting the model.
+We ca see the pruned tree has cut out some leaf nodes.
+This would help in avoiding overfitting the model.
+As a reminder, Overfitting is when a model learns the detail and noise in the training data to the extent that it negatively impacts the performance of the model on new data.
 

### 2023-12-16 Submission.Rmd
yuhh

@@
 
-Both of us have been very interetsted in the intersection of medicine and machine learning.
-The idea that the concepts we are learning in our DS majors could have such an impact on the health industry was really special.
-Another important aspect was that there is so much potential for advancement of medicine, we both agreed that this area is MLs most promising aspect.
-The cirrhosis data set perticularly stood out to us because the data set included so many interesting attributes and we anted to see how they are connected with the outcomes of the patient.
-
-The data set also includes some important limitations.
-Most importantly, there are only 418 cases, greatly affecting the types of analysis we could do.
-This shows realm world problems data scientists face.
-Thus, our goal is to use our Statistical DS skillset to approach this problem, despite the limitations.
+We are very interested in the intersection of medicine and machine learning.
+To us, clinical research is engaging and interesting to learn about.
+It is also special because of the impact it can have on the healthcare industry.
+There is so much potential for advancement of medicine using AI, and we are passionate about this.
+The 'cirrhosis' data set stood out to us because the data set included many diverse attributes.
+A limitation was that this dataset had only 418 cases.
+This reflects a very probable issue when using clinical data.
+Thus, our goal is to use our Statistical DS skillet to approach this problem despite its limitations.
 
@@
 'Status' was either Dead, Censored, or Censored due to liver transplant.
-A link to the original study is here: <https://faculty.washington.edu/abansal/ShortCourse_DynamicDecisionMaking/Dickson1989_MayoPBCOriginalArticle.pdf>
-
-The data contains the following attributes ([Data Source](https://www.kaggle.com/datasets/fedesoriano/cirrhosis-prediction-dataset/data)):
+A link to the original study is in our references.
 
@@
 
-Our goal is to explore the data through multiple approaches including regression, classification and clustering. Let us frist clear some definitions and differences overall.
+Our goal is to explore the data through our research questions while demonstrating our understandings of including regression, classification and clustering.
+Let us first clear some definitions and differences overall.
 
@@
 
--   **Regression Learning**: A type of machine learning where the model predicts a continuous output. For example, predicting house prices based on features like size and location.
+-   **Regression Learning(logistic)**:Logistic regression is a process of modeling the probability of a dichotomous outcome given an input variable.
 
@@
 
-### Survival Analysis Based on Treatment and Stage: How does the survival rate differ among patients at different stages of PBC who received D-penicillamine versus placebo?
-
-This question investigates the effectiveness of the drug D-penicillamine compared to a placebo, considering the stage of PBC.
-By analyzing survival rates across different disease stages and treatment types, we can assess the drug's effectiveness at various disease stages.
-This information is vital for clinicians to make informed decisions about treatment plans and for researchers to understand the drug's impact.
-
-```{r}
+### Impact of Liver Complications on Survival: How do the presence of ascites, hepatomegaly, and spiders relate to survival time?
+
+Exploring how liver-related conditions (ascites, hepatomegaly, and spiders) affect patient survival provides insights into the severity of these complications in PBC progression.
+
+```{r}
+
 surv_object <- Surv(cirrhosis$N_Days, cirrhosis$Status == 'D')
-surv_fit <- survfit(surv_object ~ cirrhosis$Drug + cirrhosis$Stage, data = cirrhosis)
-ggsurvplot(surv_fit, 
-           data = cirrhosis,
-           conf.int = TRUE, 
-           #palette = c("#00AFBB", "#E7B800", "#FC4E07"),
-           xlab = "Days", 
-           ylab = "Survival Probability", 
-           title = "Survival Curves by Treatment and Stage")
-```
-
-### Impact of Liver Complications on Survival: How do the presence of ascites, hepatomegaly, and spiders relate to survival time?
-
-Exploring how liver-related symptoms (ascites, hepatomegaly, and spiders) affect patient survival provides insights into the severity of these complications in PBC progression.
-This can help in identifying high-risk patients and understanding the disease's impact on liver function.
-
-```{r}
 surv_fit_complications <- survfit(surv_object ~ cirrhosis$Ascites + cirrhosis$Hepatomegaly + cirrhosis$Spiders, data = cirrhosis)
@@
 
-### Correlation between Biochemical Markers and Disease Stage: What is the correlation between biochemical markers (Bilirubin, Cholesterol, Albumin, Copper, Alk_Phos, SGOT, Triglycerides) and the histologic stage of the disease?
-
-This analysis is crucial to understand how different biochemical markers correlate with the disease's progression.
-Such correlations can aid in the early detection of disease severity, help in monitoring the disease progression, and potentially guide treatment adjustments.
+Survival Curves are commonly used in medical research to measure the ability of a patient to survive.
+There was a survivial curve in the original study.
+Out of curiosity, we attempted one here.
+The combination of of Ascites, Hepatomegaly, and Spiders shows many overlapping curves with often large variation.
+The labels cut off here, but this is more of a visual exploration.
 
@@
 To do this, we will be using Decision Trees.
-Let us start by exploring what these are and why we chose to use them.
 
@@
 However there are also some drawbacks, the model assumes a linear relationship between the independent variables and the log odds of the dependent variable, which may not always hold true in complex real-world scenarios.
-Furthermore, unlike the decision tree, linear regression models cannot ignore N/A values.
+Furthermore, unlike the decision tree, regression models cannot ignore N/A values.
 
@@
 The rows with N/A values were also removed.
-Lastly, we transformed various categorical variables into numeric formats, which is necessary for clustering algorithms.
-
-```{r}
-#do the ml, take about observational stages vs now quantify, look into the different variables and what they mean.
+Lastly, we transformed various categorical variables into numeric formats which is the easiest way to deal with them in clustering algorithms.
+
+```{r}
 cirrhosisCluster <- cirrhosis
@@
 cirrhosisCluster$Age_Years <- round(cirrhosisCluster$Age_Years) #round age to whole numbers
-
-#reode sex... recode others later if need be #female is 0
-
-#cirrhosisCluster  <- cirrhosisCluster %>%
-#  select( -c(Status, Drug, Edema)) %>%
-
-# mutate(Sex = ifelse(Sex == 'Female', 0, 1))
 
@@
 
-#Data is already factored
-
-#mutate(Sex = ifelse(Sex == 'Female', 0, 1)) %>%
-
-   #mutate(Status = ifelse(Status %in% c('C', 'CL'), 0 , 1)) %>%
-          #    mutate(Drug = ifelse(Drug == "D-penicillamine", 0, 1))
-
-#???:
-
-#make column yes no edema #then yes no under treatment
-
-#same for transplant stuff
 ```
@@
 
-We were able to visualize the results of the k-means algorithm. The first step was making a color palette and plotting the dendrogram tree.
+We were able to visualize the results of the k-means algorithm. We created a new color palette and plotted the dendrogram.
 
@@
 cirrhosisCluster2$cluster <- hybridCirrhosis2$cluster
-# Calculate mean (or median, etc.) for each variable in each cluster
-library(dplyr)
+#library(dplyr)
 cluster_summary <- cirrhosisCluster2 %>%
@@
   summarise_all(funs(mean(., na.rm = TRUE))) # Replace mean with median or any other function as necessary
-kable(cluster_summary %>% dplyr::select(-Status, -Drug, - Sex, -Edema, -Stage), label = "Cluster Summary")
+kable(cluster_summary %>% dplyr::select(-Status, -Drug, - Sex, -Edema,-Stage), label = "Cluster Summary")
 ggplot(cirrhosisCluster2, aes(x = factor(cluster), fill = factor(Stage))) +
@@
 ```
-
-We can see that the clusters have a somewhat connection to the Stage, but is not as direct as we hypothesized.
-We were also able to find the cluster summary compared to the rest of the variables and were able to plot that into a table.
+From the table, we see the mean values for each variable belonging to each group.Group 5, for instance, has the largest mean value for Cholestorol.
+
+From the graph, we can see that the breakdown of the different stages in our clusters.
+It's important to note that the amount of patients with each stage was not equal in that data and that is shown here.
+However, there are new observations to be made as our findings produced 5 clusters.
 
@@
 To conclude, in this report, we were able to explore Biliary Cirrhosis and Treatments through various angles.
-Though our clustering we were able to find new patterns among the data.
-Though the logistic and CART algorithms we were able to build models that were able to predict the chance of a patient's death using various predictors.
+We were able to build two different kinds of models: decision tree and logistic regression models.
+Bilirubin, Age, and Prothrombin were strong predictors to predict the status of the patient.
+From our clustering algorithm, we were able to create 5 clusters for our data and compared it with the stages of the patients.
 
@@
 However, due to our data's and report's limitations, such as the limited data quantities, we were unable to get concrete solutions.
-We hope that our results can be used to build even better models, making healthcare more inclusive, efficient and affordable.
+s
 
@@
 
-Dataset: <https://www.kaggle.com/datasets/fedesoriano/cirrhosis-prediction-dataset/data>
-
-Study: <https://faculty.washington.edu/abansal/ShortCourse_DynamicDecisionMaking/Dickson1989_MayoPBCOriginalArticle.pdf>
+Fedesoriano. (2021, August 2). Cirrhosis prediction dataset. Kaggle. https://www.kaggle.com/datasets/fedesoriano/cirrhosis-prediction-dataset/data
+
+Dickson, et.
+al.
+(n.d.).
+Prognosis in primary biliary cirrhosis: Model for decision making.
+Prognosis in Primary Biliary Cirrhosis: Model for Decision Making.
+https://faculty.washington.edu/abansal/ShortCourse_DynamicDecisionMaking/Dickson1989_MayoPBCOriginalArticle.pdf
+
+Penn Medicine.
+(n.d.).
+Primary Biliary Cirrhosis.
+Pennmedicine.org.
+https://www.pennmedicine.org/for-patients-and-visitors/patient-information/conditions-treated-a-to-z/primary-biliary-cirrhosis#:~:text=All%20of%20the%20bile%20ducts,This%20is%20called%20biliary%20cirrhosis.
+
+Mayo Foundation for Medical Education and Research.
+(2023, November 14).
+Primary biliary cholangitis.
+Mayo Clinic.
+https://www.mayoclinic.org/diseases-conditions/primary-biliary-cholangitis/symptoms-causes/syc-20376874
+
+professional, C.
+C.
+medical.
+(n.d.).
+Primary biliary cholangitis: What it is, symptoms, treatment.
+Cleveland Clinic.
+https://my.clevelandclinic.org/health/diseases/17715-primary-biliary-cholangitis-pbc
 
